---
title: "Average Marginal Effects Benchmark"
format: gfm
execute:
  echo: true
  output: true
  warning: true
  message: true
---

```{r setup}
helper <- if (file.exists("analysis/_analysis_helpers.R")) "analysis/_analysis_helpers.R" else "../_analysis_helpers.R"
source(helper)
```

## Legacy diagnostic intent

Use automatic differentiation (AD) for standard-error delta-method gradients, but keep the final method as `avg_slopes(model_probit_selection, newdata = sel_data, wts = "weight", vcov = TRUE, type = "response")` with `options(marginaleffects_parallel = FALSE)`.  The legacy notebook noted that raw AME derivation is slow because the count of numeric variables, observations, and centered finite-difference prediction calls is large.

```{r ame-current-metadata}
ame_methods <- analysis_target_csv("bench_ame_methods", "ame_methods_benchmark.csv")
ame_notes <- analysis_target_csv("bench_ame_methods", "ame_legacy_benchmark_notes.csv")
ame_parallel <- analysis_target_csv("bench_ame_methods", "ame_parallelization_notes.csv")
full_n <- analysis_value(ame_methods, column = "n_observations")
n_numeric <- analysis_value(ame_methods, column = "n_numeric_variables")
predict_calls <- analysis_value(ame_methods, column = "centered_predict_calls_full_data")
```

The current active model frame has `r analysis_format_number(n_numeric, 0)` numeric variables and `r analysis_format_number(full_n, 0)` observations.  Centered finite differences therefore imply `r analysis_format_number(predict_calls, 0)` full-data calls of `predict()` if each numeric variable is perturbed up and down once.  This is the current-codebase analog of the legacy note that `avg_slopes()` uses centered finite differences.

Method 1 was straight sampling/subsampling.  The current benchmark keeps `set.seed(999)`, samples from the active model frame, and calls the production `marginaleffects::avg_slopes()` method.

```{r ame-methods-table}
analysis_table(
  ame_methods[intersect(c("method", "sample_size", "n_observations", "n_numeric_variables", "centered_predict_calls_full_data", "elapsed_seconds", "status"), names(ame_methods))],
  "Current AME benchmark attempts"
)
```

With `newdata = sel_data` (no subsampling, as if `num_samp` were `r analysis_format_number(full_n, 0)`), the legacy run took several minutes; the default current benchmark tier intentionally does not rerun the full-data AME timing every public build.  With the current default benchmark tier, the largest sampled benchmark uses `num_samp` equal to `r analysis_format_number(max(ame_methods$sample_size, na.rm = TRUE), 0)` and the slowest recorded current run took `r analysis_format_number(max(ame_methods$elapsed_seconds, na.rm = TRUE), 3)` seconds.

Method 2 was subsampling with forward differences.  The current analog is the row whose method is `avg_slopes_fdforward`.

```{r ame-forward-difference}
analysis_table(
  ame_methods[ame_methods$method == "avg_slopes_fdforward", intersect(c("method", "sample_size", "elapsed_seconds", "status", "reason", "fallback"), names(ame_methods)), drop = FALSE],
  "Forward-difference AME benchmark rows"
)
```

Method 3 split numeric `slopes()` and factor `comparisons()` just to check whether it improved runtime.  The legacy conclusion was no: use `avg_slopes()`.  Because the current production pipeline uses `avg_slopes()` directly, this note is retained as a benchmark note rather than a second production path.

Method 4 attempted parallelization with `future.apply`, `plan(multicore, workers = 4)`, and `options(marginaleffects_parallel = TRUE)`.  The legacy attempt failed because the exported globals exceeded available RAM.  The current benchmark therefore records the parallelization attempt as documented-not-run instead of rerunning an unsafe memory-heavy path.

```{r ame-notes}
analysis_table(ame_notes, "Legacy AME benchmark notes retained as target output")
analysis_table(ame_parallel, "Parallelization notes")
```
