---
title: "Average Marginal Effects Benchmark"
format: gfm
execute:
  echo: true
  output: true
  warning: true
  message: true
---

```{r setup}
helper <- if (file.exists("analysis/_analysis_helpers.R")) "analysis/_analysis_helpers.R" else "../_analysis_helpers.R"
source(helper)
```

## Legacy prose retained with current results

The legacy chunk began: Use automatic differentiation (AD) for SE delta method gradients. It also set `options(marginaleffects_parallel = FALSE)` and kept the final method as `avg_slopes(model_probit_selection, newdata = sel_data, wts = "weight", vcov = TRUE, type = "response")`.

```{r ame-current-metadata}
ame_methods <- analysis_target_csv("bench_ame_methods", "ame_methods_benchmark.csv")
ame_notes <- analysis_target_csv("bench_ame_methods", "ame_legacy_benchmark_notes.csv")
ame_parallel <- analysis_target_csv("bench_ame_methods", "ame_parallelization_notes.csv")
full_n <- analysis_value(ame_methods, column = "n_observations")
n_numeric <- analysis_value(ame_methods, column = "n_numeric_variables")
predict_calls <- analysis_value(ame_methods, column = "centered_predict_calls_full_data")
```

With `newdata = sel_data` (no subsampling, as if `num_samp = ` `r analysis_format_number(full_n, 0)`): the default current benchmark tier intentionally does not rerun the full-data AME timing every public build. With the current default benchmark tier, the largest sampled benchmark uses `num_samp = ` `r analysis_format_number(max(ame_methods$sample_size, na.rm = TRUE), 0)`, and the slowest recorded current run took `r analysis_format_number(max(ame_methods$elapsed_seconds, na.rm = TRUE), 3)` seconds.

The rest of this chunk is saved for quick replications of robustness checks on my final method.

Raw AME derivation is very slow: `r analysis_format_number(n_numeric, 0)` numeric variables * `r analysis_format_number(full_n, 0)` observations * 2 calls of `predict()` per observation per variable (`avg_slopes()` uses centered finite difference, see <https://marginaleffects.com/bonus/uncertainty.html#numerical-derivatives-sensitivity-to-step-size>) = `r analysis_format_number(predict_calls, 0)` `predict()` calls.

```{r ame-prose-deviation-note, results='asis'}
analysis_deviation_note("Legacy full-data and 20,000-row timings are preserved in target notes, but the rendered sentence reports current target-backed default-tier timings because rerunning the legacy full-data benchmark on every analysis refresh would defeat the analysis-notes workflow.")
```

```{r ame-predict-call-count-current-code}
data.frame(
  current_code_analog = "n_numeric_variables * n_observations * 2 predict() calls",
  n_numeric_variables = n_numeric,
  n_observations = full_n,
  centered_predict_calls = n_numeric * full_n * 2
)
```

Method 1: Straight Sampling and Subsampling. Note: "`slopes()` functions will automatically revert to `comparisons()` for binary or categorical variables" (<https://marginaleffects.com/man/r/slopes.html>). The current benchmark keeps `set.seed(999)`, samples from the active model frame, and calls the production `marginaleffects::avg_slopes()` method.

```{r ame-methods-table}
analysis_table(
  ame_methods[intersect(c("method", "sample_size", "n_observations", "n_numeric_variables", "centered_predict_calls_full_data", "elapsed_seconds", "status"), names(ame_methods))],
  "Current AME benchmark attempts"
)
```

```{r ame-current-runtime-analog}
ame_methods[, intersect(c("method", "sample_size", "elapsed_seconds", "status", "reason"), names(ame_methods)), drop = FALSE]
```

Method 2: Subsampling with forward differences. The legacy note was: calls `predict()` once per perturbation instead of twice (a la central differences, `fdcenter`). Run the above, but add `numderiv = "fdforward"` into `avg_slopes()`. Cheaper numeric differentiation didn't help at all! Reflective of how many discrete variables I have. Centered differences are generally more accurate than forward differences for continuous vars.

```{r ame-forward-difference}
analysis_table(
  ame_methods[ame_methods$method == "avg_slopes_fdforward", intersect(c("method", "sample_size", "elapsed_seconds", "status", "reason", "fallback"), names(ame_methods)), drop = FALSE],
  "Forward-difference AME benchmark rows"
)
```

Method 3: Split slopes and comparisons, just to be sure. Ran `avg_slopes()` on numeric variables and `avg_comparisons()` on factor variables. No. Just use `avg_slopes()`.

Method 4: Parallelization. The legacy attempt used `future.apply`, `plan(multicore, workers = 4)`, and `options(marginaleffects_parallel = TRUE)`. Failed! "The total size of the 18 globals exported for future expression (`FUN()`) is 7.90 GiB." As documentation states: "There is always considerable overhead when using parallel computation, mainly involved in passing the whole dataset to the different processes." To parallelize for the whole dataset after trimming the model object still required 53.96 GiB. My laptop has 24 GB of RAM! So no parallelization. The current benchmark therefore records the parallelization attempt as documented-not-run instead of rerunning an unsafe memory-heavy path.

```{r ame-notes}
analysis_table(ame_notes, "Legacy AME benchmark notes retained as target output")
analysis_table(ame_parallel, "Parallelization notes")
```
