---
title: "Fuzzy Matching Benchmark"
format: gfm
execute:
  echo: true
  output: true
  warning: true
  message: true
---

```{r setup}
helper <- if (file.exists("analysis/_analysis_helpers.R")) "analysis/_analysis_helpers.R" else "../_analysis_helpers.R"
source(helper)
```

## Legacy diagnostic intent

The legacy tuning chunk compared string-distance methods and thresholds before settling on `soundex = 0`, `qgram = 0`, `jw <= 0.15`, `dl <= 2`, and `osa <= 1`.  The current benchmark replaces the hand-inspected legacy objects with target-generated candidate pairs from the active tracker/join-map objects, while keeping the final thresholds visible.

```{r fuzzy-targets}
fuzzy_cov <- analysis_target_csv("bench_fuzzy_matching", "fuzzy_matching_candidate_pair_coverage.csv")
fuzzy_threshold <- analysis_target_csv("bench_fuzzy_matching", "fuzzy_matching_threshold_sensitivity.csv")
fuzzy_reference <- analysis_target_csv("bench_fuzzy_matching", "fuzzy_matching_legacy_tuning_reference.csv")
```

The current benchmark evaluates `r analysis_format_number(sum(fuzzy_cov$n_pairs, na.rm = TRUE), 0)` candidate pairs across `r analysis_format_number(nrow(fuzzy_cov), 0)` pair-source groups.

```{r fuzzy-current-candidate-count-analog}
data.frame(
  current_code_analog = "sum(fuzzy_cov$n_pairs, na.rm = TRUE)",
  candidate_pairs = sum(fuzzy_cov$n_pairs, na.rm = TRUE),
  pair_source_groups = nrow(fuzzy_cov)
)
```

```{r fuzzy-coverage}
analysis_table(fuzzy_cov, "Benchmark candidate-pair coverage", max_rows = 30)
```

```{r fuzzy-reference}
analysis_table(fuzzy_reference, "Legacy fuzzy-tuning reference results")
```

```{r fuzzy-thresholds}
analysis_table(fuzzy_threshold, "Threshold sensitivity by method and pair source", max_rows = 40)
```
