---
title: "Fuzzy Matching Benchmark"
format: gfm
execute:
  echo: true
  output: true
  warning: true
  message: true
---

```{r setup}
helper <- if (file.exists("analysis/_analysis_helpers.R")) "analysis/_analysis_helpers.R" else "../_analysis_helpers.R"
source(helper)
```

## Legacy prose retained with current results

Number of rows from fuzzy full joining `district_tracker` (734 rows) and `mother_tongues_01` (593 rows) with different metrics: `osa`: 859; `lv`: 859; `dl`: 859; `hamming`: 872; `lcs`: 825; `qgram`: 829; `cosine`, `jaccard`, and `jw`: 435,262. See <https://cran.r-project.org/web/packages/stringdist/stringdist.pdf#page=23> for more.

Testing different methods of fuzzy joining. With (`lcs`, `osa`), (3, 3): 189/734 rows have an NA. With (`jw`, `dl`, `osa`, `lcs`), (0.10, 2, 3, 5): 180/734. With final helper vectors `methods <- c("soundex", "qgram", "jw", "dl", "osa")` and `thresholds <- c(0, 0, 0.15, 2, 1)`: 166/734.

`soundex = 0` means phonetic variants in anglicization allowed. `qgram = 0` means rearrangements of words allowed. `jw <= 0.15` means respellings and vowel swaps with 0.85 similarity allowed. `dl <= 2` means no more than two insertions, deletions, substitutions, and transpositions allowed. `osa <= 1` means one typo allowed.

```{r fuzzy-prose-deviation-note, results='asis'}
analysis_deviation_note("The legacy hand-counts are preserved as prose/reference rows; the current benchmark additionally evaluates target-generated candidate pairs from the active tracker and join-map objects.")
```

```{r fuzzy-targets}
fuzzy_cov <- analysis_target_csv("bench_fuzzy_matching", "fuzzy_matching_candidate_pair_coverage.csv")
fuzzy_threshold <- analysis_target_csv("bench_fuzzy_matching", "fuzzy_matching_threshold_sensitivity.csv")
fuzzy_reference <- analysis_target_csv("bench_fuzzy_matching", "fuzzy_matching_legacy_tuning_reference.csv")
```

The current benchmark evaluates `r analysis_format_number(sum(fuzzy_cov$n_pairs, na.rm = TRUE), 0)` candidate pairs across `r analysis_format_number(nrow(fuzzy_cov), 0)` pair-source groups.

```{r fuzzy-current-candidate-count-analog}
data.frame(
  current_code_analog = "sum(fuzzy_cov$n_pairs, na.rm = TRUE)",
  candidate_pairs = sum(fuzzy_cov$n_pairs, na.rm = TRUE),
  pair_source_groups = nrow(fuzzy_cov)
)
```

```{r fuzzy-coverage}
analysis_table(fuzzy_cov, "Benchmark candidate-pair coverage", max_rows = 30)
```

```{r fuzzy-reference}
analysis_table(fuzzy_reference, "Legacy fuzzy-tuning reference results")
```

```{r fuzzy-thresholds}
analysis_table(fuzzy_threshold, "Threshold sensitivity by method and pair source", max_rows = 40)
```
