---
title: "Spatial IV Benchmark"
format: gfm
execute:
  echo: true
  output: true
  warning: true
  message: true
---

```{r setup}
helper <- if (file.exists("analysis/_analysis_helpers.R")) "analysis/_analysis_helpers.R" else "../_analysis_helpers.R"
source(helper)
```

## Legacy prose retained with current results

Build spatial lags of dependent variables: `W_consY` and `W_giniY`. Rebuild the spatial lags of endogenous `EMIE` and IV: `W_EMIE` and `W_wLing`. Build the second-order lag `W2_wLing`. Rebuild spatial lags of exogenous controls. Spatial-2SLS for consumption change, also instrumenting `W_consY`.

The legacy comments left `summary(model_sdm2sls_cons, diagnostics = TRUE)` and `summary(model_sdm2sls_gini, diagnostics = TRUE)` commented out and concluded: Don't work even when diagnostics = FALSE.

Region clustered standard errors. `vcovCL(..., cluster = ~ region, data = joined_df)`. HC0 by default. Then `coeftest(model_sdm2sls_cons, vcov. = vcov_cluster_cons)` and `coeftest(model_sdm2sls_gini, vcov. = vcov_cluster_gini)`.

```{r spatial-iv-prose-deviation-note, results='asis'}
analysis_deviation_note("The current benchmark preserves the legacy experimental specifications and clustered-SE attempt, but separates ivreg object creation from diagnostic numerical suitability.")
```

```{r spatial-iv-targets}
spatial_iv_status <- analysis_target_csv("bench_spatial_iv_experimental", "spatial_iv_model_status.csv")
spatial_iv_diag <- analysis_target_csv("bench_spatial_iv_experimental", "spatial_iv_diagnostics_summary.csv")
spatial_iv_coef <- analysis_target_csv("bench_spatial_iv_experimental", "spatial_iv_coefficient_summary.csv")
spatial_iv_cluster <- analysis_target_csv("bench_spatial_iv_experimental", "spatial_iv_clustered_coefficient_summary.csv")
spatial_iv_fail <- analysis_target_csv("bench_spatial_iv_experimental", "spatial_iv_failure_summary.csv")
```

The current benchmark estimates `r analysis_format_number(nrow(spatial_iv_status), 0)` experimental specifications.  A status of `estimated` means that `ivreg()` returned an object, not that the model is suitable for the final paper.

```{r spatial-iv-status}
analysis_table(spatial_iv_status[setdiff(names(spatial_iv_status), "formula")], "Spatial-IV model status")
```

```{r spatial-iv-formulas}
spatial_iv_status[, c("model", "formula"), drop = FALSE]
```

```{r spatial-iv-diagnostics}
analysis_table(spatial_iv_diag, "IV diagnostic summaries")
analysis_table(spatial_iv_coef, "Default coefficient summaries")
analysis_table(spatial_iv_cluster, "Clustered-SE coeftest attempt")
analysis_table(spatial_iv_fail, "Failure/status summary")
```
