---
title: "Spatial IV Benchmark"
format: gfm
execute:
  echo: true
  output: true
  warning: true
  message: true
---

```{r setup}
helper <- if (file.exists("analysis/_analysis_helpers.R")) "analysis/_analysis_helpers.R" else "../_analysis_helpers.R"
source(helper)
```

## Legacy diagnostic intent

The legacy experimental spatial-IV chunk tried spatial lags of the dependent variable, `EMIE`, the linguistic-distance instrument, second-order instrument lags, and lagged controls.  The legacy comments said the attempts did not work even when IV diagnostics were disabled.  The current benchmark distinguishes between `ivreg()` returning an object and the model being numerically meaningful.

```{r spatial-iv-targets}
spatial_iv_status <- analysis_target_csv("bench_spatial_iv_experimental", "spatial_iv_model_status.csv")
spatial_iv_diag <- analysis_target_csv("bench_spatial_iv_experimental", "spatial_iv_diagnostics_summary.csv")
spatial_iv_coef <- analysis_target_csv("bench_spatial_iv_experimental", "spatial_iv_coefficient_summary.csv")
spatial_iv_cluster <- analysis_target_csv("bench_spatial_iv_experimental", "spatial_iv_clustered_coefficient_summary.csv")
spatial_iv_fail <- analysis_target_csv("bench_spatial_iv_experimental", "spatial_iv_failure_summary.csv")
```

The current benchmark estimates `r analysis_format_number(nrow(spatial_iv_status), 0)` experimental specifications.  A status of `estimated` means that `ivreg()` returned an object, not that the model is suitable for the final paper.

```{r spatial-iv-status}
analysis_table(spatial_iv_status[setdiff(names(spatial_iv_status), "formula")], "Spatial-IV model status")
```

```{r spatial-iv-formulas}
spatial_iv_status[, c("model", "formula"), drop = FALSE]
```

```{r spatial-iv-diagnostics}
analysis_table(spatial_iv_diag, "IV diagnostic summaries")
analysis_table(spatial_iv_coef, "Default coefficient summaries")
analysis_table(spatial_iv_cluster, "Clustered-SE coeftest attempt")
analysis_table(spatial_iv_fail, "Failure/status summary")
```
