\documentclass[letterpaper,12pt]{article}
\usepackage{graphicx} % Required for inserting images

\usepackage[
  left=.9in,
  right=.9in,
  top=1in,
  bottom=1in,
  footskip=30pt
]{geometry}

\usepackage{csquotes}
\MakeOuterQuote{"}
\usepackage{comment}

\pagenumbering{gobble}

\title{Coding Sample: \\ Escaping Inequality in India\vspace{-2mm}}

\author{Rishav Roy}
\date{}

\begin{document}

\maketitle
\vspace{-3mm}
\noindent What follows is the 138-page code appendix to my independent paper and writing sample, "Escaping Inequality in India: The Role of English-Medium Instruction." \\

\vspace{-2mm}
\noindent Due to the length of this code, the time-constrained reviewer is encouraged to prioritize these 47 pages, excerpted from the code chunks listed below them: 

\begin{enumerate}
\item[14-16:] Scalable \textbf{import, cleaning, and merging} of multi-file census and geospatial data.
\vspace{-2mm}
\begin{itemize}
\item Chunk 4: Import key data files
\end{itemize}
\vspace{-2mm}

\item[31-32:] \textbf{QA checks, disambiguation of identifiers} for district-level crosswalk.
\vspace{-2mm}
\begin{itemize}
\item Chunk 6: Match districts: Construct district tracking dfs
\end{itemize}
\vspace{-2mm}

\item[50-52:] \textbf{Testing for systematic missingness} using the Benjamini–Hochberg procedure and parallelized logistic regressions. 
\vspace{-2mm}
\begin{itemize}
\item Chunk 8: Participation model: Are NAs randomly distributed
\end{itemize}
\vspace{-2mm}

\item[55-57:] \textbf{Survey-weighted probit estimation}, average marginal effects derivation, and setup for sample selection correction under a complex survey design.
\vspace{-2mm}
\begin{itemize}
\item Chunk 9: Participation model: Estimation and IMR 
\vspace{-1mm}
\item Chunk 10: Participation model: Derive AME
\end{itemize}
\vspace{-2mm}

\item[80-88:] 
Reusable \textbf{record-linkage functions} using cascading fuzzy matches and unmatched/false-match diagnostics to harmonize district identifiers across years and data sources.
\vspace{-2mm}
\begin{itemize}
\item Chunk 16: Match districts: Test joining methods
\vspace{-1mm}
\item Chunk 17: Match districts: Define joining functions
\end{itemize}
\vspace{-2mm}

\item[105-107:] Reusable \textbf{IV specification and estimation pipeline} with clustered inference, plus contiguity-based spatial weights and setup for spatially lagged models.
\vspace{-2mm}
\begin{itemize}
\item Chunk 24: Geospatial: Neighbor list construction
\vspace{-1mm}
\item Chunk 25: 2SLS: Controls and 2SLS formula
\vspace{-1mm}
\item Chunk 26: 2SLS: Baseline 2SLS models
\end{itemize}
\vspace{-2mm}

\item[111-112:] \textbf{Robustness checks and diagnostics} for multicollinearity, spatial autocorrelation.
\vspace{-2mm}
\begin{itemize}
\item Chunk 28: Multicollinearity check
\vspace{-1mm}
\item Chunk 29: Geospatial: Spatial autocorrelation tests
\end{itemize}
\vspace{-2mm}

\item[117-138:] \textbf{Automated generation of publication-quality figures and tables} (including both summary statistics and regression outputs).
\vspace{-2mm}
\begin{itemize}
\item Chunks 31-40: [Output for paper]
\end{itemize}
\vspace{-2mm}

\end{enumerate}


\end{document}