---
title: "Long Paths and 8.3 Filename Troubleshooting"
format: gfm
execute:
  echo: true
  output: true
  warning: true
  message: true
---

```{r setup}
helper <- if (file.exists("analysis/_analysis_helpers.R")) "analysis/_analysis_helpers.R" else "../_analysis_helpers.R"
source(helper)
```

## Legacy prose retained with current results

---TROUBLESHOOTING--- If getting "file does not exist" or similar errors: switch the function used to read in the file; for example, if the relevant file is being read in using `read_sav()` or `read_sav_short()`, switch to `read_sav_short()` or `read_sav()` respectively.

If that didn't work and you're not using Windows: either modify this code chunk for your OS or manually shorten the error-inducing file paths.

Purpose of this chunk: to ensure R can identify and read in all necessary files. This project found me making frequent use of file paths over 260 characters in length. Out of my naivete, I didn't realize Windows File Explorer automatically shortens such file paths by rewriting them in the 8.3 filename convention---meaning R can no longer access their standard "long name." Conversely, functions like `read_sav()` automatically lengthen 8.3 filenames via functions like `normalizePath()`, turning the true 8.3 file path into a long name which, from R's and thus our perspective, is useless. This code chunk turns our long name inputs into 8.3 filenames while tricking the `read_*()`-type functions into not turning them back.

Goal: make alternative readers which accept 8.3 filenames. The current production implementation is in `R/io/read_long_paths.R`; this analysis note keeps the troubleshooting material out of production code while rendering runnable current-code analogs directly.

```{r long-path-prose-deviation-note, results='asis'}
analysis_deviation_note("The legacy comment included extensive Windows Registry and readr/vroom internals notes. The rendered note preserves the operational troubleshooting prose and runs the current production reader contract; full implementation details live in R/io/read_long_paths.R.")
```

```{r long-path-source-current-functions}
source(analysis_path("R", "io", "read_long_paths.R"))
read_with_short_path
get_windows_short_path
normalize_path_for_os
```

```{r long-path-current-reader-contract}
tmp <- tempfile(fileext = ".csv")
write.csv(data.frame(x = 1:2), tmp, row.names = FALSE)
read_csv_short(tmp)
data.frame(
  current_code_analog = c(
    "normalize_path_for_os(tmp)",
    "get_windows_short_path(tmp)",
    "read_csv_short(tmp)"
  ),
  result = c(
    normalize_path_for_os(tmp),
    get_windows_short_path(tmp),
    paste(names(read_csv_short(tmp)), collapse = ", ")
  )
)
```
