---
title: "District Matching Diagnostics"
format: gfm
execute:
  echo: true
  output: true
  warning: true
  message: true
---

```{r setup}
helper <- if (file.exists("analysis/_analysis_helpers.R")) "analysis/_analysis_helpers.R" else "../_analysis_helpers.R"
source(helper)
```

## Legacy prose retained with current results

Errors are fixed in the chunk "Match districts: Manually fix errors". Changed for diagnostic purposes.

To extract mismatches: the legacy chunk pulled the rows whose merged district/year fields were absent after `merge_dfs_into_tracker()`, tagged their source, and combined them for inspection. Compare outcomes with different tracker dfs. Identify errors for each tracker dr. Combine all errors. Create a df with all rows from original data, so I can search for close matches easily.

#### Correct even more NAs

The legacy row-count checks were: `flagged_df_timeseries %>% .[!complete.cases(.),] %>% nrow()` = 155 rows; `joined_df_timeseries %>% .[!complete.cases(.),] %>% nrow()` = 305 rows; `flagged_df_tracker %>% .[!complete.cases(.),] %>% nrow()` = 92 rows; `joined_df_tracker %>% .[!complete.cases(.),] %>% nrow()` = 250 rows. ***Why the gigantic discrepancy???

```{r district-matching-prose-deviation-note, results='asis'}
analysis_deviation_note("The legacy helper objects no longer exist in the active architecture. The rendered current-code analog counts incomplete rows in target-backed tracker/panel, source-inventory, unmatched-row, and many-to-many outputs.")
```

```{r district-matching-targets}
dm_summary <- analysis_target_csv("diag_ext_district_matching", "district_matching_summary.csv")
dm_reference <- analysis_target_csv("diag_ext_district_matching", "district_matching_legacy_reference.csv")
dm_tracker_panel <- analysis_target_csv("diag_ext_district_matching", "district_matching_tracker_panel_comparison.csv")
dm_key_comparison <- analysis_target_csv("diag_ext_district_matching", "district_matching_key_comparison.csv")
dm_key_roles <- analysis_target_csv("diag_ext_district_matching", "district_matching_key_role_counts.csv")
dm_source_inventory <- analysis_target_csv("diag_ext_district_matching", "district_matching_source_key_inventory.csv")
dm_unmatched <- analysis_target_csv("diag_ext_district_matching", "district_matching_unmatched_rows.csv")
dm_many <- analysis_target_csv("diag_ext_district_matching", "district_matching_many_to_many_cases.csv")
```

The current analog keeps true unmatched rows separate from fallback source-key inventory rows.  It reports `r analysis_format_number(analysis_value(dm_summary, column = "n_unmatched_rows"), 0)` true unmatched rows and `r analysis_format_number(analysis_value(dm_summary, column = "n_source_key_inventory_rows"), 0)` fallback source-key inventory rows.

```{r district-matching-summary}
analysis_table(dm_summary, "District-matching diagnostic summary")
analysis_table(dm_reference, "Legacy Chunk 20 diagnostic reference")
analysis_table(dm_tracker_panel, "Tracker/panel/intermediate row-count comparison")
```

The current analog of `flagged_df_timeseries %>% .[!complete.cases(.),] %>% nrow()` is to count incomplete rows in the active tracker/panel comparison table and source-key diagnostic tables rather than rerunning the deleted legacy helper.

```{r district-matching-current-code-analogs}
sapply(
  list(
    tracker_panel_comparison = dm_tracker_panel,
    source_key_inventory = dm_source_inventory,
    true_unmatched_rows = dm_unmatched,
    many_to_many_cases = dm_many
  ),
  function(x) sum(!stats::complete.cases(as.data.frame(x)))
)
```

```{r district-matching-keys}
if (all(c("panel_key_status", "join_key_status") %in% names(dm_key_comparison))) {
  key_summary <- as.data.frame(table(panel_key_status = dm_key_comparison$panel_key_status, join_key_status = dm_key_comparison$join_key_status), stringsAsFactors = FALSE)
  analysis_table(key_summary, "Panel-vs-join key-status summary")
} else {
  analysis_table(dm_key_comparison, "Panel-vs-join key-status summary")
}
analysis_table(dm_key_roles, "Panel/join key roles and interpretation")
```

```{r district-matching-details}
analysis_table(dm_source_inventory, "Fallback source-key inventory separated from true unmatched rows", max_rows = 20)
analysis_table(dm_unmatched, "True unmatched rows from current matching diagnostics", max_rows = 20)
analysis_table(dm_many, "Many-to-many cases from current matching diagnostics", max_rows = 20)
```
