---
title: "District Lineage Diagnostics"
format: gfm
execute:
  echo: true
  output: true
  warning: true
  message: true
---

```{r setup}
helper <- if (file.exists("analysis/_analysis_helpers.R")) "analysis/_analysis_helpers.R" else "../_analysis_helpers.R"
source(helper)
```

## Scope

```{r lineage-deviation, results='asis'}
analysis_deviation_note("This notebook reports the current production district-lineage system. The inherited harmonization panel is retained only for historical comparison and is not a validation target.")
```

The public analysis uses the reviewed 573-district primary panel. The 408-district conservative panel and 587-district full reviewed panel are rebuilt with the same measure, pooled-Gini, IV, and first-stage code.

```{r lineage-inputs}
summary <- analysis_target_csv("diag_ext_district_lineage", "summary.csv")
readiness <- analysis_target_csv("diag_ext_district_lineage", "readiness.csv")
blockers <- analysis_target_csv("diag_ext_district_lineage", "blockers.csv")
completion <- analysis_target_csv("diag_ext_district_lineage", "completion_status.csv")
inventory <- analysis_target_csv("diag_ext_district_lineage", "source_inventory.csv")
registry <- analysis_target_csv("diag_ext_district_lineage", "source_registry.csv")
reference_issues <- analysis_target_csv("diag_ext_district_lineage", "source_reference_issues.csv")
transition <- analysis_target_csv("diag_ext_district_lineage", "district_transition_2001_2011.csv")
allocation_validation <- analysis_target_csv("diag_ext_district_lineage", "allocation_weight_validation.csv")
reviewed_allocations <- analysis_target_csv("diag_ext_district_lineage", "adjudicated_allocation_weights.csv")
reviewed_allocation_validation <- analysis_target_csv("diag_ext_district_lineage", "adjudicated_allocation_validation.csv")
eligibility <- analysis_target_csv("diag_ext_district_lineage", "conservative_mapping_eligibility.csv")
conservative_crosswalk <- analysis_target_csv("diag_ext_district_lineage", "conservative_source_crosswalk.csv")
primary_crosswalk <- analysis_target_csv("diag_ext_district_lineage", "primary_source_crosswalk.csv")
full_reviewed_crosswalk <- analysis_target_csv("diag_ext_district_lineage", "full_reviewed_source_crosswalk.csv")
primary_reviews <- analysis_target_csv("diag_ext_district_lineage", "primary_reviews.csv")
multi_parent_queue <- analysis_target_csv("diag_ext_district_lineage", "multi_parent_review_queue.csv")
loss_audit <- analysis_target_csv("diag_ext_district_lineage", "district_loss_audit.csv")
reclassification <- analysis_target_csv("diag_ext_district_lineage", "identity_reclassification.csv")
geometry_qa <- analysis_target_csv("diag_ext_district_lineage", "geometry_2001_qa.csv")
geometry_coverage <- analysis_target_csv("diag_ext_district_lineage", "geometry_2001_unit_coverage.csv")
panel_variants <- analysis_target_csv("diag_ext_district_lineage", "panel_variant_summary.csv")
variant_models <- analysis_target_csv("diag_ext_lineage_panel_variants", "panel_variant_model_summary.csv")
variant_coefficients <- analysis_target_csv("diag_ext_lineage_panel_variants", "panel_variant_coefficients.csv")
variant_first_stage <- analysis_target_csv("diag_ext_lineage_panel_variants", "panel_variant_first_stage.csv")
variant_ginis <- analysis_target_csv("diag_ext_lineage_panel_variants", "panel_variant_gini_reconstruction.csv")
legacy_comparison <- analysis_target_csv("diag_ext_legacy_crosswalk_comparison", "legacy_crosswalk_comparison.csv")
downstream_gates <- analysis_target_csv("diag_ext_lineage_downstream", "downstream_review_gates.csv")
```

## Production readiness

```{r lineage-summary}
analysis_table(summary, "District-lineage summary")
analysis_table(readiness, "Current lineage invariants")
analysis_table(blockers, "Current lineage blockers")
analysis_table(completion, "Six-step completion status")
analysis_table(inventory, "Input inventory", max_rows = 60)
analysis_table(registry, "Evidence registry", max_rows = 60)
analysis_table(reference_issues, "Unregistered evidence references", max_rows = 50)
```

Production readiness depends only on current lineage invariants. Historical crosswalk comparisons are generated separately and cannot block the public panel.

## Transition and allocation integrity

```{r lineage-transition}
analysis_table(transition, "Census 2011 to Census 2001 transitions", max_rows = 60)
analysis_table(allocation_validation, "Generated allocation checks", max_rows = 60)
analysis_table(reviewed_allocation_validation, "Reviewed allocation checks", max_rows = 60)
analysis_table(reviewed_allocations, "Reviewed allocations", max_rows = 60)
```

## Three panel specifications

```{r lineage-panels}
analysis_table(panel_variants, "Panel definitions", max_rows = 10)
analysis_table(variant_models, "Analysis-sample comparison", max_rows = 10)
analysis_table(variant_first_stage, "First-stage comparison", max_rows = 30)
analysis_table(variant_coefficients, "2SLS coefficient comparison", max_rows = 60)
analysis_table(variant_ginis, "Pooled-Gini reconstruction comparison", max_rows = 30)
analysis_table(conservative_crosswalk, "Conservative crosswalk", max_rows = 50)
analysis_table(primary_crosswalk, "Primary crosswalk", max_rows = 50)
analysis_table(full_reviewed_crosswalk, "Full reviewed crosswalk", max_rows = 50)
analysis_table(primary_reviews, "Reviewed primary-panel additions", max_rows = 50)
```

## Remaining bounded review work

```{r lineage-review}
analysis_table(loss_audit, "Census-2001 district accounting", max_rows = 60)
analysis_table(reclassification, "NSS-75 identity classes", max_rows = 60)
analysis_table(multi_parent_queue, "Multi-parent fractional-allocation review queue", max_rows = 60)
analysis_table(geometry_qa, "Census-2001 geometry QA")
analysis_table(geometry_coverage, "Geometry unit coverage", max_rows = 60)
```

The 21 multi-parent identities remain confined to the full reviewed sensitivity panel until official territorial evidence validates their shares. Six Census-2001 districts lack 2007–08 support.

## Historical comparison

```{r lineage-legacy}
analysis_table(legacy_comparison, "Archived legacy crosswalk comparison", max_rows = 60)
analysis_table(downstream_gates, "Historical downstream comparison checks", max_rows = 50)
```

The legacy panel and archived review ledger are loaded only by optional comparison targets.
