---
title: "District Tracker Source Diagnostics"
format: gfm
execute:
  echo: true
  output: true
  warning: true
  message: true
---

```{r setup}
helper <- if (file.exists("analysis/_analysis_helpers.R")) "analysis/_analysis_helpers.R" else "../_analysis_helpers.R"
source(helper)
```

## Legacy prose retained with current results

The legacy tracker-source diagnostics compared district names across years, inspected state and union-territory changes, noted unrecorded state changes requiring manual attention, and checked same-name districts that appeared across states. The analysis note keeps the legacy recorded-change, unrecorded-change, in-period district-name-change, and same-name-district benchmarks beside the current row-level outputs so differences in counting level are visible.

```{r tracker-prose-deviation-note, results='asis'}
analysis_deviation_note("The current source diagnostics report row-level outputs, event-level summaries, and per-year same-name-district counts beside the legacy expected-event benchmarks, because active tracker sources can count more rows than the legacy comments' event-level summaries.")
```

```{r tracker-targets}
tracker_counts <- analysis_target_csv("diag_ext_district_tracker_sources", "tracker_source_counts.csv")
tracker_state_changes <- analysis_target_csv("diag_ext_district_tracker_sources", "tracker_state_changes.csv")
tracker_state_events <- analysis_target_csv("diag_ext_district_tracker_sources", "tracker_state_change_events.csv")
tracker_expected_state <- analysis_target_csv("diag_ext_district_tracker_sources", "tracker_legacy_expected_state_changes.csv")
tracker_unrecorded <- analysis_target_csv("diag_ext_district_tracker_sources", "tracker_unrecorded_state_changes.csv")
tracker_inperiod <- analysis_target_csv("diag_ext_district_tracker_sources", "tracker_inperiod_district_changes.csv")
tracker_expected_inperiod <- analysis_target_csv("diag_ext_district_tracker_sources", "tracker_legacy_expected_inperiod_district_changes.csv")
tracker_same <- analysis_target_csv("diag_ext_district_tracker_sources", "tracker_same_name_districts.csv")
tracker_same_by_year <- analysis_target_csv("diag_ext_district_tracker_sources", "tracker_same_name_districts_by_year.csv")
tracker_expected_same <- analysis_target_csv("diag_ext_district_tracker_sources", "tracker_legacy_expected_same_name_districts.csv")
```

```{r tracker-current-count-analogs}
data.frame(
  current_code_analog = c(
    "nrow(tracker_state_events)",
    "nrow(tracker_expected_state)",
    "nrow(tracker_inperiod)",
    "nrow(tracker_expected_inperiod)",
    "nrow(tracker_same_by_year[tracker_same_by_year$within_legacy_range, ])"
  ),
  rows = c(
    nrow(tracker_state_events),
    nrow(tracker_expected_state),
    nrow(tracker_inperiod),
    nrow(tracker_expected_inperiod),
    if ("within_legacy_range" %in% names(tracker_same_by_year)) sum(tracker_same_by_year$within_legacy_range %in% TRUE) else NA_integer_
  )
)
```

```{r tracker-tables}
analysis_table(tracker_counts, "Tracker source row counts")
analysis_table(tracker_state_changes, "Current row-level state/UT changes")
analysis_table(tracker_state_events, "Current state/UT change events")
analysis_table(tracker_expected_state, "Legacy expected recorded state/UT changes")
analysis_table(tracker_unrecorded, "Legacy unrecorded state/UT changes requiring manual attention")
analysis_table(tracker_inperiod, "Current district-name changes inside sampling periods")
analysis_table(tracker_expected_inperiod, "Legacy in-period district-name-change benchmark")
analysis_table(tracker_same, "Current same-name districts appearing in multiple states", max_rows = 30)
analysis_table(tracker_same_by_year, "Current same-name districts per year compared to legacy 6-10 range")
analysis_table(tracker_expected_same, "Legacy same-name-district benchmark")
```
