---
title: "District Lineage v2 Diagnostics"
format: gfm
execute:
  echo: true
  output: true
  warning: true
  message: true
---

```{r setup}
helper <- if (file.exists("analysis/_analysis_helpers.R")) "analysis/_analysis_helpers.R" else "../_analysis_helpers.R"
source(helper)
```

## Scope

```{r lineage-v2-deviation, results='asis'}
analysis_deviation_note("This notebook is a new current-code diagnostic rather than preserved legacy prose. It does not treat the inherited 454-district panel or its legacy IV estimates as validation targets. The legacy matching was almost certainly flawed, and the legacy partial F-statistic of 37.77 was computed incorrectly.")
```

The v2 pipeline runs beside the current reviewed crosswalk. It does not replace
production geography until all source identities and administrative events used
by the preferred panel are adjudicated. Exact names and fuzzy scores generate
review candidates. Only tracked code, alias, or dated-lineage adjudications can
accept a source identity.

```{r lineage-v2-inputs}
summary <- analysis_target_csv("diag_ext_district_lineage_v2", "summary.csv")
readiness <- analysis_target_csv("diag_ext_district_lineage_v2", "migration_readiness.csv")
inventory <- analysis_target_csv("diag_ext_district_lineage_v2", "source_inventory.csv")
source_registry <- analysis_target_csv("diag_ext_district_lineage_v2", "source_registry.csv")
source_reference_issues <- analysis_target_csv("diag_ext_district_lineage_v2", "source_reference_issues.csv")
bridge_summary <- analysis_target_csv("diag_ext_district_lineage_v2", "shrid_bridge_summary.csv")
bridge_qa <- analysis_target_csv("diag_ext_district_lineage_v2", "shrid_bridge_qa.csv")
weight_validation <- analysis_target_csv("diag_ext_district_lineage_v2", "allocation_weight_validation.csv")
adjudicated_weights <- analysis_target_csv("diag_ext_district_lineage_v2", "adjudicated_allocation_weights.csv")
adjudicated_weight_validation <- analysis_target_csv("diag_ext_district_lineage_v2", "adjudicated_allocation_validation.csv")
transition <- analysis_target_csv("diag_ext_district_lineage_v2", "district_transition_2001_2011.csv")
eligibility <- analysis_target_csv("diag_ext_district_lineage_v2", "primary_mapping_eligibility.csv")
primary_crosswalk <- analysis_target_csv("diag_ext_district_lineage_v2", "primary_source_crosswalk.csv")
excluded_sources <- analysis_target_csv("diag_ext_district_lineage_v2", "excluded_source_rows.csv")
candidates <- analysis_target_csv("diag_ext_district_lineage_v2", "source_match_candidates.csv")
events <- analysis_target_csv("diag_ext_district_lineage_v2", "candidate_admin_events.csv")
current_components <- analysis_target_csv("diag_ext_district_lineage_v2", "current_component_registry.csv")
urban_coverage <- analysis_target_csv("diag_ext_district_lineage_v2", "current_urban_coverage.csv")
components <- analysis_target_csv("diag_ext_district_lineage_v2", "changed_component_roster.csv")
evidence <- analysis_target_csv("diag_ext_district_lineage_v2", "evidence_requests.csv")
duplicates <- analysis_target_csv("diag_ext_district_lineage_v2", "duplicate_keys.csv")
gold <- analysis_target_csv("diag_ext_district_lineage_v2", "match_gold_scored.csv")
gold_summary <- analysis_target_csv("diag_ext_district_lineage_v2", "match_gold_summary.csv")
```

## Build and migration status

```{r lineage-v2-summary}
analysis_table(summary, "District-lineage v2 summary")
analysis_table(readiness, "Production-migration gates")
analysis_table(inventory, "Available and missing lineage inputs", max_rows = 60)
analysis_table(source_registry, "District-lineage evidence registry", max_rows = 50)
analysis_table(source_reference_issues, "Unregistered or missing adjudication evidence", max_rows = 50)
```

A failed migration gate is expected while the adjudication files are incomplete.
The diagnostic remains intentionally separate from the current production panel.

## Census transition and allocation integrity

```{r lineage-v2-transition}
analysis_table(bridge_summary, "SHRUG bridge-status summary", max_rows = 50)
analysis_table(bridge_qa, "Non-deterministic or incomplete SHRID bridge records", max_rows = 50)
analysis_table(weight_validation, "SHRUG source-district allocation checks", max_rows = 50)
analysis_table(adjudicated_weight_validation, "Tracked sensitivity-allocation checks", max_rows = 50)
analysis_table(adjudicated_weights, "Tracked sensitivity allocations", max_rows = 50)
analysis_table(transition, "Census 2011 to Census 2001 district transitions", max_rows = 50)
analysis_table(duplicates, "Duplicate-key diagnostics", max_rows = 50)
```

SHRID units that cross district boundaries or lack either Census membership are
not silently renormalized away. Their omission lowers coverage and prevents a
transition from being classified as deterministic.

## Review queues

```{r lineage-v2-review}
analysis_table(eligibility, "Primary-panel source eligibility", max_rows = 50)
analysis_table(primary_crosswalk, "Adjudicated deterministic source-to-2001 crosswalk", max_rows = 50)
analysis_table(excluded_sources, "Excluded or unresolved source rows", max_rows = 50)
analysis_table(candidates, "Top source-match candidates; never automatic adjudications", max_rows = 50)
analysis_table(events, "Candidate administrative events", max_rows = 50)
analysis_table(current_components, "Current LGD subdistrict and urban-local-body registry", max_rows = 50)
analysis_table(urban_coverage, "Current LGD urban coverage bridge", max_rows = 50)
analysis_table(components, "LGD changed-component roster", max_rows = 50)
analysis_table(evidence, "Targeted official-evidence requests", max_rows = 50)
```

## Fuzzy-name calibration

```{r lineage-v2-gold}
analysis_table(gold_summary, "Hand-reviewed candidate-rule summary")
analysis_table(gold, "Hand-reviewed fuzzy-match cases", max_rows = 50)
```

The calibration table is intentionally small and conservative. It documents
known spelling variants and hard negatives, but it is not large enough to
establish a 99.5 percent automatic-match precision claim. Candidate thresholds
therefore remain review aids rather than production decision rules.
