---
title: "District Tracker Source Diagnostics"
format: gfm
---
```{r setup, include=FALSE}
helper <- if (file.exists("analysis/_analysis_helpers.R")) "analysis/_analysis_helpers.R" else "../_analysis_helpers.R"
source(helper)
```

This notebook carries forward the long district-tracker source QA comments from legacy Chunk 6. It renders the current diagnostic outputs instead of copying counts from comments by hand.

The legacy workflow used these diagnostics to justify cross-year district matching, identify recorded and unrecorded state/UT changes, count district names changing inside multi-year survey periods, and check same-name districts across states.

```{r source-counts, echo=FALSE, results='asis'}
analysis_table(read_analysis_csv("diagnostics", "extended", "district_tracker_sources", "tracker_source_counts.csv"), "Raw district-change source coverage")
```

```{r legacy-reference, echo=FALSE, results='asis'}
analysis_table(read_analysis_csv("diagnostics", "extended", "district_tracker_sources", "tracker_legacy_comment_reference.csv"), "Legacy comment benchmarks versus current detections")
```

```{r state-changes, echo=FALSE, results='asis'}
analysis_table(read_analysis_csv("diagnostics", "extended", "district_tracker_sources", "tracker_state_change_events.csv"), "Current recorded state/UT change events")
```

```{r expected-state-changes, echo=FALSE, results='asis'}
analysis_table(read_analysis_csv("diagnostics", "extended", "district_tracker_sources", "tracker_legacy_expected_state_changes.csv"), "Recorded state/UT changes noted in the legacy comments")
```

```{r unrecorded-state-changes, echo=FALSE, results='asis'}
analysis_table(read_analysis_csv("diagnostics", "extended", "district_tracker_sources", "tracker_unrecorded_state_changes.csv"), "Unrecorded state/UT changes requiring manual attention")
```

```{r inperiod-changes, echo=FALSE, results='asis'}
analysis_table(read_analysis_csv("diagnostics", "extended", "district_tracker_sources", "tracker_inperiod_district_changes.csv"), "Current district-name changes inside sampling periods")
```

```{r expected-inperiod, echo=FALSE, results='asis'}
analysis_table(read_analysis_csv("diagnostics", "extended", "district_tracker_sources", "tracker_legacy_expected_inperiod_district_changes.csv"), "Legacy in-period district-name-change benchmark")
```

```{r same-name-districts, echo=FALSE, results='asis'}
analysis_table(read_analysis_csv("diagnostics", "extended", "district_tracker_sources", "tracker_same_name_districts.csv"), "Same-name districts appearing in multiple states")
```
