---
title: "Fuzzy Matching Diagnostics"
format: gfm
execute:
  echo: true
  output: true
  warning: true
  message: true
---

```{r setup}
helper <- if (file.exists("analysis/_analysis_helpers.R")) "analysis/_analysis_helpers.R" else "../_analysis_helpers.R"
source(helper)
```

## Legacy prose retained with current results

The legacy fuzzy-matching chunk tested string-distance methods, inspected troublesome pairs, and compared thresholds before fixing the production thresholds. Some sample pairs included `Baleshwar`/`Balasore`, `Jammu & Kashmir`/`Jammu and Kashmir`, `East Godavari`/`Godavari East`, `Sikim`/`Sikkim`, and long truncated district names such as `Sahibzada Ajit Singh Nag*`/`Sahibzada Ajit Singh Nagar`.

The final helper vectors were `methods <- c("soundex", "qgram", "jw", "dl", "osa")` and `thresholds <- c(0, 0, 0.15, 2, 1)`. `soundex = 0` means phonetic variants in anglicization allowed. `qgram = 0` means rearrangements of words allowed. `jw <= 0.15` means respellings and vowel swaps with 0.85 similarity allowed. `dl <= 2` means no more than two insertions, deletions, substitutions, and transpositions allowed. `osa <= 1` means one typo allowed.

```{r fuzzy-diagnostic-prose-deviation-note, results='asis'}
analysis_deviation_note("The current diagnostic renders active target outputs rather than legacy View() calls, but preserves the sample-pair and threshold prose in the note body.")
```

```{r fuzzy-diagnostic-targets}
fuzzy_summary <- analysis_target_csv("diag_ext_fuzzy_matching", "fuzzy_matching_summary.csv")
fuzzy_methods <- analysis_target_csv("diag_ext_fuzzy_matching", "fuzzy_matching_legacy_methods.csv")
fuzzy_tuning <- analysis_target_csv("diag_ext_fuzzy_matching", "fuzzy_matching_legacy_tuning_reference.csv")
fuzzy_coverage <- analysis_target_csv("diag_ext_fuzzy_matching", "fuzzy_matching_candidate_pair_coverage.csv")
fuzzy_trouble <- analysis_target_csv("diag_ext_fuzzy_matching", "fuzzy_matching_troublesome_pairs.csv")
fuzzy_status <- analysis_target_csv("diag_ext_fuzzy_matching", "fuzzy_matching_join_status_counts.csv")
```

```{r fuzzy-diagnostic-tables}
analysis_table(fuzzy_summary, "Fuzzy matching diagnostic summary")
analysis_table(fuzzy_methods, "Legacy final string-distance methods and thresholds")
analysis_table(fuzzy_tuning, "Legacy tuning comments retained as benchmarks")
analysis_table(fuzzy_coverage, "Candidate-pair coverage", max_rows = 30)
analysis_table(fuzzy_trouble, "Troublesome legacy name-pair checks")
analysis_table(fuzzy_status, "Current join status counts")
```
