---
title: "Missingness Diagnostics"
format: gfm
execute:
  echo: true
  output: true
  warning: true
  message: true
---

```{r setup}
helper <- if (file.exists("analysis/_analysis_helpers.R")) "analysis/_analysis_helpers.R" else "../_analysis_helpers.R"
source(helper)
```

## Legacy prose retained with current results

#### All variables: Variables/Regions with NAs

Variables with NAs. Only variables in the probit. Regions with the _most_ NAs.

The legacy case study inspected Rajasthan's Southern region and concluded: Southern region of Rajasthan: People with one missing cost variable often have ones which aren't missing. The person as a whole wasn't excluded from the data. Explanations: Surveyor messed up writing some costs but not others. Or surveyor meant to put a 0.

Potential chi-square tests for independence: ANY NA vs. state; ANY NA vs region. The target output preserves these checks as notes rather than rerunning exploratory `View()` calls.

#### Correlation matrix

For all observations: make new data frame with missingness indicator variables, expand factors to dummy variables, and compute the correlation matrix using pairwise-complete observations. For enrolled observations: repeat the same logic for missing variables defined only for enrolled children.

#### Logistic per missing variable (parallelized)

One logistic per missing variable. Summarize coefficient significance and overall fit. So `n_sig` = number of predictors which survive FDR at 0.05 significance (i.e., number of covariates associated with missingness).

```{r missingness-prose-deviation-note, results='asis'}
analysis_deviation_note("The rendered note preserves the legacy diagnostic order and prose but replaces interactive View() calls and platform-specific parallelization with target-backed CSVs, heatmaps, and pseudo-R-squared figures.")
```

```{r missingness-targets}
missing_counts <- analysis_target_csv("diag_ext_missingness", "missingness_counts.csv")
regional_cost <- analysis_target_csv("diag_ext_missingness", "regional_missingness_cost.csv")
regional_distance <- analysis_target_csv("diag_ext_missingness", "regional_missingness_distance.csv")
regional_father <- analysis_target_csv("diag_ext_missingness", "regional_missingness_father_education.csv")
logit_summary <- analysis_target_csv("diag_ext_missingness", "missingness_logit_summary.csv")
legacy_notes <- analysis_target_csv("diag_ext_missingness", "missingness_legacy_notes.csv")
corr_all_pairs <- analysis_target_csv("diag_ext_missingness", "missingness_correlation_all_top_pairs.csv")
corr_enrolled_pairs <- analysis_target_csv("diag_ext_missingness", "missingness_correlation_enrolled_top_pairs.csv")
```

The current active data contain `r analysis_format_number(analysis_metric(missing_counts, "missing_var", "Total probit-relevant with NA", "n_missing"), 0)` probit-relevant rows with at least one missing value and `r analysis_format_number(analysis_metric(missing_counts, "missing_var", "Total probit-relevant with no NA", "n_missing"), 0)` probit-relevant rows with no missing value.

```{r missingness-current-code-analog}
missing_counts[order(-missing_counts$n_missing), c("missing_var", "n_missing", "pct_missing"), drop = FALSE]
```

```{r missingness-counts}
analysis_table(missing_counts, "Current missingness counts")
```

```{r missingness-regions}
analysis_table(regional_cost, "Regions with the most cost-variable missingness", max_rows = 20)
analysis_table(regional_distance, "Regions with the most distance-to-school missingness", max_rows = 20)
analysis_table(regional_father, "Regions with the most father-education missingness", max_rows = 20)
```

The current analog of the legacy correlation-matrix work writes both CSV matrices and rendered heatmaps as target outputs.  The legacy logit-screen plots are likewise rendered from the current target output rather than copied from the old notebook.

```{r missingness-correlation-figures, results='asis'}
analysis_image("diag_ext_missingness", "missingness_correlation_all.png", "Missingness correlation heatmap for all observations")
analysis_image("diag_ext_missingness", "missingness_correlation_enrolled.png", "Missingness correlation heatmap for enrolled observations")
analysis_image("diag_ext_missingness", "missingness_logit_pseudo_r2.png", "Pseudo-R-squared bar chart for missingness logit screens")
```

```{r missingness-correlation-tables}
analysis_table(corr_all_pairs, "Largest absolute missingness correlations: all observations", max_rows = 20)
analysis_table(corr_enrolled_pairs, "Largest absolute missingness correlations: enrolled observations", max_rows = 20)
```

The current analog of the legacy one-logit-per-missing-variable screen is summarized below.  `n_sig` counts covariates that survive Benjamini-Hochberg/FDR adjustment at the target's configured threshold.

```{r missingness-logits}
analysis_table(logit_summary, "Logit summaries for structured missingness")
analysis_table(legacy_notes, "Legacy missingness notes retained as diagnostics")
```
