---
title: "Spatial Autocorrelation Diagnostics"
format: gfm
execute:
  echo: true
  output: true
  warning: true
  message: true
---

```{r setup}
helper <- if (file.exists("analysis/_analysis_helpers.R")) "analysis/_analysis_helpers.R" else "../_analysis_helpers.R"
source(helper)
```

## Legacy diagnostic intent

The legacy spatial-autocorrelation chunk used rook contiguity, a binary adjacency matrix, row-standardized weights, and `moran.test(..., zero.policy = TRUE)`.  It checked IV residuals, first-stage residuals, `EMIE`, linguistic distance, consumption growth, and Gini growth, while also leaving a Monte Carlo Moran scaffold commented out.

```{r moran-target}
moran <- analysis_target_csv("diag_public_spatial_autocorrelation", "spatial_moran_tests.csv")
```

The current `m_cons_resid` p-value is `r analysis_format_number(analysis_metric(moran, "legacy_name", "m_cons_resid", "p.value"), 12)`, and the current `m_cons` p-value is `r analysis_format_number(analysis_metric(moran, "legacy_name", "m_cons", "p.value"), 12)`.  These are current active-pipeline results, not hard-coded legacy comments.

```{r moran-table}
analysis_table(moran, "Current Moran's I diagnostics")
```
