---
title: "District Matching and Harmonization"
format:
  html: default
  pdf: default
---

```{r}
library(targets)
```

## Purpose

This note documents why district matching is necessary, how external district trackers are combined, where they disagree, and how manual corrections are applied.

## Core distinction

The matching problem mixes several very different phenomena: political district changes, renamings, spelling variants, typos, splits, mergers, and source inconsistencies. The code should not treat all of these as mere fuzzy-string problems.

## Sources

This note draws on the district-change files read by `read_district_change_sources()` and on `data/metadata/manual_district_corrections.csv`.

## Diagnostics

Generated diagnostics should include unmatched districts, many-to-many joins, tracker-source disagreements, threshold sensitivity, and manually applied corrections.

```{r}
# Example after pipeline exists:
# tar_read(diag_district_tracker_sources)
# tar_read(diag_district_matching)
```

## Current research implication

This note is meant to directly address the concern that the district matching must be more transparent before the exclusion restriction can be defended with geography controls and state fixed effects.
