---
title: "Instrument Exploration"
format: gfm
execute:
  echo: true
  output: true
  warning: true
  message: true
---

```{r setup}
helper <- if (file.exists("analysis/_analysis_helpers.R")) "analysis/_analysis_helpers.R" else "../_analysis_helpers.R"
source(helper)
```

## Legacy prose retained with current results

Preliminary test of IV strength: Dotplot of EMIE values by `district_code`. These are the current target-backed instrument-strength plots for the legacy exploratory check.

EMIE has three peaks: in Jammu and Kashmir; in Sikkim, Arunachal Pradesh, Nagaland, Manipur, Mizoran, Tripura, maybe Meghalaya; and in Andhra Pradesh, Karnataka, Goa, Lakshadweep, Kerala, Tamil Nadu, Pondicheri, and Andaman & Nicobar! The regions which historically were the furthest from Hindi! Many districts in the second group seem to have EMIE around 1, and the range of EMIE outside peaks is between 0.4 and 0.1. Justification for looking at smaller units of analysis?

Process the dataframe and compute the weighted average `ling_distance` for each `(state, district)` group. Group by state and district to ensure that same district names in different states are treated separately. For each group, keep only the top three rows by `spkr_tot`. Create the `ling_degrees` column based on the mother tongue values and @shastry2012a's 0-5 measure of degrees of linguistic distance. Calculate the weighted average linguistic distance for each group.

```{r instrument-prose-deviation-note, results='asis'}
analysis_deviation_note("The active note renders the current EMIE dotplot target and IV-panel diagnostics, while preserving the legacy interpretation of the peaks as prose rather than copied static output.")
```

```{r instrument-targets}
iv_match <- read_analysis_csv("diagnostics", "public", "iv_panel_match_summary.csv")
iv_state <- read_analysis_csv("diagnostics", "public", "iv_panel_state_summary.csv")
iv_rows <- read_analysis_csv("diagnostics", "public", "iv_summary_keyed_rows.csv")
iv_dotplot <- analysis_target_csv("diag_ext_instrument_exploration", "instrument_strength_dotplot_data.csv")
iv_notes <- analysis_target_csv("diag_ext_instrument_exploration", "instrument_exploration_legacy_notes.csv")
```

The current analog of the legacy `ggplot(..., aes(x = district_code_0708, y = EMIE, color = district_prefix)) + geom_point()` check uses `r analysis_format_number(nrow(iv_dotplot), 0)` active district-panel rows.

```{r instrument-current-code-analog}
instrument_dotplot_rows <- iv_dotplot[, intersect(c("district_order", "district_code", "state", "district", "EMIE", "wavg_ling_degrees", "state_prefix"), names(iv_dotplot)), drop = FALSE]
analysis_table(instrument_dotplot_rows, "Current EMIE-by-district dotplot data", max_rows = 30)
```

```{r instrument-dotplot, results='asis'}
analysis_image("diag_ext_instrument_exploration", "emie_by_district_dotplot.png", "Current EMIE-by-district exploratory dotplot")
```

```{r instrument-tables}
analysis_table(iv_match, "Current IV-panel match summary")
analysis_table(iv_state, "Current IV-panel state summary", max_rows = 30)
analysis_table(iv_rows, "Current keyed IV summary rows", max_rows = 30)
analysis_table(iv_notes, "Legacy instrument-exploration notes retained as target output")
```
