---
title: "Instrument Exploration"
format: gfm
execute:
  echo: true
  output: true
  warning: true
  message: true
---

```{r setup}
helper <- if (file.exists("analysis/_analysis_helpers.R")) "analysis/_analysis_helpers.R" else "../_analysis_helpers.R"
source(helper)
```

## Legacy diagnostic intent

The legacy IV-construction chunk included exploratory checks of the 2001 instrument, district joins, and potential instrument-strength plots that were not used as final paper outputs.  The current note reads the parity-audit diagnostics produced by the active pipeline instead of copying exploratory values from deleted code.

```{r instrument-targets}
iv_match <- read_analysis_csv("diagnostics", "public", "iv_panel_match_summary.csv")
iv_state <- read_analysis_csv("diagnostics", "public", "iv_panel_state_summary.csv")
iv_rows <- read_analysis_csv("diagnostics", "public", "iv_summary_keyed_rows.csv")
iv_dotplot <- analysis_target_csv("diag_ext_instrument_exploration", "instrument_strength_dotplot_data.csv")
iv_notes <- analysis_target_csv("diag_ext_instrument_exploration", "instrument_exploration_legacy_notes.csv")
```

The current analog of the legacy `ggplot(..., aes(x = district_code_0708, y = EMIE, color = district_prefix)) + geom_point()` check uses `r analysis_format_number(nrow(iv_dotplot), 0)` active district-panel rows.

```{r instrument-current-code-analog}
instrument_dotplot_rows <- iv_dotplot[, intersect(c("district_order", "district_code", "state", "district", "EMIE", "wavg_ling_degrees", "state_prefix"), names(iv_dotplot)), drop = FALSE]
analysis_table(instrument_dotplot_rows, "Current EMIE-by-district dotplot data", max_rows = 30)
```

```{r instrument-dotplot, results='asis'}
analysis_image("diag_ext_instrument_exploration", "emie_by_district_dotplot.png", "Current EMIE-by-district exploratory dotplot")
```

```{r instrument-tables}
analysis_table(iv_match, "Current IV-panel match summary")
analysis_table(iv_state, "Current IV-panel state summary", max_rows = 30)
analysis_table(iv_rows, "Current keyed IV summary rows", max_rows = 30)
analysis_table(iv_notes, "Legacy instrument-exploration notes retained as target output")
```
