---
title: "Instrument Exploration"
format: gfm
execute:
  echo: true
  output: true
  warning: true
  message: true
---

```{r setup}
helper <- if (file.exists("analysis/_analysis_helpers.R")) "analysis/_analysis_helpers.R" else "../_analysis_helpers.R"
source(helper)
```

## Current instrument and treatment diagnostics

The active construction uses the full mutually exclusive Census 2001 C-16 mother-tongue distribution. Language-group subtotal rows are removed before aggregation, linguistic distances are resolved mother-tongue-first from the documented Shastry concordance, and unmapped speaker mass remains visible. The preferred public scalar is `ling_distance_nonzero_mean`, the speaker-weighted mean distance among mapped languages with distance greater than zero. Native Hindi and Urdu are treated as the zero-distance reference and their mother-tongue shares are reported separately; distinct leaves inside the Census Hindi language group do not inherit zero distance.

The preferred public treatment is `emi_exposure_all_children_0708`, the survey-weighted share of all children ages 5-19 who are both enrolled and studying in English medium. The historical `EMIE` field remains a compatibility measure of EMI among enrolled children but is not used in the current district-level dotplot.

The tables below diagnose how the preferred exploratory first stage changes with six-region fixed effects, state fixed effects, main and expanded Census control sets, sequential thematic blocks, VIF/GVIF, state deletion, and district influence. These diagnostics do not change the public IV specification.

```{r instrument-prose-deviation-note, results='asis'}
analysis_deviation_note("The active note describes the same all-child EMI exposure and full-distribution linguistic-distance scalar used by the public IV specification. Historical comments are retained only as provenance.")
```

```{r instrument-targets}
iv_match <- read_analysis_csv("diagnostics", "public", "iv_panel_match_summary.csv")
iv_state <- read_analysis_csv("diagnostics", "public", "iv_panel_state_summary.csv")
iv_rows <- read_analysis_csv("diagnostics", "public", "iv_summary_keyed_rows.csv")
iv_dotplot <- analysis_target_csv("diag_ext_instrument_exploration", "instrument_strength_dotplot_data.csv")
iv_notes <- analysis_target_csv("diag_ext_instrument_exploration", "instrument_exploration_legacy_notes.csv")
first_stage_absorption <- analysis_target_csv("diag_ext_first_stage_absorption", "first_stage_absorption_ladder.csv")
first_stage_support <- analysis_target_csv("diag_ext_first_stage_absorption", "first_stage_absorption_common_support.csv")
first_stage_state_ranges <- analysis_target_csv("diag_ext_first_stage_absorption", "first_stage_state_residual_ranges.csv")
first_stage_state_deletion <- analysis_target_csv("diag_ext_first_stage_absorption", "first_stage_state_deletion.csv")
first_stage_district_influence <- analysis_target_csv("diag_ext_first_stage_absorption", "first_stage_district_influence.csv")
first_stage_vif <- analysis_target_csv("diag_ext_first_stage_absorption", "first_stage_vif.csv")
alternative_distance_summary <- analysis_target_csv("diag_ext_alternative_distance_first_stages", "alternative_distance_first_stage_summary.csv")
alternative_distance_coefficients <- analysis_target_csv("diag_ext_alternative_distance_first_stages", "alternative_distance_first_stage_coefficients.csv")
alternative_distance_support <- analysis_target_csv("diag_ext_alternative_distance_first_stages", "alternative_distance_first_stage_common_support.csv")
mapping_coverage_sensitivity <- analysis_target_csv("diag_ext_alternative_distance_first_stages", "alternative_distance_mapping_coverage_sensitivity.csv")
distance4_languages <- analysis_target_csv("diag_ext_alternative_distance_first_stages", "distance4_language_decomposition.csv")
unmapped_languages <- analysis_target_csv("diag_ext_alternative_distance_first_stages", "unmapped_language_decomposition.csv")
distance4_leave_one_out <- analysis_target_csv("diag_ext_alternative_distance_first_stages", "distance4_leave_one_language_out.csv")
weak_iv_outcomes <- analysis_target_csv("diag_ext_alternative_distance_first_stages", "alternative_distance_weak_iv_outcomes.csv")
iv_diagnostic_applicability <- analysis_target_csv("diag_ext_alternative_distance_first_stages", "iv_diagnostic_applicability.csv")
iv_specification_registry <- analysis_target_csv("diag_ext_alternative_distance_first_stages", "iv_specification_registry.csv")
iv_overidentification <- analysis_target_csv("diag_ext_alternative_distance_first_stages", "iv_overidentification.csv")
iv_monotonicity <- analysis_target_csv("diag_ext_alternative_distance_first_stages", "iv_monotonicity_summary.csv")
iv_joint_balance <- analysis_target_csv("diag_ext_census_2001_controls", "instrument_balance_joint.csv")
```

## District-level treatment and instrument check

The district-level dotplot uses `r analysis_format_number(nrow(iv_dotplot), 0)` active panel rows and the same all-child EMI exposure used in the public IV specification. The accompanying table keeps the preferred linguistic-distance scalar visible alongside the treatment.

```{r instrument-current-code-analog}
instrument_dotplot_rows <- iv_dotplot[, intersect(c("district_order", "district_code", "state", "district", "emi_exposure_all_children_0708", "ling_distance_nonzero_mean", "state_prefix"), names(iv_dotplot)), drop = FALSE]
exposure_scale_summary <- data.frame(
  current_scale = "0-100 percentage scale",
  min_exposure = min(iv_dotplot$emi_exposure_all_children_0708, na.rm = TRUE),
  median_exposure = stats::median(iv_dotplot$emi_exposure_all_children_0708, na.rm = TRUE),
  max_exposure = max(iv_dotplot$emi_exposure_all_children_0708, na.rm = TRUE)
)
analysis_table(exposure_scale_summary, "All-child EMI-exposure scale check")
analysis_table(instrument_dotplot_rows, "Current treatment and instrument by district", max_rows = 30)
```

```{r instrument-dotplot, results='asis'}
analysis_image("diag_ext_instrument_exploration", "emie_by_district_dotplot.png", "All-child EMI exposure by district")
```

```{r instrument-tables}
analysis_table(iv_match, "Current IV-panel match summary")
analysis_table(iv_state, "Current IV-panel state summary", max_rows = 30)
analysis_table(iv_rows, "Current keyed IV summary rows", max_rows = 30)
analysis_table(iv_notes, "Legacy instrument-exploration notes retained as target output")
analysis_table(first_stage_support, "Fixed common support for first-stage absorption diagnostics")
analysis_table(first_stage_absorption, "First-stage absorption ladder", max_rows = 30)
analysis_table(alternative_distance_support, "Fixed support for alternative linguistic-distance first stages")
analysis_table(
  alternative_distance_summary[alternative_distance_summary$adjustment_id %in% c("region_expanded", "state_expanded"), ],
  "Alternative linguistic-distance first stages: preferred adjustment sets",
  max_rows = 30
)
analysis_table(
  mapping_coverage_sensitivity[
    mapping_coverage_sensitivity$adjustment_id == "state_expanded" &
      mapping_coverage_sensitivity$construction_id %in% c("nonzero_mean", "distance_shares_all_unmapped", "distance_shares_mapped"),
  ],
  "Mapping-coverage sensitivity under state fixed effects and expanded controls",
  max_rows = 30
)
analysis_table(unmapped_languages, "Largest unmapped C-16 language-state cells", max_rows = 20)
analysis_table(distance4_languages, "Distance-four speaker composition by language and state", max_rows = 20)
analysis_table(distance4_leave_one_out, "Distance-four leave-one-language-out joint tests", max_rows = 15)
analysis_table(weak_iv_outcomes, "Weak-IV-aware exploratory outcome estimates", max_rows = 10)
analysis_table(iv_diagnostic_applicability, "IV diagnostic applicability and implementation status", max_rows = 30)
analysis_table(iv_joint_balance, "Joint holdout-covariate balance tests", max_rows = 30)
analysis_table(iv_specification_registry, "De-duplicated IV diagnostic specification registry", max_rows = 30)
analysis_table(
  iv_overidentification[iv_overidentification$status != "not_applicable", ],
  "Overidentifying-restrictions diagnostics for multi-instrument specifications",
  max_rows = 30
)
analysis_table(
  iv_monotonicity[iv_monotonicity$adjustment_id %in% c("region_main", "state_main", "state_expanded"), ],
  "Residualized scalar first-stage shape diagnostics",
  max_rows = 30
)
analysis_table(
  alternative_distance_coefficients[
    grepl("distance_shares", alternative_distance_coefficients$specification_id) &
      grepl("state_expanded", alternative_distance_coefficients$specification_id),
  ],
  "State-expanded distance-share coefficients",
  max_rows = 20
)
analysis_table(first_stage_vif, "Main and expanded-control VIF/GVIF diagnostics", max_rows = 40)
analysis_table(first_stage_state_ranges, "State-by-state residual ranges", max_rows = 40)
analysis_table(first_stage_state_deletion[order(abs(first_stage_state_deletion$estimate_change), decreasing = TRUE), ], "Leave-one-state-out influence", max_rows = 30)
analysis_table(first_stage_district_influence[order(first_stage_district_influence$cooks_distance, decreasing = TRUE), ], "Most influential districts in the expanded first stage", max_rows = 30)
```
