---
title: "Instrument Exploration"
format: gfm
execute:
  echo: true
  output: true
  warning: true
  message: true
---

```{r setup}
helper <- if (file.exists("analysis/_analysis_helpers.R")) "analysis/_analysis_helpers.R" else "../_analysis_helpers.R"
source(helper)
```

## Current instrument and treatment diagnostics

The active Phase 1 construction uses the full mutually exclusive Census 2001 C-16 mother-tongue distribution. Language-group subtotal rows are removed before aggregation, linguistic distances are attached from the documented Shastry concordance, and unmapped speaker mass remains visible. The preferred exploratory scalar is `ling_distance_nonzero_mean`, the speaker-weighted mean distance among mapped languages with distance greater than zero. Hindi and Urdu are treated as distance zero and their shares are reported separately.

The preferred exploratory treatment is `emi_exposure_all_children_0708`, the survey-weighted share of all children ages 5-19 who are both enrolled and studying in English medium. The historical `EMIE` field remains a compatibility measure of EMI among enrolled children and is shown below only in the legacy dotplot section.

The Phase 2 and Phase 3 tables diagnose how the preferred exploratory first stage changes with six-region fixed effects, state fixed effects, main and expanded Census control sets, sequential thematic blocks, VIF/GVIF, state deletion, and district influence. These diagnostics do not change the public IV specification.

```{r instrument-prose-deviation-note, results='asis'}
analysis_deviation_note("The active note now describes the full-distribution linguistic-distance construction and all-child EMI exposure used by the Phase 2-3 diagnostics. The historical top-three/EMIE dotplot is retained below as an explicitly labeled compatibility check rather than as the current instrument definition.")
```

```{r instrument-targets}
iv_match <- read_analysis_csv("diagnostics", "public", "iv_panel_match_summary.csv")
iv_state <- read_analysis_csv("diagnostics", "public", "iv_panel_state_summary.csv")
iv_rows <- read_analysis_csv("diagnostics", "public", "iv_summary_keyed_rows.csv")
iv_dotplot <- analysis_target_csv("diag_ext_instrument_exploration", "instrument_strength_dotplot_data.csv")
iv_notes <- analysis_target_csv("diag_ext_instrument_exploration", "instrument_exploration_legacy_notes.csv")
first_stage_absorption <- analysis_target_csv("diag_ext_first_stage_absorption", "first_stage_absorption_ladder.csv")
first_stage_support <- analysis_target_csv("diag_ext_first_stage_absorption", "first_stage_absorption_common_support.csv")
first_stage_state_ranges <- analysis_target_csv("diag_ext_first_stage_absorption", "first_stage_state_residual_ranges.csv")
first_stage_state_deletion <- analysis_target_csv("diag_ext_first_stage_absorption", "first_stage_state_deletion.csv")
first_stage_district_influence <- analysis_target_csv("diag_ext_first_stage_absorption", "first_stage_district_influence.csv")
first_stage_vif <- analysis_target_csv("diag_ext_first_stage_absorption", "first_stage_vif.csv")
```

## Legacy top-three and EMI-among-enrolled compatibility check

The current analog of the historical `ggplot(..., aes(x = district_code_0708, y = EMIE, color = district_prefix)) + geom_point()` check uses `r analysis_format_number(nrow(iv_dotplot), 0)` active district-panel rows. It is retained to make earlier drafts reproducible, not as the preferred Phase 1 definition.

```{r instrument-current-code-analog}
instrument_dotplot_rows <- iv_dotplot[, intersect(c("district_order", "district_code", "state", "district", "EMIE", "wavg_ling_degrees", "state_prefix"), names(iv_dotplot)), drop = FALSE]
emie_scale_summary <- data.frame(
  current_scale = "0-100 percentage scale",
  min_EMIE = min(iv_dotplot$EMIE, na.rm = TRUE),
  median_EMIE = stats::median(iv_dotplot$EMIE, na.rm = TRUE),
  max_EMIE = max(iv_dotplot$EMIE, na.rm = TRUE),
  legacy_scale_note = "legacy prose values 1 and 0.4-0.1 correspond to 100 and 40-10 on the current scale"
)
analysis_table(emie_scale_summary, "Current EMIE scale check")
analysis_table(instrument_dotplot_rows, "Current EMIE-by-district dotplot data", max_rows = 30)
```

```{r instrument-dotplot, results='asis'}
analysis_image("diag_ext_instrument_exploration", "emie_by_district_dotplot.png", "Current EMIE-by-district exploratory dotplot")
```

```{r instrument-tables}
analysis_table(iv_match, "Current IV-panel match summary")
analysis_table(iv_state, "Current IV-panel state summary", max_rows = 30)
analysis_table(iv_rows, "Current keyed IV summary rows", max_rows = 30)
analysis_table(iv_notes, "Legacy instrument-exploration notes retained as target output")
analysis_table(first_stage_support, "Fixed common support for first-stage absorption diagnostics")
analysis_table(first_stage_absorption, "First-stage absorption ladder", max_rows = 30)
analysis_table(first_stage_vif, "Main and expanded-control VIF/GVIF diagnostics", max_rows = 40)
analysis_table(first_stage_state_ranges, "State-by-state residual ranges", max_rows = 40)
analysis_table(first_stage_state_deletion[order(abs(first_stage_state_deletion$estimate_change), decreasing = TRUE), ], "Leave-one-state-out influence", max_rows = 30)
analysis_table(first_stage_district_influence[order(first_stage_district_influence$cooks_distance, decreasing = TRUE), ], "Most influential districts in the expanded first stage", max_rows = 30)
```
