---
title: "Map and Palette Tuning"
format: gfm
execute:
  echo: true
  output: true
  warning: true
  message: true
---

```{r setup}
helper <- if (file.exists("analysis/_analysis_helpers.R")) "analysis/_analysis_helpers.R" else "../_analysis_helpers.R"
source(helper)
```

## Legacy prose retained with current results

For the names of all color palette: `cols4all::c4a_palettes()`. To explore the palettes: `cols4all::c4a_gui()`. To plot a palette: `cols4all::c4a("zissou1continuous", 7) %>% cols4all::c4a_plot()`. To view the color codes of a palette: `cols4all::c4a("rd_yl_bu", 5)`. To explore some of the palettes supported by `tmap`: `tmaptools::palette_explorer()`.

Based on geometry from @bhatiaMergingUpdatedDistrictlevel2020, and code from @nowosadElegantInformativeMaps2021. Recall that this change does not account for inflation! To help determine where breaks should be: `tapply((joined_df %>% drop_na(EMIE))$EMIE, (kmeans((joined_df %>% drop_na(EMIE))$EMIE, centers = 6))$cluster, range)`.

To print all of the maps: `for(nm in ls(pattern = "^map_")) {print(get(nm))}`. Save the maps. Uncomment when/if `final_draft == TRUE`. Save each `tmap` object as a PNG file with print-quality `dpi = 300`.

```{r map-prose-deviation-note, results='asis'}
analysis_deviation_note("The current note keeps the legacy palette/break/export prose and renders the active figure_files outputs rather than rerunning GUI palette tools or writing duplicate map-generation code in analysis/.")
```

```{r map-targets}
figure_outputs <- tryCatch(analysis_read_target("figure_files"), error = function(e) character())
figure_manifest <- if (file.exists(analysis_path("outputs", "figures", "main", "figure_manifest.csv"))) {
  read_analysis_csv("figures", "main", "figure_manifest.csv")
} else {
  data.frame(path = figure_outputs, stringsAsFactors = FALSE)
}
figure_manifest
```

The current figure target produced `r analysis_format_number(length(figure_outputs), 0)` files available to this render.  The map-collage previews below are target outputs, not copied legacy exploratory graphics.

```{r map-current-code-analog}
figure_manifest[grepl("collage|map", figure_manifest$path, ignore.case = TRUE), , drop = FALSE]
```

```{r map-previews, results='asis'}
analysis_image("figure_files", "collage_main_maps.png", "Current main-map collage")
analysis_image("figure_files", "collage_iv_region_maps.png", "Current IV and region map collage")
```

```{r map-output-table}
analysis_table(figure_manifest, "Current figure manifest", max_rows = 30)
```
