---
title: "Map and Palette Tuning"
format: gfm
execute:
  echo: true
  output: true
  warning: true
  message: true
---

```{r setup}
helper <- if (file.exists("analysis/_analysis_helpers.R")) "analysis/_analysis_helpers.R" else "../_analysis_helpers.R"
source(helper)
```

## Legacy diagnostic intent

The legacy map-making chunk used commented exploratory code for palette selection, break checks, `cols4all`/GUI inspection, `tmap_save()` sizing experiments, and k-means/break exploration for the EMI and linguistic-distance maps.  The current production maps are generated by the `figure_files` target; this note keeps the tuning rationale in `analysis/` and renders the current figure artifacts instead of copying static screenshots or comments.

```{r map-targets}
figure_outputs <- tryCatch(analysis_read_target("figure_files"), error = function(e) character())
figure_manifest <- if (file.exists(analysis_path("outputs", "figures", "main", "figure_manifest.csv"))) {
  read_analysis_csv("figures", "main", "figure_manifest.csv")
} else {
  data.frame(path = figure_outputs, stringsAsFactors = FALSE)
}
figure_manifest
```

The current figure target produced `r analysis_format_number(length(figure_outputs), 0)` files available to this render.  The map-collage previews below are target outputs, not copied legacy exploratory graphics.

```{r map-current-code-analog}
figure_manifest[grepl("collage|map", figure_manifest$path, ignore.case = TRUE), , drop = FALSE]
```

```{r map-previews, results='asis'}
analysis_image("figure_files", "collage_main_maps.png", "Current main-map collage")
analysis_image("figure_files", "collage_iv_region_maps.png", "Current IV and region map collage")
```

```{r map-output-table}
analysis_table(figure_manifest, "Current figure manifest", max_rows = 30)
```
