---
title: "Technical Note for Replication: 8.3 Filenames"
author: "Rishav Roy"
format:
  html: default
  pdf:
    pdf-engine: xelatex
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| results: asis
legacy_path <- "../archive/legacy-paper-drafts/580-Draft-ECON-580.Rmd"
legacy <- readLines(legacy_path, warn = FALSE)
legacy <- paste(legacy, collapse = "\n")
strip_code_chunks <- function(x) {
  lines <- strsplit(x, "\n", fixed = TRUE)[[1]]
  out <- character()
  in_chunk <- FALSE
  for (line in lines) {
    if (startsWith(line, "```")) {
      in_chunk <- !in_chunk
      next
    }
    if (!in_chunk) out <- c(out, line)
  }
  paste(out, collapse = "\n")
}
extract_between <- function(x, start, end = NULL) {
  start_pos <- regexpr(start, x, perl = TRUE)
  if (start_pos < 0) stop("Could not find: ", start, call. = FALSE)
  if (is.null(end)) return(substr(x, start_pos, nchar(x)))
  rest <- substr(x, start_pos + attr(start_pos, "match.length"), nchar(x))
  end_pos <- regexpr(end, rest, perl = TRUE)
  if (end_pos < 0) return(substr(x, start_pos, nchar(x)))
  substr(x, start_pos, start_pos + attr(start_pos, "match.length") + end_pos - 2)
}
body <- strip_code_chunks(legacy)
body <- sub("^---.*?---", "", body, perl = TRUE)
cat(extract_between(body, "^# Technical Note for Replication", "^# References"))
```

```{r}
#| echo: true
#| eval: false
# ---TROUBLESHOOTING---
# If getting "file does not exist" or similar errors:
# Switch the function used to read in the file e.g., if the relevant file is being read in using read_sav() or read_sav_short(), switch to read_sav_short() or read_sav() respectively.

# If that didn't work and you're not using Windows: 
# Either modify this code chunk for your OS (note the "if" statement at the start of each function) or manually shorten the error-inducing file paths.

# ---

# Purpose of this chunk: 
# To ensure R can identify and read in all necessary files.
# This project found me making frequent use of file paths over 260 characters in length. Out of my naivete, I didn't realize Windows File Explorer automatically shortens such file paths by rewriting them in the 8.3 filename convention---meaning R can no longer access their standard "long name." Conversely, functions like read_sav() automatically lengthen 8.3 filenames via functions like normalizePath(), turning the true 8.3 file path into a long name which, from R's and (thus our) perspective, is useless.
# This code chunk turns our long name inputs into 8.3 filenames while tricking the read_*()-type functions into not turning them back.


# Context for nerds:
# In my Registry Editor, under Computer\HKEY_LOCAL_MACHINE\SYSTEM\CurrentControlSet\Control\FileSystem, I had set the LongPathsEnabled value to 1 (which enables Windows to use long file paths) and the NtfsDisable8dot3NameCreation value to 2. I had also changed Group Policy to "Enable Win32 long paths".
# But none of that mattered! Windows File Explorer lacks a longPathAware entry in its .exe file's manifest--so despite LongPathsEnabled=1, my file paths were still capped at 260 characters. Ergo this code chunk.
# It's possible that setting NtfsDisable8dot3NameCreation=0 would've prevented all of this code from being necessary. But I'd rather not force replicators to change their registry settings.
```
