---
title: "Technical Note for Replication: 8.3 Filenames"
author: "Rishav Roy"
format:
  html: default
  pdf:
    pdf-engine: xelatex
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| results: asis
legacy_path <- "../archive/legacy-paper-drafts/580-Draft-ECON-580.Rmd"
legacy_lines <- readLines(legacy_path, warn = FALSE)

strip_yaml <- function(lines) {
  if (length(lines) >= 2L && identical(lines[[1]], "---")) {
    close <- which(lines[-1L] == "---")
    if (length(close)) return(lines[-seq_len(close[[1]] + 1L)])
  }
  lines
}

strip_code_chunks <- function(lines) {
  out <- character()
  in_chunk <- FALSE
  for (line in lines) {
    if (startsWith(line, "```")) {
      in_chunk <- !in_chunk
      next
    }
    if (!in_chunk) out <- c(out, line)
  }
  out
}

find_line <- function(lines, pattern) {
  hit <- grep(pattern, lines, fixed = TRUE)
  if (!length(hit)) stop("Could not find: ", pattern, call. = FALSE)
  hit[[1]]
}

extract_lines <- function(lines, start, end = NULL) {
  i <- find_line(lines, start)
  j <- if (is.null(end)) length(lines) + 1L else find_line(lines, end)
  lines[i:(j - 1L)]
}

body <- strip_code_chunks(strip_yaml(legacy_lines))
out <- extract_lines(body, "# Technical Note for Replication", "# References")
out <- gsub("`r ", "` r ", out, fixed = TRUE)
cat(paste(out, collapse = "\n"))
```

```{r}
#| echo: true
#| eval: false
# ---TROUBLESHOOTING---
# If getting "file does not exist" or similar errors:
# Switch the function used to read in the file e.g., if the relevant file is being read in using read_sav() or read_sav_short(), switch to read_sav_short() or read_sav() respectively.

# If that didn't work and you're not using Windows: 
# Either modify this code chunk for your OS (note the "if" statement at the start of each function) or manually shorten the error-inducing file paths.

# ---

# Purpose of this chunk: 
# To ensure R can identify and read in all necessary files.
# This project found me making frequent use of file paths over 260 characters in length. Out of my naivete, I didn't realize Windows File Explorer automatically shortens such file paths by rewriting them in the 8.3 filename convention---meaning R can no longer access their standard "long name." Conversely, functions like read_sav() automatically lengthen 8.3 filenames via functions like normalizePath(), turning the true 8.3 file path into a long name which, from R's and (thus our) perspective, is useless.
# This code chunk turns our long name inputs into 8.3 filenames while tricking the read_*()-type functions into not turning them back.


# Context for nerds:
# In my Registry Editor, under Computer\HKEY_LOCAL_MACHINE\SYSTEM\CurrentControlSet\Control\FileSystem, I had set the LongPathsEnabled value to 1 (which enables Windows to use long file paths) and the NtfsDisable8dot3NameCreation value to 2. I had also changed Group Policy to "Enable Win32 long paths".
# But none of that mattered! Windows File Explorer lacks a longPathAware entry in its .exe file's manifest--so despite LongPathsEnabled=1, my file paths were still capped at 260 characters. Ergo this code chunk.
# It's possible that setting NtfsDisable8dot3NameCreation=0 would've prevented all of this code from being necessary. But I'd rather not force replicators to change their registry settings.
```
