---
title: "Economic Returns to English-Medium Instruction in India"
author: "Rishav Roy"
date: "2024-04-22"
header-includes: \usepackage{setspace}\doublespacing
output:
  pdf_document: default
  html_notebook: default
  bookdown::pdf_document2: default
  tufte::tufte_handout: default
subtitle: "A District-Level Analysis of Human Capital Investments and Economic Growth"
abstract: Despite the country's economic growth, many Indias
bibliography: English Education Economic Returns.bib
citation_package: biblatex
link-citations: yes
csl: https://www.zotero.org/styles/apa
editor_options:
  canonical: yes
  chunk_output_type: inline
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
	fig.align = "center",
	fig.height = 5,
	fig.width = 8,
	message = FALSE,
	warning = FALSE,
	cache = TRUE,
	include = FALSE
)
library(tidyverse)
library(bookdown)
library(here)
library(haven)
library(janitor)
library(equatiomatic)
#Despite these warnings, my code wouldn't work unless I explicitly imported in this library. 
#The coding gods disagreed with RStudio.

```





```{r import data, include=FALSE}

nss07block1.2 <- read_sav(
  here("NSS 2007-08 Participation and Expenditure in Education 64th Round","Block-1 & 2 Identification of sample household.sav")
  ) %>% 
  mutate(SAMPLE = SAMPLE %>% as.numeric()) %>% 
  select(-LEVEL, -CENTRE_ROUND_SHIFT)

nss07block3 <- read_sav(
  here("NSS 2007-08 Participation and Expenditure in Education 64th Round","Block-3  Household  characteristics.sav")
  ) %>% 
  mutate(SAMPLE = SAMPLE %>% as.numeric()) %>% 
  select(-LEVEL, -CENTRE_ROUND_SHIFT)

nss07block4 <- read_sav(
  here("NSS 2007-08 Participation and Expenditure in Education 64th Round","Block-4  Demographic and other particulars of household members.sav")
  ) %>% 
  mutate(SAMPLE = SAMPLE %>% as.character()) %>% 
  mutate(SAMPLE = SAMPLE %>% as.numeric()) %>% 
  select(-LEVEL, -CENTRE_ROUND_SHIFT)

nss07block5 <- read_dta(
  here("NSS 2007-08 Participation and Expenditure in Education 64th Round","Block 5.dta")
  ) %>% 
  mutate(SAMPLE = SAMPLE %>% as.numeric()) %>% 
  select(-LEVEL, -CENTRE_ROUND_SHIFT)

nss07block7 <- read_sav(
  here("NSS 2007-08 Participation and Expenditure in Education 64th Round","Block-7  Particulars of currently not attending persons aged 5-29 years.sav")
  ) %>% 
  mutate(SAMPLE = SAMPLE %>% as.numeric()) %>% 
  select(-LEVEL, -CENTRE_ROUND_SHIFT)

HHnss0708 <- nss07block1.2 %>% 
  full_join(nss07block3)

# get_dupes(nss0708,HHID) #Used to check for rows which are duplicated 

Pnss0708 <- HHnss0708 %>% 
  full_join(nss07block4) %>% 
  full_join(nss07block5) %>% 
  full_join(nss07block7)

# get_dupes(nss0708,PID)

Pnss0708 <- clean_names(Pnss0708)
HHnss0708 <- clean_names(HHnss0708)

# remove(nss07block1.2, nss07block3, nss07block4, nss07block5, nss07block7)

```

```{r}

Pnss0708 <- Pnss0708 %>% 
  unite(state_district, c(state, district),sep="")

datasetf <- Pnss0708 %>% 
  drop_na(lvl_curr_attendence) %>% 
  group_by(state_district) %>%
  summarize(
    n_students=n(),
    EMI_I = sum(medium_instruction=="02")/n_students,
    avg_consumption_0708 =
          mean(total,na.rm=T)
    )

```

```{r}

plfs1819 <- read_dta(here("PLFS 2018-19 Periodic Labour Force Survey July 2018 - June 2019","HHRV-2018-19.dta"))

plfs1819 <- plfs1819 %>% 
  rename(total_consumption_1819 = b3q5_hh_fv) %>%
  filter(visit_hh_fv == "V2") %>% 
  unite(state_district, c(B0q1state_hh_fv, B1q4dist_code_hh_fv),sep="")

datasetf <- datasetf %>% 
  full_join(
    plfs1819 %>% 
      mutate(total_consumption_1819 =
               as.numeric(total_consumption_1819)) %>% 
      group_by(state_district) %>% 
      summarize(
        n_homes_final=n(),
        avg_consumption_1819 =
          mean(total_consumption_1819,na.rm=T))
  ) %>% 
  mutate(consumption_growth = 
           avg_consumption_1819 -
           avg_consumption_0708
         )


summary(lm(log(consumption_growth)~EMI_I, data=datasetf))


```


Try executing this chunk by clicking the *Run* button within the chunk or by placing your cursor inside it and pressing *Ctrl+Shift+Enter*. 

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

\begin{equation} 
  f\left(k\right) = \binom{n}{k} p^k\left(1-p\right)^{n-k}
  (\#eq:binom)
\end{equation}

See \@ref(eq:binom)

\newpage
# References
