---
title: "Inequality in a Potential Equalizer of Opportunity"
subtitle: "Variation in the Accessibility and Net Benefits of English-Medium Instruction in India"
author: "Rishav Roy"
thanks: "Acknowledgments: I would like to thank Laura Schechter, Corina Mommaerts, and Agustín Gutiérrez, for their feedback on earlier stages of this project. The Internet Archive also proved indispensable for this project; a plurality of the data sources I use were only accessible because of their preservation work. GPT models 4.5, 5, 5.4, 5.5, and 5.6 were used for troubleshooting, proofreading, and generating parts of this paper and its code from detailed outlines and pseudocode. All ideas, analytical work, and errors are my own."
format:
  pdf:
    pdf-engine: xelatex
bibliography: references.bib
execute:
  echo: false
  warning: false
  message: false
abstract: "English-medium instruction (EMI), or the teaching of school subjects in English, is often viewed as a potential palliative for the integenerational rigidty of socioeconomic status in India, where English skills command substantial wage premia. I study whether the conditions governing households’ access to and potential net gains from EMI are themselves unequally distributed. I organize my analysis around three such conditions: inherited linguistic costs of acquiring English, access to English-medium schooling, and the local economic setting in which English skills may be rewarded. By harmonizing district data on languages, schooling, consumption, migration, human capital, and sectoral employment, I find inequality across all three. Inherited linguistic conditions are highly persistent, but their effects on language acquisition vary across insitutional settings. EMI exposure and provision are strongly organized by schooling system and sharply stratified along socioeconomic lines, even within the same local education market. Consumption growth is more associated with EMI enrollment than enrollment alone, and their strength increases in pre-existing human capital and urbanization. Though linguistic conditions predict more human capital, financial access, and structural change, they do not provide persuasive within-state variation in EMI and I interpret all associations descriptively. In sum, a high-return educational opportunity need not equalize opportunity when access to it and the resources required to benefit from it are themselves unequal."
geometry: "left=0.85in, right=0.85in, top=0.9in, bottom=0.9in"
header-includes:
  - \usepackage{setspace}\doublespacing
  - \usepackage{mathtools}
  - \usepackage{longtable}
  - \usepackage{booktabs}
  - \usepackage{array}
  - \usepackage{xcolor}
  - \definecolor{gray35}{gray}{0.35}
  - \usepackage{colortbl}
  - \usepackage{threeparttablex}
  - \usepackage{makecell}
  - \usepackage{tabularray}
  - \usepackage{float}
  - \usepackage{threeparttable}
  - \usepackage{etoolbox}
  - \AtBeginEnvironment{CSLReferences}{\setstretch{1}\setlength{\parskip}{0pt}\setlength{\itemsep}{0pt}}
  - \UseTblrLibrary{booktabs}
  - \UseTblrLibrary{siunitx}
  - \usepackage{caption}
  - \usepackage{subcaption}
  - \captionsetup{width=.95\linewidth,justification=centering}
  - \captionsetup[subtable]{labelformat=empty}
link-citations: true
cite-method: citeproc
---

<!--
WORKING PAPER: paper-new.qmd is the manuscript-facing draft built around the
registered final exhibits. paper.qmd remains the legacy compatibility document
until application-sample excerpts are migrated separately.
-->

```{r public-output-table-helper}
#| include: false
source_public_qmd_helpers <- function() {
  candidates <- c(
    "R/output/public_qmd_helpers.R",
    "../R/output/public_qmd_helpers.R",
    "../../R/output/public_qmd_helpers.R"
  )
  hit <- candidates[file.exists(candidates)]
  if (!length(hit)) stop("Could not locate R/output/public_qmd_helpers.R", call. = FALSE)
  source(hit[[1]], local = knitr::knit_global())
}
source_public_qmd_helpers()
initialize_public_qmd_helpers(knitr::knit_global())
```

# Introduction {#sec-intro}

In an Indian economy characterized by rapid GDP growth alongside wage stagnation, profuse informality, and high rates of both youth unemployment and underemployment [@organization2024a], what can a household do to improve the future prospects of their children? The history of economics highlights *education* as one answer [@mincerHumanCapitalEarnings1974; @card1999; @bjorklundEstimatingReturnInvestments2002; @bhandariIncomeDifferentialsReturns2006]. The history of India highlights another: *English*. In a context of extraordinary linguistic diversity [@gurevichDatasetLinguisticConnectivity2025] and thus extraordinary communication and coordination costs [@nakagawa2021a], English has come to serve as a primary domestic lingua franca [@clingingsmithIndustrializationBilingualismIndia2014]. In its capacity as a global lingua franca [@melitzEnglishGlobalLanguage2016], meanwhile, English opens pathways both internationally---e.g., to education at prestigious universities or employment at multinational firms [@ramanathanTeacherKnowledgeBase2014; @bediEnglishLanguageIndia2019, pp. 32-33]---and at home. The domestic expansion of internationally traded services and information technology, in particular, has led to a flurry of jobs whose higher socioeconomic returns [@chamarbagwalaIndustrialDelicensingTrade2011; @smokotin2014a] are largely accessible only to those who can use English [@mankiwPoliticsEconomicsOffshore2006; @shastry2012a].

Empirical study corroborates the notion that English expands economic mobility [@itaniMeaningLanguageSkills2015]. Using nationally representative household data, @azam2013a find that English fluency was, at least up until 2005, associated with a 34% hourly wage premium for men and a 22% premium for women; even limited English proficiency was associated with premia of 13% and 10%, respectively. @refeque2022a find similar differentials extending through 2011--12. Evidence for causality in this association comes from @chakraborty2016a, who interpret West Bengal's 1983--2004 ban on English teaching in public primary schools as a natural experiment where cohorts more *exposed* to the policy earned lower subsequent wages. "Language skills are human capital," after all [@chiswick1995a, p. 4], just as "English-language skills [specifically] are a form of human capital" [@azam2013a, p. 344].

It is thus by combining our two answers---education and English---that we can begin to understand the demand for *English-medium instruction (EMI)*, or for the teaching of all school subjects in English, currently seen in India [@gooptu2023a; @lahoti2019a; @chattopadhayLowFeePrivateSchools2017]. For its potential to expand equality of opportunity, EMI has been touted as a tool of social development [@commission2009a] and described as a route to escape caste [@mathur2013a; @mathew2011a; @vij2020a] alongside the rigid wage, employment, and occupation distributions it enforces across generations [@madheswaranSocialExclusionCaste2016; @munshi2006a].

But it is ultimately households who decide whether or not a child attends EMI, and the determinants of their choice---the costs they face, the benefits they expect, and the forces constraining the choice set in front of them---are affected by factors which are not just national and historical, but also specific and local. My focus is on whether such factors are associated with the same inequalities that English skills are often expected to help households escape. I study inequalities which exist along dimensions such as social group, religion, parental education, household resources, and urban or rural residence. If the ability to access or benefit from EMI is systematically related to these characteristics, even within the same local education market, then we have evidence that a high economic payoff to English skills alone is not enough for EMI to become an equalizer of opportunity.

Though I never estimate a structural model of household choice, their investment problem still provides a helpful organizing framework. I will first study a type of cost whose variation is predetermined relative to the institutional and economic factors that otherwise affect the household's problem: **inherited linguistic costs to learning English**. These will reflect historically persistent features of the local linguistic environment [see @roweLanguageDevelopmentContext2020] which affect the opportunity cost of learning English over other domestic bridge languages, namely Hindi. This work, detailed in @sec-language, owes much credit to @shastry2012a; in line with her methodology, I create several different measures of local linguistic distance from *Hindi* to capture this opportunity cost. In @sec-access I analyze the **accessibility of English-medium instruction** through enrollment, several measures of EMI exposure, and school management. Private schooling is a distinct household choice with its own costs and selection patterns, so I do not treat private management as a measure of instructional quality [@endow2019a; @muralidharan2015a]. Finally, in @sec-conversion I study potential determinants of **expected returns to English skills**. My goal here is to find evidence on the value of English skills in the local labor market, with the assumption that households incorporate the local history of such returns into their beliefs regarding English, akin to households having adaptive expectations. I ask whether EMI exposure predicts later district consumption and whether that association is stronger where pre-existing human capital and urbanization are greater. I then examine whether inherited linguistic conditions accompany human-capital growth and skilled in-migration, as well as changes in financial access and economic structure.

The local linguistic environment, the local education market, and the local labor market---all three of the factors I focus on are studied locally. These choices are justifiable: the linguistic conditions which households pass down to their children are highly persistent at the local level, local schooling conditions constrain the menu of educational opportunities households choose from, and local economic conditions plausibly cause households to update their beliefs on the economic returns to learning English. I will use a household's district to proxy for their 'local' environment.^[A district in India is a political unit akin to a county in the United States, albeit with an average population of around 2 million. Table \ref{tbl-paper-core-summary} contains summary statistics derived from the distribution of district populations.] Though the district is the most granular unit at which I can harmonize the longitudinal data sources I rely on, it must be noted that interdistrict migration would confound and attenuate any association between the characteristics of one district's education market and its future economic outcomes. Older studies found interdistrict migration rates (and particularly rates of economically-motivated interdistrict migration) to be so low even after the expansion of IT and services in the 1990s [@topalovaTradeLiberalizationPoverty2005; @office2021a] that they "assume[d] that district of current residence (or of employment) of an individual is approximately the same as the schooling district" [@chakraborty2016a, p. 6]. More recent work, however, complicates [@mahapatroInternalMigrationEmerging2020; @rajanCOVID19PandemicInternal2020] and even outright repudiates [@lucasCountrySpecificMagnitudesMigration2021] this notion. These studies emphasize the role of the IT expansion in driving interdistrict migration [@divison2015a; @ghoseTradeInternalMigration2026], and my own results associate potential determinants of household investment into EMI with the skill composition of recent work migrants. I return to those results in @sec-app-migration. Regardless, it is evident that district outcomes are not an aggregation of individual returns to EMI, but a function of places and their changing populations.

The results show that the conditions surrounding English-skill investment are themselves unequally distributed. The relative cost of acquiring a bridge language is partly inherited, and language behavior varies with these inherited linguistic conditions. Formal English-medium schooling is strongly organized by state boundaries, particularly in the public sector. Within the same district, social-group differences remain large in English-medium and private-school participation, with larger differences in urban schooling markets. EMI exposure is more strongly associated with later local consumption than enrollment alone, and this association is stronger in districts with higher initial human capital and urbanization. The inherited linguistic gradient is also associated with later changes in human capital, the skill composition of recent work migrants, financial access, and sectoral composition. These broader associations are selective rather than evidence of general employment expansion.

Taken together, the evidence limits the equalizing promise of a high-return educational opportunity. A child may face high future payoffs to English skills while the household inherits a less favorable linguistic starting point, has less access to EMI, or lives in a district with fewer complementary resources through which English skills can be converted into economic opportunity. English skills can be valuable while access to English-medium schooling remains unequal. A high-return educational opportunity does not automatically equalize opportunity when entry into it, and the resources needed to benefit from it, are themselves unequally distributed.

# Data and Variables {#sec-data}

I construct the district datasets used in the paper by reconciling changing district boundaries, reconstructing archived schooling records, linking Census mother tongues to several linguistic classifications, and harmonizing household consumption surveys across major changes in survey design. I place survey, Census, and school records from different boundary regimes onto Census-2001 district geography using a reviewed district-lineage system informed by official identifiers, district histories, and published concordances [@kumarCreatingLongPanels2016; @ministryPanchayatiRajLocalGovernmentDirectory2026; @asherDevelopmentResearch2021]. I reconstruct district-level EMI and school-system measures from archived DISE workbooks and report cards whose formats and medium-of-instruction fields vary across releases [@mehtaUDISEPublications2005062022; @diseDistrictReportCardsArchive2017]. I combine Census mother-tongue populations with Shastry's Hindi-distance classification, reviewed Census-language crosswalks, Glottolog genealogy, and Dyen lexical data [@shastry2012a; @officeRegistrarLanguageAtlasIndia2004; @hammarstromGlottologGlottologGlottolog2026; @dyenIndoeuropeanClassification1992; @dyenComparativeIndoeuropeanDatabase1997]. I construct comparable district consumption measures from nine NSS/HCES survey designs spanning 2000--01 through 2023--24, placing households on Census-2001 district geography and expressing consumption in 2011--12 prices. Five survey designs provide the baseline and later outcomes used in the main longitudinal analysis [@nssHouseholdConsumerExpenditure2004; @nssHouseholdConsumerExpenditure2009Type2; @nssHouseholdConsumerExpenditure2011Type2; @nsoHCES2022; @nsoHCES2023]. Construction details appear in @sec-app-data, including the additional Census and Economic Census panels used for local-development outcomes. Identification checks are reported in @sec-app-identification.

**Geography and coverage.** District boundaries change often enough that a district name cannot serve as a stable longitudinal identifier. I use Census-2001 districts as the reference geography and reconcile later and earlier units using reviewed district-lineage evidence. The lineage system draws on Census materials and independent district-harmonization resources, including the India State and District Evolution Database, the Indian Census Data Collection, the India District Changes Tracker, and SHRUG [@joladIndiaStateDistrict2026; @joladIndianCensusData2026; @jaacksIndiaDistrictChanges2020; @asherDevelopmentResearch2021]. The reference geography contains 593 Census-2001 districts. The 2007--08 schooling survey reaches 587 of them under the reviewed lineage, while the primary 2017--18 lineage reaches 578; requiring support in both schooling rounds yields a 573-district panel. Because the household surveys are repeated cross-sections rather than repeated observations of the same respondents, I aggregate the relevant measures to harmonized districts.

The principal schooling measures come from the 2007--08 National Sample Survey (NSS) 64th Round, *Participation and Expenditure in Education* [@nationalsamplesurveyoffice2008]. I use survey weights when forming district aggregates. DISE provides a separate school-reported measure of medium of instruction, management, enrollment, and school characteristics. I reconstruct the DISE district series from report cards covering 2005--06 through 2007--08 that were preserved online after the original publication pages disappeared [@mehtaUDISEPublications2005062022; @diseDistrictReportCardsArchive2017]. Census-2001 language tables provide mother-tongue populations and language behavior [@officeoftheregistrargeneralandcensuscommissionerofindiaCensusIndia20012001]. Later sections also use Census-2001 and Census-2011 migration, household, housing, and worker tables. I use the 2005 and 2013--14 Economic Censuses for changes in local economic structure [@centralstatisticsofficeIndiaFifthEconomic2008; @office2019a]. Labor outcomes come from the 2009--10 NSS employment survey and the 2017--18 Periodic Labour Force Survey [@nationalsamplesurveyofficeIndiaEmploymentUnemployment2010; @nationalstatisticalofficeINDIAPeriodicLabour2019].

**Access to English-medium instruction.** The NSS schooling variables separate school participation from medium of instruction. Let $N_d$ be the survey-weighted number of children ages 5--19 in district $d$, $E_d$ the number enrolled, $K_d$ the number enrolled for whom medium of instruction is observed, and $M_d$ the number in $K_d$ enrolled in English-medium instruction (EMI). I call

$$
\mathrm{Enrollment}_d = 100\frac{E_d}{N_d}
$$ {#eq-enrollment}

the *enrollment rate*. This is the extensive margin: whether a child is enrolled at all. Among enrolled children whose medium of instruction is observed, I call

$$
\mathrm{EnglishMediumShare}_d = 100\frac{M_d}{K_d}
$$ {#eq-emi-enrolled}

the *English-medium share among enrolled children*. This is the conditional medium-of-instruction margin among children with a reported medium. Finally,

$$
\mathrm{EnglishMediumExposure}_d = 100\frac{M_d}{N_d}
$$ {#eq-emie}

is *English-medium exposure among all children*. It incorporates enrollment, whether medium is reported among enrolled children, and EMI conditional on a reported medium. Enrolled children with unreported medium remain in an explicit unclassified category. The all-child exposure measure is therefore lower than the conditional English-medium share whenever some children are unenrolled or some enrolled children have unreported medium. I also distinguish *public-school English-medium exposure* and *private-school English-medium exposure*, each defined as the share of all children observed in English-medium schooling in the relevant management sector. Children whose medium or school management is unreported remain outside those sector-medium cells. *Private-school enrollment* is the share of enrolled children with observed school management who attend private aided or unaided schools. These measures describe observed formal-school enrollment and reported medium of instruction. NSS 64 excludes classes taken from private tutors and therefore does not measure supplemental English instruction [@nationalsamplesurveyoffice2008].

**Inherited linguistic conditions.** The primary language measure begins with Shastry's 0--5 ordinal scale of separation from Hindi, developed with linguist Jay Jasanoff [@shastry2012a]. Shastry uses linguistic distance from Hindi to capture variation in the relative cost of acquiring English rather than Hindi as a lingua franca. For each mapped, non-English mother tongue $l$ in district $d$, let $S_{ld}$ denote its Census-2001 speaker population and $D_l$ its Shastry distance from Hindi. The *speaker-weighted distance from Hindi* is

$$
\overline{D}_d =
\frac{\sum_l S_{ld}D_l\,\mathbf{1}\{D_l>0\}}
     {\sum_l S_{ld}\,\mathbf{1}\{D_l>0\}}.
$$ {#eq-speaker-weighted-distance}

The positive-distance restriction excludes Hindi and Urdu, the zero-distance reference languages, from both the numerator and denominator. The resulting mean describes the linguistic composition of speakers whose mapped mother tongue has positive distance from Hindi. I use *linguistic distance from Hindi* as a shorter name for this conditional speaker-weighted measure in the main text. A second measure, the *share speaking languages distant from Hindi*, is the percentage of all mother-tongue speakers in the district whose mapped language has Shastry distance at least three. The distant-language share uses all mother-tongue speakers in its denominator, so the size of the distant-language population enters the measure directly.

I use several alternatives when testing whether results depend on one linguistic classification. The *top-three-language distance from Hindi* is the speaker-weighted Shastry distance among the three largest mapped mother tongues in a district, corresponding to the older construction used in the original analysis. *Genealogical distance from Hindi* is the unweighted number of ancestry-tree edges separating each language from Hindi in Glottolog [@hammarstromGlottologGlottologGlottolog2026]. *Lexical noncognacy with Hindi* is one hundred minus the percentage of cognate items shared with Hindi in the Dyen lexicostatistical data [@dyenIndoeuropeanClassification1992; @dyenComparativeIndoeuropeanDatabase1997]. I also reconstruct the speaker-weighted and distant-language measures using the 1991 Language Atlas for historical validation [@officeRegistrarLanguageAtlasIndia2004]. The crosswalk and adjudication rules for Census mother tongues are documented in @sec-app-data.

**Language-behavior controls.** The Census language-acquisition regressions use state-by-native-language cells rather than districts, so their controls differ from the district covariates used elsewhere. Within each state I condition on the native language's share of speakers and an indicator for the state's most common native language. The relevant continuous and threshold distance specifications also include an indicator for Hindi/Urdu, the zero-distance reference category. I refer to the first two variables together as *within-state language-prevalence controls*. Their summary statistics appear separately in Table \ref{tbl-paper-core-summary} because their unit of observation is a state-language cell rather than a district.

**Economic outcomes and baseline covariates.** I use household consumption as the principal measure of local material living standards. At the household level, monthly per-capita consumption expenditure (MPCE) is consumption expenditure divided by household size. I express MPCE in 2011--12 prices and estimate district means using the survey design and person weights, so the district statistic is the person-level average of real MPCE implied by the survey weights. The baseline comes from the 2004--05 NSS 61st Round [@nssHouseholdConsumerExpenditure2004]. The 2009--10 and 2011--12 NSS rounds provide intermediate endpoints [@nssHouseholdConsumerExpenditure2009Type2; @nssHouseholdConsumerExpenditure2011Type2], and HCES 2022--23 and 2023--24 provide the long-run endpoints [@nsoHCES2022; @nsoHCES2023]. I reconstruct each survey's household consumption measure under its documented design and express the resulting MPCE series in 2011--12 prices using official price series, including the Reserve Bank of India and Labour Bureau series [@reserveBankIndiaDBIE; @labourBureauCPIIW]. HCES 2022--23 and 2023--24 divide consumption across three questionnaires administered to each selected household in three consecutive monthly visits [@nsoHCES2022; @nsoHCES2023]. I reconstruct household MPCE from those questionnaire components before district estimation; @sec-app-data gives the details.

The preferred district regressions compare districts within states and adjust for covariates measured in 2001, before the 2007--08 schooling outcomes. I call these the *baseline Census-2001 covariates*. The registered covariates cover six groups: scale and geography, social composition, human capital, demography, economic structure, and basic development. They include log population, urban share, and log population density; Scheduled Caste, Scheduled Tribe, and Muslim shares; secondary-or-higher attainment; the dependency ratio; the agricultural share of workers; and household electricity access [@officeoftheregistrargeneralandcensuscommissionerofindiaCensusIndia20012001; @asherDevelopmentResearch2021]. These covariates adjust the district regressions for observed differences in baseline conditions. In the heterogeneity analysis, baseline secondary-or-higher attainment and urban share enter as moderators and are omitted from the baseline covariate vector.

Table \ref{tbl-paper-core-summary} reports the scale and dispersion of the principal variables and both covariate families. Enrollment averages about 74 percent of children ages 5--19, while English-medium exposure among all children averages about 15 percent. Private-school English-medium exposure averages about 9 percent and public-school English-medium exposure about 6 percent. The speaker-weighted distance from Hindi averages 3.05 Shastry degrees across districts, and the district mean of the share speaking languages at distance three or greater is 42 percent. The table also reports all ten baseline district covariates and, in a separate state-language panel, the variables used in the Census language-behavior regressions. Mean real consumption in the available district samples is about Rs. 928 per month in 2004--05, Rs. 2,152 in 2022--23, and Rs. 2,242 in 2023--24, all expressed in 2011--12 prices.

::: {.landscape}

```{r}
#| label: render-main-core-summary
#| echo: false
#| results: asis

render_public_table("../outputs/tables/main/paper_core_summary.tex", "paper_core_summary")
```

:::

# Inherited Linguistic Conditions {#sec-language}

## Proxies and Persistence of Linguistic Conditions {#sec-language-persistence}

Why measure distance from Hindi? India's linguistic diversity gives languages of wider communication an important role, and both Hindi and English have served that role across groups [@clingingsmithIndustrializationBilingualismIndia2014; @shastry2012a]. Shastry's argument is comparative: linguistic diversity creates incentives to learn English or Hindi, and greater distance between a mother tongue and Hindi raises the cost of learning Hindi relative to English [@shastry2012a]. The speaker-weighted measure defined in @sec-data summarizes a persistent linguistic component of this relative cost.

Most district variation in the measure occurs between states. In the 573-district schooling panel used later in the paper, state indicators alone explain 84.4% of its cross-district variation. The preferred district specifications therefore compare districts within states and adjust for the baseline Census-2001 covariates. @fig-linguistic-geography shows the observed measure and the residual after removing state indicators and the linear contribution of those covariates.

::: {#fig-linguistic-geography layout-ncol=2}
![Raw linguistic distance.](../outputs/figures/main/map_linguistic_distance.pdf)

![Residual linguistic distance: state fixed effects + baseline Census-2001 covariates.](../outputs/figures/main/map_residual_linguistic_distance_state_main.pdf)

District linguistic distance before and after the preferred state-level adjustment. Sources: Census of India 2001 language and district characteristics [@officeoftheregistrargeneralandcensuscommissionerofindiaCensusIndia20012001; @asherDevelopmentResearch2021]. `r paper_map_boundary_note()` [@datameetCensus2001Districts; @naturalearthDisputedBoundaries].
:::

I call these conditions *inherited* because the district pattern is highly persistent across the independently reconstructed 1991 and 2001 measures. Among the 91 Census-2001 districts meeting the historical reconstruction requirements in both years, the correlation between the 1991 and 2001 speaker-weighted distance measures is 0.898; weighting districts by their 1991 populations raises the correlation to 0.928. The population-weighted correlation for the share of speakers whose mother tongues are at least three Shastry degrees from Hindi is 0.919. @fig-linguistic-persistence shows the district comparison. These estimates agree with Shastry's finding that district linguistic distance is highly correlated across earlier Census years [@shastry2012a].

![Persistence of speaker-weighted linguistic distance from Hindi from 1991 to 2001. Each point is a Census-2001 district meeting the historical language-reconstruction requirements in both years; the dashed line is the 45-degree line. Sources: Census of India 1991 Language Atlas [@officeRegistrarLanguageAtlasIndia2004] and Census of India 2001 language tables [@officeoftheregistrargeneralandcensuscommissionerofindiaCensusIndia20012001].](../outputs/figures/appendix/appendix_historical_language_persistence.pdf){#fig-linguistic-persistence fig-pos="H" width="82%"}

An independent reconstruction provides a second validation of the 1991 series. Helms and Lim's replication data include a Shastry-derived 1991 district linguistic-distance measure [@helmsReplicationDataGlobalization2025]. Across the 148 districts shared by their extract and my reconstruction, the two measures correlate at 0.989. The agreement shows that the independently assembled series contain nearly identical cross-district variation.

The measure captures inherited linguistic distance from Hindi. Family members, peers, schools, media, and other social settings also shape the language input children receive [@roweLanguageDevelopmentContext2020].

## Linguistic Conditions and Reported Language Choice {#sec-language-behavior}

I next test whether the Shastry relative-cost measure is associated with the subsidiary languages people report. The Census-2001 bilingualism and trilingualism tables record native-language populations and reported subsidiary languages [@officeoftheregistrargeneralandcensuscommissionerofindiaCensusIndia20012001]. For English and Hindi, the outcome is the share of multilingual native speakers who report the language as a first subsidiary language or, where applicable, a second subsidiary language. The multilingual outcome is the share of native speakers who report any subsidiary language. Each regression observation is a state-by-native-language cell, weighted by its native-speaker population. The specifications compare language groups within states and include the within-state language-prevalence controls defined in @sec-data; the registered continuous and threshold specifications also include the Hindi/Urdu reference indicator.

::: {.landscape}

```{r}
#| label: tbl-main-language-behavior
#| echo: false
#| results: asis

render_public_table("../outputs/tables/main/paper_language_behavior.tex", "paper_language_behavior")
```

:::

The continuous English estimate is 2.23 percentage points per Shastry distance degree, with a standard error of 1.45 ($p=0.123$). The threshold specification is much larger: speakers of languages at least three degrees from Hindi report English at a rate 16.73 percentage points higher than speakers of the remaining classified mother tongues (SE 4.39; $p<0.001$). This specification parallels Shastry's district measure based on the share of speakers whose languages are at least three degrees from Hindi [@shastry2012a].

The largest continuous coefficient in the table is for reported Hindi. Each additional Shastry degree is associated with 9.69 percentage points less reported Hindi among multilingual speakers (SE 2.57; $p<0.001$). Greater distance is also associated with a 3.58 percentage-point decline in the share of native speakers who report any subsidiary language (SE 0.92; $p<0.001$). The pattern therefore includes both the reported subsidiary language and the prevalence of multilingualism.

The alternative linguistic measures produce different English estimates. Glottolog genealogical distance yields a coefficient of 0.21 percentage points (SE 0.43; $p=0.627$), and Dyen lexical noncognacy yields 0.11 percentage points (SE 0.08; $p=0.167$) in the corresponding continuous specifications [@hammarstromGlottologGlottologGlottolog2026; @dyenIndoeuropeanClassification1992; @dyenComparativeIndoeuropeanDatabase1997]. The Shastry threshold specification yields the large English association.

## Context Dependence of Reported Language Choice {#sec-language-context}

Within the predeclared Hindi-belt sample, the continuous English coefficient is $-0.04$ percentage points per Shastry degree (SE 2.94; $p=0.990$). The threshold estimate is 10.16 percentage points (SE 10.81; $p=0.347$). Reported Hindi remains negatively associated with distance at 9.02 percentage points per degree (SE 5.32; $p=0.091$).

The 1991 Scheduled Tribe language data show a different allocation across reported languages [@officeRegistrarLanguageAtlasIndia2004]. In the validated all-state sample, each Shastry degree is associated with 3.38 percentage points more reported Hindi and 0.11 percentage points less reported English. The corresponding bilingualism estimate is 3.64 percentage points. In the Hindi-belt restriction, the Hindi estimate is 5.69 percentage points per degree. The English estimate is $-0.17$ percentage points, and the bilingualism estimate is 5.82 percentage points. That restriction contains four state clusters, so I use it as descriptive evidence. Clingingsmith's historical analysis provides independent evidence that changing economic incentives increased second-language learning in India [@clingingsmithIndustrializationBilingualismIndia2014].

The sex-specific estimates are similar. In the Census-2001 language data [@officeoftheregistrargeneralandcensuscommissionerofindiaCensusIndia20012001], the continuous English coefficients are 2.33 percentage points for men and 2.01 percentage points for women per Shastry distance degree. Geographic and historical comparisons produce much larger differences in the relationship between linguistic distance and reported language.

Across these samples, linguistic distance from Hindi captures persistent variation in the relative cost of learning Hindi and English. The reported subsidiary language associated with greater distance differs across settings. In the national 2001 estimates, greater distance is associated with less Hindi, while the threshold specification shows more English. The English relationship is weak within the Hindi belt. In the 1991 Scheduled Tribe data, greater distance is associated primarily with more Hindi. Reported bridge-language choice therefore varies with setting as well as inherited linguistic conditions. The next section examines formal schooling as one institutional setting in which access to English differs.

# Access to English-Medium Instruction {#sec-access}

## Measuring Access to English-Medium Schooling {#sec-schooling-menu}

Inherited linguistic conditions can change the relative cost of acquiring English, but households can invest only through schooling options they can reach. I therefore use the measures defined in @sec-data to describe realized access to English-medium schooling. The NSS 64th Round records household choices in 2007--08: whether children ages 5--19 are enrolled, the reported medium of instruction among enrolled children, and school management among enrolled children with reported management [@nationalsamplesurveyoffice2008]. These outcomes reflect available schools, household preferences, and household resources. They are not direct measures of school supply.

The schooling measures separate distinct margins of realized access. Enrollment records whether an age-eligible child attends school. EMI among enrolled children records medium of instruction among children already in school, and all-child EMI exposure incorporates both enrollment and reported medium. I also separate public- and private-school EMI exposure and private-school enrollment. School management and medium of instruction identify different schooling choices: households choosing private schools often cite perceived learning quality or school environment more often than English medium itself [@joshiUnderstandingConsumerDemographics2017]. I therefore keep enrollment, EMI among enrolled children, all-child EMI exposure, public EMI, private EMI, and private enrollment separate.

Reported medium of instruction is also distinct from English proficiency and instructional quality. In the rural sample studied by @lahoti2019a, 57% of children described as attending English-medium schools were actually taught in the dominant regional language, while another 18% used English books with explanation in the regional language. I therefore interpret the NSS and DISE measures as classifications of reported school medium rather than direct measures of English acquisition or the amount of English used in class. The NSS also does not observe private tutoring or other supplemental English instruction.

DISE provides a school-reported perspective. Its district aggregates describe the composition of the local school system, including enrollment by medium and management, rather than household survey choices. Enrollment shares still combine available provision with family demand, so I use DISE as a measure of the reported school system rather than a pure supply measure. The agreement between NSS and DISE later in this section is informative because the two datasets measure the same schooling margin from different respondents [@mehtaUDISEPublications2005062022; @diseDistrictReportCardsArchive2017].

@fig-schooling-geography compares the four EMI measures directly. EMI among enrolled children isolates the medium-of-instruction margin among children already in school, while all-child EMI also incorporates enrollment. Splitting all-child exposure by school management shows that public-school EMI is concentrated especially in northern and northeastern districts, whereas private-school EMI is more geographically dispersed. The four panels use the same percentage bins so that differences in both levels and geography can be compared directly.

::: {#fig-schooling-geography layout-ncol=2}
![EMI share among enrolled children.](../outputs/figures/main/map_emi_share_enrolled.pdf)

![EMI exposure among all children.](../outputs/figures/main/map_emi_exposure.pdf)

![Public-school EMI exposure among all children.](../outputs/figures/main/map_public_emi_exposure.pdf)

![Private-school EMI exposure among all children.](../outputs/figures/main/map_private_emi_exposure.pdf)

Geography of English-medium instruction on the common Census-2001 district geography. All panels use a common percentage scale. Source: NSS 64th Round, 2007--08 [@nationalsamplesurveyoffice2008]. `r paper_map_boundary_note()` [@datameetCensus2001Districts; @naturalearthDisputedBoundaries].
:::

## EMI Access as a Function of Institutional Factors {#sec-schooling-market}

The schooling measures are organized very differently across states. Table \ref{tbl-paper-schooling-market-geography} presents the same three geographic specifications for each schooling outcome: an unadjusted association, a specification with region fixed effects and predetermined Census-2001 controls, and a specification with state fixed effects and the same controls. The table reports standardized coefficients so that attenuation can be compared across outcomes, with state-clustered standard errors beneath each estimate. Nationally, linguistic distance is positively associated with every English-medium schooling margin in the table. The association attenuates sharply within states: the state-adjusted standardized coefficient is about 0.12 for enrollment, 0.03 for EMI among enrolled children, 0.06 for all-child EMI, 0.05 for public EMI, 0.09 for private EMI, and 0.05 for DISE English-medium enrollment.

The state-membership $R^2$ rows ask a separate descriptive question: how much district variation is associated with state membership? State indicators explain roughly 84% of linguistic-distance variation. Across the schooling measures, they explain about 83--84% of all-child and conditional EMI, about 83% of DISE English-medium enrollment, and nearly 90% of public EMI, but only about 57% of private EMI and 36% of enrollment. These statistics locate the variation; they do not identify a causal effect of state institutions. Still, the public/private contrast is consistent with the institutional importance of state language policy. Indian state boundaries and schooling systems have historically been organized around language, and @jain2017a shows that aligning official instructional language with local language after linguistic state reorganization substantially improved educational attainment.

```{r}
#| label: render-main-schooling-market-geography
#| echo: false
#| results: asis

render_public_table("../outputs/tables/main/paper_schooling_market_geography.tex", "paper_schooling_market_geography")
```

The NSS pattern is not an idiosyncrasy of household reporting. DISE English-medium enrollment and NSS EMI among enrolled children have a raw district correlation of 0.896, while DISE and all-child NSS EMI exposure correlate at 0.882. After I subtract each state's mean from both measures, those correlations fall to 0.607 and 0.562, respectively. The decline is informative. A raw district correlation combines agreement across states with agreement among districts within the same state; state residualization removes the between-state component and asks only whether districts that are high relative to their own state are also high in the other source. Because the schooling geography above is strongly organized along state boundaries, removing that common state component should reduce the correlations if part of the cross-source agreement reflects broad state-level schooling systems. The remaining within-state correlations are still substantial, which makes simple NSS measurement error an implausible explanation for the weak within-state relationship between linguistic distance and EMI. The corresponding district-level NSS--DISE comparison is shown in @fig-nss-dise-agreement.

Private schooling has a distinct household cost. Even schools described as low-fee can absorb a substantial share of poor households' income [@endow2019a], and private enrollment is strongly selected on household characteristics [@muralidharan2008a; @rajuPrivateSchoolingIndia2023]. Raw public-private achievement differences therefore do not identify differences in school effectiveness. In a randomized voucher experiment in Andhra Pradesh, @muralidharan2015a find no statistically significant difference between lottery winners and losers in Telugu, mathematics, English, or science and social studies after two and four years, despite large cross-sectional differences between public- and private-school students.

Private management is not a measure of instructional quality. The project's DISE indicators of pupil-teacher ratios, single-teacher schools, sanitation, private-sector composition, and EMI do not form a single quality dimension. I therefore use management to distinguish the schooling options households attend, without ranking public and private schools by quality.

## EMI Access as a Function of Social Factors {#sec-social-access}

The existence of options within a district does not imply equal access to them. @fig-unequal-schooling-access compares Scheduled Tribe (ST), Scheduled Caste (SC), and Other Backward Class (OBC) children with children in the residual Other category within common districts. The social-group differences are present for enrollment but substantially larger for English-medium and private schooling. Relative to Other children, mean district enrollment is lower by about 13.6 percentage points for ST children, 9.8 points for SC children, and 6.4 points for OBC children. EMI among enrolled children is lower by about 13.3, 17.2, and 13.2 points respectively. Private enrollment is lower by about 18.1 points for ST children, 23.5 for SC children, and 14.5 for OBC children. Household resources and social institutions can shape these schooling choices [@endow2019a; @rajuPrivateSchoolingIndia2023]. In Mumbai, @munshi2006a show that caste networks continued to direct lower-caste boys toward local-language schools associated with traditional occupations as returns to nontraditional white-collar work rose, while lower-caste girls shifted more rapidly toward English schools.

![Social-group differences in access to English-medium and private schooling. Panel A reports mean within-district percentage-point differences for OBC, SC, and ST children relative to the Other category in enrollment, EMI among enrolled children, private enrollment, and all-child private EMI exposure. Panel B separates rural and urban differences for EMI among enrolled children and private EMI exposure. Source: NSS 64th Round, 2007--08 [@nationalsamplesurveyoffice2008].](../outputs/figures/main/paper_unequal_schooling_access.pdf){#fig-unequal-schooling-access fig-pos="H" width="100%"}

The rural/urban comparison makes the social stratification more pronounced. For SC children, for example, the difference in EMI among enrolled children is about 8.4 percentage points in rural districts and 24.6 points in urban districts; the corresponding difference in private EMI exposure is about 7.5 and 21.4 points. OBC and ST children show the same qualitative pattern. A plausible interpretation is that more differentiated urban schooling markets offer a wider range of formal options while household resources and social position still condition which options are reachable. I treat that as a descriptive sorting pattern rather than evidence that urbanization itself causes the larger differences.

The corresponding decompositions for boys and girls are directionally similar. Government programs operating before the NSS round explicitly targeted girls' education [@pressinformationbureauInitiativesHumanResource2008], so I also examine the sex split directly. In this national cross-section, social group and rural/urban setting distinguish access more strongly than sex.

The result is a constraint on how the schooling menu should be interpreted. A district can contain English-medium and private schools without placing those options within equal reach of every household. The next section asks whether these observed EMI choices are associated with later economic outcomes that might help explain household demand for them.

# Local Economic Outcomes Associated with English-Medium Instruction {#sec-conversion}

## Later Consumption and English-Medium Instruction

English-language skills are a form of human capital [@chiswick1995a; @azam2013a], and evidence from India gives households an economic reason to value them. Using nationally representative 2005 household data, @azam2013a find English fluency associated with a 34% hourly wage premium for men and a 22% premium for women; workers with a little English receive smaller positive premia, and the returns are greater for more educated workers. @refeque2022a later documents substantial English-skill earnings associations through 2011--12. The strongest evidence for a causal wage effect comes from @chakraborty2016a, who exploit West Bengal's withdrawal of English from primary schools and estimate substantially lower later wages for cohorts more exposed to the policy.

Households making schooling decisions in 2007--08 could observe contemporary labor-market rewards to English, including the wage differences documented in this literature. I use later district consumption to describe the economic outcomes eventually associated with the schooling choices observed in 2007--08. These later outcomes do not measure parents' expectations at the time of enrollment.

Later district consumption varies substantially across India. @fig-welfare-geography shows real consumption per person in 2022--23 and the long change from 2004--05 to 2022--23 among districts satisfying the preferred support criteria for both rounds. The figure shows the raw geographic distribution. The regressions below condition on state fixed effects and baseline Census-2001 covariates.

::: {#fig-welfare-geography layout-ncol=2}
![Real consumption per person, 2022--23.](../outputs/figures/main/map_paper_real_consumption_2022_23.pdf)

![Log change in real consumption per person, 2004--05 to 2022--23.](../outputs/figures/main/map_paper_real_consumption_change.pdf)

Later district consumption and long-run consumption growth on common preferred support. Real consumption is expressed in 2011--12 rupees. Sources: NSS 61st Round, 2004--05, and HCES 2022--23 [@nssHouseholdConsumerExpenditure2004; @nsoHCES2022]. `r paper_map_boundary_note()` [@datameetCensus2001Districts; @naturalearthDisputedBoundaries].
:::

Table \ref{tbl-paper-economic-conversion} compares five schooling margins on common district support. Each entry comes from a separate district regression with state fixed effects and baseline Census-2001 covariates. The table places the coefficients on the same scale for comparison. In the 2022--23 ANCOVA specification, a 10 percentage-point increase in enrollment is associated with 0.7% higher real mean consumption (SE 1.70). The corresponding estimates are 3.9% for EMI among enrolled children, 5.1% for all-child EMI, 4.1% for public EMI, and 5.3% for private EMI. The EMI-among-enrolled, all-child EMI, and private-EMI estimates survive the predeclared family correction. In the 2004--05 to 2022--23 long-change specification, all-child EMI is associated with 3.9% higher real mean consumption and private EMI with 5.4%; both survive family correction. The analogous 2023--24 estimates are smaller, and none survives the predeclared family correction.

```{r}
#| label: render-main-economic-conversion
#| echo: false
#| results: asis

render_public_table("../outputs/tables/main/paper_economic_conversion.tex", "paper_economic_conversion")
```

These are conditional district associations. English-medium and private schooling are socially selected, so the coefficients cannot be interpreted as individual returns to attending those schools. The district-consumption coefficients and individual wage returns are different estimands. This distinction is especially important for private schooling: in @muralidharan2015a's voucher experiment, randomized school access yielded similar scores in Telugu, math, English, and science/social studies despite large cross-sectional public--private achievement differences. For the demand question, the relevant pattern is that EMI margins are much more strongly associated with later consumption than enrollment alone. This pattern is also consistent with @azam2013a's evidence that English skills and education are complementary in the labor market.

## Predetermined Complements to English-Medium Instruction

The same Table \ref{tbl-paper-economic-conversion} asks whether the 2004--05 to 2022--23 association varies with three characteristics measured before the schooling round: baseline human capital, urbanization, and ST concentration. The interaction rows standardize each characteristic so that they report how much the schooling association differs with a one-standard-deviation increase. For all-child EMI, the schooling--consumption association is about 1.5 percentage points larger for a one-standard-deviation increase in baseline human capital and about 1.8 points larger for the same increase in urbanization. Both interactions survive the predeclared family correction. The ST-concentration interaction is near zero. The private-EMI interaction estimates are also positive: 1.47 points for baseline human capital and 2.00 for urbanization. They do not survive family correction.

I use *complement* descriptively for a baseline characteristic associated with a larger schooling--consumption slope. Baseline human capital and urbanization fit that description for all-child EMI. The predeclared IT-employment interaction provides no comparable evidence for either all-child EMI or linguistic distance.

The EMI margins most strongly associated with later district consumption are also the schooling margins that are sharply stratified by social group. The all-child EMI association with consumption growth is larger where baseline human capital and urbanization are greater. The inequality argument therefore concerns both access to EMI and variation in the district characteristics associated with a stronger consumption relationship.

## Economic Outcomes as a Function of Linguistic Conditions {#sec-development}

The final part of the local-outcomes analysis examines whether inherited linguistic conditions are associated with later economic change beyond their weak within-state relationship with formal EMI.

The consumption association appears in the long horizon. @fig-consumption-dynamics follows the coefficient on linguistic distance across the registered later-consumption endpoints. The 2009--10 and 2011--12 estimates are small relative to their uncertainty. The 2004--05 to 2022--23 long-change estimate is positive and precisely estimated, and the 2023--24 estimate remains positive with greater uncertainty. The local-development outcomes below help characterize which economic margins accompany this later consumption association. The timing alone does not identify a mechanism.

![Association between linguistic distance and real mean household consumption across later horizons. Points show ANCOVA and long-change coefficients with 95% state-clustered Wald confidence intervals; all specifications include state fixed effects and predetermined Census-2001 controls. Sources: Census of India 2001; NSS 61st, 66th, and 68th Rounds; HCES 2022--23 and 2023--24 [@officeoftheregistrargeneralandcensuscommissionerofindiaCensusIndia20012001; @nssHouseholdConsumerExpenditure2004; @nssHouseholdConsumerExpenditure2009Type2; @nssHouseholdConsumerExpenditure2011Type2; @nsoHCES2022; @nsoHCES2023].](../outputs/figures/main/consumption_iv_dynamics.pdf){#fig-consumption-dynamics fig-pos="H" width="90%"}

Table \ref{tbl-paper-local-development} reports the predeclared local-development outcomes spanning human capital, migration, household financial access, sectoral composition, and employment. Greater linguistic distance is associated with deeper household literacy and greater graduate access. It is also associated with a higher skilled share among recent work in-migrants and larger gains in household banking access. Employment composition shifts toward services and away from manufacturing. Household television ownership also increases. The migration outcome measures the skill composition of recent work in-migrants at their destination. The interstate share among migrants is close to zero in the preferred specification.

```{r}
#| label: render-main-local-development
#| echo: false
#| results: asis

render_public_table("../outputs/tables/main/paper_local_development.tex", "paper_local_development")
```

The local-development pattern is concentrated in human capital, skilled in-migration, financial access, and sectoral composition. Nonfarm employment, labor-force participation, and the employment rate show no corresponding broad increase. The strongest welfare evidence is for real mean MPCE; the alternative welfare measures do not show a common pattern. The Economic Census results are also compositional. Greater linguistic distance is associated with a larger increase in the services employment share and a larger decline in the manufacturing share. Total nonfarm employment shows no comparable association.^[The sector-composition results are strongest with region fixed effects: the raw state-clustered $p$-values are 0.002 for services and 0.004 for manufacturing. Under state fixed effects the corresponding raw $p$-values are 0.031 and 0.027, and both adjust to about 0.110 under the predeclared family correction.]

The sectoral result is consistent with the service- and skill-intensive pattern documented in India's broader development literature. @kochharIndiasPatternsDevelopment2006 emphasize the country's service- and skill-intensive pattern relative to the manufacturing-led path followed by several other fast-growing Asian economies. More recent work stresses how tradeable services can become increasingly important to development as technology changes the tradability of service work [@baldwinGloboticsDevelopmentWhen2023]. These descriptive estimates associate inherited linguistic conditions with human-capital deepening, skilled in-migration, financial access, and sectoral reallocation. They do not identify a causal effect of linguistic distance on structural transformation.

The Census household and housing outcomes cover 2001--11, while the Economic Census changes cover 2005--13; both windows overlap the 2007--08 schooling measurement. These outcomes therefore cannot be ordered as downstream mediators of measured EMI. I treat them as contemporaneous or co-evolving local conditions. The results identify two district characteristics associated with a stronger schooling--consumption relationship: pre-existing human capital and urbanization. Inherited linguistic conditions are also associated with human capital, skilled in-migration, financial access, and sectoral composition. Together these results show substantial variation in the local economic conditions surrounding EMI.

# Conclusion {#sec-discussion}

English skills are often presented as a way for households to loosen the connection between children's prospects and the circumstances into which they are born [@itaniMeaningLanguageSkills2015; @bediEnglishLanguageIndia2019]. The premise is economically plausible: English proficiency carries substantial labor-market rewards in India [@azam2013a; @refeque2022a], and English schooling has been associated with access to occupations outside traditional caste-linked networks [@munshi2006a]. I began from a narrower question. If English-medium schooling is to expand access to valuable English skills, what conditions surround a household's access to it, and are those conditions themselves related to inherited and social inequality?

The evidence points to inequality in each of the three conditions studied in this paper. The linguistic conditions I call inherited are highly persistent across the 1991 and 2001 reconstructions. Their association with reported English and Hindi differs substantially across settings. State schooling systems account for most of the geographic variation in EMI exposure and provision: state membership explains about 84% of the variation in EMI among enrolled children and about 90% of government-school EMI exposure. Within the same districts, realized access to English-medium and private schooling differs sharply across social groups.

The later consumption results show a corresponding economic pattern. In the 2022--23 ANCOVA comparison, a 10 percentage-point increase in all-child EMI is associated with 5.1% higher real mean MPCE, compared with 0.7% for enrollment. The EMI-among-enrolled, all-child EMI, and private-EMI estimates survive the predeclared family correction. In the long-change specification, the all-child and private-EMI estimates also survive that correction. The schooling--consumption association is larger in districts with greater baseline human capital and urbanization. Greater linguistic distance is also associated with deeper human capital, a more skilled composition of recent work in-migration, greater banking access, and a shift in employment toward services. Nonfarm employment, labor-force participation, and the employment rate show no corresponding broad increase. India's service- and skill-intensive development path provides context for the sectoral pattern [@kochharIndiasPatternsDevelopment2006; @baldwinGloboticsDevelopmentWhen2023]. These estimates describe conditional district relationships. They do not identify individual returns or a causal sequence from linguistic conditions through schooling to later development.

The identification analysis reaches a clear limit. Under the preferred state-adjusted specification, linguistic distance has a partial $R^2$ of about 0.004 for all-child EMI, and the registered IV variants fail the effective-$F$ strength criterion. Alternative linguistic-distance measures also remain weak after state adjustment. I therefore cannot use linguistic distance to identify the causal effect of English-medium instruction. This result also reinforces the substantive finding that state schooling systems account for most of the observed geographic variation in EMI.

English skills carry economic returns. The language used to teach school subjects also affects learning. Evidence from India documents educational costs when students are taught through an unfamiliar language [@jain2017a; @nair2015a; @mohanty2022a]. This paper does not estimate the effect of EMI on learning or follow the same children from schooling into later economic outcomes. Migration also changes the population represented by later district outcomes.

Households therefore begin investment in English skills from unequal positions. Children face different inherited linguistic conditions across districts, and realized access to EMI differs sharply by social group within the same districts. The later consumption relationship is also larger where baseline human capital and urbanization are higher. A high-return educational opportunity can coexist with substantial inequality when inherited linguistic conditions, realized access to EMI, and the local conditions associated with its later economic returns are themselves unequally distributed.

# Technical Note for Replication {#sec-replication}

All reported tables and figures are generated through the project's `{targets}` build. Exact R package versions are recorded in `renv.lock`. Required raw files are declared in `data/metadata/file_manifest.csv` with expected paths and sizes; the manifest also records SHA-256 identities for verified inputs. `DATA_AVAILABILITY.md` records availability and redistribution constraints for each required dataset. The appendices document data construction (@sec-app-data), migration (@sec-app-migration), selection (@sec-app-selection), and weak identification (@sec-app-identification). The repository also provides machine-readable results for the larger validation and local-development analyses. `REPLICATION.md` gives the reproduction procedure and system requirements. The full build checks package synchronization and parses the active R and Quarto files before running the test suite. It then renders the manuscript, verifies cross-references and required outputs, and checks reproduction from the tracked processed district inputs.

\newpage

```{=latex}
\appendix
```

# Major Data Contributions {#sec-app-data}

The empirical analysis requires several original data constructions before observations from different years and collections can be compared. I place district observations on a common Census-2001 geography, reconstruct historical DISE medium-of-instruction measures, build district linguistic measures from Census mother-tongue populations and historical language records, and construct comparable real consumption estimates from nine survey designs spanning 2000--01 through 2023--24. I also construct the Census and Economic Census panels used for the local-development outcomes. This appendix describes each construction and the checks used to validate it.

## District Lineage Harmonization {#sec-app-lineage}

District names cannot serve as longitudinal identifiers in India. Between the 1961 and 2001 Censuses the number of districts rose from 339 to 593 through renamings, partitions, mergers, and boundary changes [@kumarCreatingLongPanels2016]. The 2011 Census contained 640 districts after another 47 were created during the intervening decade [@censusIndiaAdministrativeAtlas2011]. I therefore construct a reviewed geographic concordance that records how districts observed in later surveys relate to the 593 Census-2001 districts used as the reference geography.

**Reference geography and evidence.** The 593 Census-2001 districts provide the common geographic basis for the paper. Historical boundary evidence extends back to 1961, and later survey links extend through the 2023--24 HCES round. Kumar and Somanathan's district histories provide the benchmark for pre-2001 partitions and population transfers [@kumarCreatingLongPanels2016]. Census records and the Ministry of Panchayati Raj's Local Government Directory establish later district identities, creations, renamings, and code continuity [@ministryPanchayatiRajLocalGovernmentDirectory2026]. The India State and District Evolution Database, the India District Changes Tracker, and published survey concordances provide additional evidence on district histories and survey-to-Census links [@joladIndiaStateDistrict2026; @jaacksIndiaDistrictChanges2020; @deshpandeIndianEnigma2024]. SHRUG locality keys measure population overlap between 2001 and 2011 places when later districts cross 2001 boundaries [@asherDevelopmentResearch2021].

**Establishing district identity.** Standardized names and spellings generate candidate matches. Accepted identities require documentary or geographic evidence. Each survey district is tied to a Census or LGD unit using its code, an exact or documented renamed identity, or a reviewed record. Ambiguous cases are recorded in a review table containing the decision and supporting evidence. The current build contains 1,254 accepted survey-district identities, and every accepted identity has cited evidence. District reorganizations are recorded as dated predecessor--successor relationships so that splits, mergers, renamings, and unchanged districts remain distinguishable.

**Assigning later districts to Census 2001.** I trace each accepted later unit to its Census-2001 predecessor or predecessors. Unchanged districts and documented renames receive unit weight on the corresponding 2001 district. A later district formed wholly within one 2001 district can also be assigned to that predecessor when the historical evidence establishes a single-parent relationship. Multi-parent districts remain outside the primary analysis. For the full reviewed sensitivity sample, SHRUG locality populations provide fractional weights across 2001 predecessors; accepted weights are nonnegative and sum to one within each later district. This follows the population-weighting principle used by @kumarCreatingLongPanels2016 for historical district partitions.

**Primary and sensitivity samples.** The *conservative* lineage variant includes direct identities, documented renames, and deterministic one-parent transitions. The *primary* variant also includes reviewed single-parent cases supported by Census or LGD code continuity or by independent district-history evidence corroborated with SHRUG locality overlap. Cases using locality overlap require at least 99 percent population coverage. The *full reviewed* variant additionally includes accepted fractional allocations across multiple 2001 predecessors. Across the two schooling waves, these definitions produce 427 common districts under the conservative variant, 573 under the primary variant used in the main analysis, and 587 under the full reviewed variant.

**Aggregating district measures.** Counts are summed across later units assigned to a common 2001 district. Rates, shares, and means are reconstructed from pooled numerators and denominators. Distributional measures such as the consumption Gini are reconstructed from pooled household records when the microdata permit it. These rules preserve the population represented by each observation when district boundaries change.

**Geographic boundaries.** The geographic files use Census-2001 district boundaries. The available 2001 boundary file omits several districts whose boundaries remained unchanged into 2011. I carry back a 2011 polygon only when the Census atlases establish exact continuity [@censusIndiaAdministrativeAtlas2001; @censusIndiaAdministrativeAtlas2011]. The current build uses this rule for eleven districts. The same district identities therefore govern the statistical concordance and the geographic displays.

The three lineage variants make the coverage trade-off explicit while preserving the same interpretation of a district observation. The appendix validation results show that the paper's identification conclusion is unchanged across the conservative, primary, and full reviewed variants.

## DISE Data Reconstruction {#sec-app-dise}

**Recovering the historical files.** I reconstruct historical District Information System for Education (DISE) data from archived workbooks and published district report cards. DISE records schools directly and therefore provides an independent measure of school-system medium of instruction. The original publication site is no longer maintained, and file structures change across years. The surviving publication index maintained by Arun Mehta identifies the district report cards and raw files released by NIEPA/NUEPA and records 2005--06 as the first year with nationwide DISE coverage under a common data-capture format [@mehtaUDISEPublications2005062022]. Using that index, I recovered an Internet Archive snapshot of the former DISE district-report-card site together with the corresponding workbooks and report cards [@diseDistrictReportCardsArchive2017]. The local archive spans 2002--03 through 2015--16. The 2005--06 through 2007--08 files provide the baseline measures used in the paper, and the later years support longitudinal checks.

**Reading year-specific files.** Each academic year has a documented file specification identifying the archived file, any contained ZIP member, the relevant enrollment, school, and teacher sheets, the associated report-card volumes, and known file problems. The year-specific specifications are necessary because workbook layouts and file extensions do not identify a stable location for each measure across vintages. The baseline construction begins in 2005--06, when nationwide coverage becomes comparable [@mehtaUDISEPublications2005062022].

**Decoding medium of instruction.** The 2005--08 enrollment workbooks contain up to five numbered medium slots with grade-specific pupil counts. The slot number does not identify the language. I sum grade counts within each slot and recover the language identity from the corresponding district report card. This reconstruction yields 7,211 district-year-slot language identities, each linked to a report-card PDF and page. English enrollment is the sum of slots identified as English, and Hindi enrollment is constructed in the same way. Positive unidentified slots remain unresolved. When the English slot can be identified, its pupil count remains usable; when unresolved positive slots prevent establishing the English count, the English measure remains missing.

**Reconstructing enrollment and school characteristics.** I reconstruct total Classes I--VIII enrollment separately from the medium slots. The baseline sequence uses grade totals when available, followed by the workbook's direct enrollment total and then the sum of government and private enrollment. Medium-classified enrollment is retained as a consistency measure. English-medium enrollment share equals English-medium pupils divided by total elementary enrollment. I separately sum government and private management categories to construct private enrollment and private-school shares. School and teacher files provide school counts, teacher counts, single-teacher-school counts, and girls'-toilet counts; ratios are calculated after assembling the underlying counts.

**School-age population denominators.** Population-scaled measures use Census counts of children ages 6--13. I first place the 2011 age counts on Census-2001 geography using deterministic district links and then use the 2001 and 2011 anchors to estimate annual district population growth on a log scale. Intervening academic-year populations are obtained by log-linear interpolation. The English-medium gross enrollment ratio divides English-medium pupils by the resulting age-6--13 population. The 2005--08 baseline measures use interpolation; post-2011 extensions appear only in longitudinal checks.

**District harmonization.** DISE pupil and school counts use a stricter geographic rule than the general lineage system. A DISE district is assigned to Census-2001 geography only through an exact identity or a reviewed one-to-one link with weight one. Several later districts can be pooled into one 2001 district when each link is deterministic. Pupil, school, and teacher counts are summed before recomputing shares and ratios.

**Checks against published DISE reports.** Published report cards both identify medium slots and validate selected reconstructed cells. Three registered 2005--06 medium-of-instruction cells for Kupwara reproduce the published report exactly. Later report-card pages provide English and Hindi pupil counts when workbook slot identities are unstable. The 2009--10 archive lacks its second report-card volume, so the reconstruction uses the archived complete *District Elementary Education Report Card: 2009--10* for affected pages [@nuepaDistrictElementaryEducation2009]. The 2010--11 enrollment workbook has a documented header/data alignment problem; matched published district totals provide the current-year denominator in affected cases. These year-specific rules are defined before estimation.

**Independent comparison with NSS.** The independently reconstructed DISE and NSS measures agree strongly in district levels. For 2007--08, the correlation between DISE English-medium enrollment share and NSS English-medium instruction among enrolled children is 0.896 across 520 districts. After subtracting state means from both series, the correlation remains 0.607. The within-state correlation shows substantial agreement between two independently constructed measures after removing their common state component.

![Independent agreement between NSS and DISE English-medium measures. The left panel compares district levels; the right panel subtracts each source's state mean before comparison. Sources: NSS 64th Round, 2007--08 [@nationalsamplesurveyoffice2008] and DISE District Report Cards, 2007--08 [@diseDistrictReportCardsArchive2017].](../outputs/figures/appendix/appendix_nss_dise_agreement.pdf){#fig-nss-dise-agreement fig-pos="H" width="90%"}

## Linguistic Measures {#sec-app-linguistic-measures}

The linguistic measures summarize differences in the relative cost of acquiring Hindi and English as languages of wider communication, following @shastry2012a. I begin with mother-tongue-level distance from Hindi and combine those scores with each district's distribution of mother-tongue speakers.

**Census speaker population.** The speaker distribution comes from the mutually exclusive mother-tongue rows of Census 2001 Table C-16 [@officeoftheregistrargeneralandcensuscommissionerofindiaCensusIndia20012001]. Language-group subtotals are excluded from the speaker count. This treatment is important for groups such as Census Hindi, which contains distinct mother tongues including Bhojpuri and Rajasthani varieties. Distance is assigned at the mother-tongue level before district aggregation.

**Shastry distance measures.** The preferred language scores use Shastry's ordinal 0--5 classification [@shastry2012a]. Hindi and Urdu form the zero-distance reference. Published Indo-European languages retain Shastry's reported degree. Languages outside Indo-European receive degree five after review of their family classification. Native English is handled separately and excluded from the weighted-distance mean. The primary *speaker-weighted distance from Hindi* is the Census-speaker-weighted mean among classified mother tongues with positive Shastry distance. The older *top-three-language distance* uses only the three largest classified mother tongues in each district. The primary measure uses the full classified mother-tongue distribution; I retain the top-three measure for comparison with the earlier analysis.

**Language classification review.** Some Census mother tongues have no unambiguous published Shastry category. I apply the published Shastry concordance first. Finer Indo-European mother tongues absent from that concordance are classified using historical evidence from the *Linguistic Survey of India* [@griersonLinguisticSurveyIndia1903] and a named published Shastry anchor. Lexical evidence from @koganGenealogicalClassificationNew2017 and ASJP v21 [@sorenwichmannCLDFDatasetDerived2025] helps distinguish nearby classifications when needed. The reviewed classification table records accepted assignments and unresolved rows. Mother tongues without sufficient evidence remain unclassified in the preferred construction. Regression results play no role in these classification decisions.

**Population shares and unresolved languages.** Unclassified speaker mass is retained explicitly. The preferred weighted mean uses classified positive-distance speakers in its numerator and denominator, with the classified-speaker share reported separately. Population-share measures keep unresolved and native-English speakers in the district population denominator. The *share speaking languages distant from Hindi* is the percentage of all mother-tongue speakers whose classified language is at least three Shastry degrees from Hindi. Five additional shares record district speaker mass at each positive distance degree, with distance zero serving as the omitted category when the shares enter jointly. Low- and high-degree assignments for reviewed ambiguous languages provide bounding sensitivity checks while the preferred assignments remain fixed.

**Historical 1991 measure.** I reconstruct a separate 1991 district-language series from the *Language Atlas of India 1991*, which reports the geographic distribution of 114 classified languages [@officeRegistrarLanguageAtlasIndia2004]. The Atlas is distributed as a PDF, so I extract district-by-language counts from the text and page coordinates embedded in its district sheets. Ambiguous cells are recorded for review and remain unresolved until supported by the page content. Districts are identified using the 1991 Census state and district codes [@censusIndiaCodingManual1991]. Extracted district populations are checked against the 1991 Primary Census Abstract before language shares are calculated. Atlas languages are then linked to the same Shastry basis, and accepted district-language counts are checked against Atlas population totals.

Unresolved language mass enters explicit bounds on the weighted mean. The preferred historical series requires at least 99 percent accepted speaker coverage and a distance-bound width no greater than 0.5 degree. Eligible 1991 districts are then placed on Census-2001 geography using @sec-app-lineage. The resulting series supports the 1991--2001 persistence analysis and the independent reconstruction checks reported in the main text.

**Genealogical and lexical alternatives.** Two alternative constructions use different linguistic information. The *genealogical distance from Hindi* uses the reviewed Census-to-Glottolog 5.3 crosswalk and counts separation along Glottolog's language-family genealogy [@hammarstromGlottologGlottologGlottolog2026]. The edge count remains on its native genealogical scale. The *lexical noncognacy with Hindi* measure uses the Dyen--Kruskal--Black Indo-European cognate data [@dyenIndoeuropeanClassification1992; @dyenComparativeIndoeuropeanDatabase1997]. At the language level it equals one hundred minus the percentage of cognate items shared with Hindi. Following Shastry's convention, reviewed non-Indo-European languages outside the Dyen comparison receive five percent shared words, giving a noncognacy value of 95. Unresolved Indo-European mother tongues remain unresolved. All three distance constructions are defined independently of their regression results.

## Price Series and Consumption Harmonization {#sec-app-consumption-construction}

I construct comparable district consumption estimates from nine survey designs spanning NSS 56 in 2000--01 through HCES 2023--24, including both NSS 66 questionnaire designs. The main longitudinal analysis uses NSS 61 as the baseline, NSS 66 Type 2 and NSS 68 Type 2 as intermediate endpoints, and the two HCES rounds as long-run endpoints. The other registered designs provide pretrend, treatment-period, and questionnaire-design checks. Survey questionnaires, recall periods, district codes, and price environments differ across these rounds [@nssHouseholdConsumerExpenditure2004; @nationalstatisticalofficeIndiaHouseholdConsumer2011; @nssHouseholdConsumerExpenditure2009Type1; @nssHouseholdConsumerExpenditure2009Type2; @nssHouseholdConsumerExpenditure2011Type2; @nsoHCES2022; @nsoHCES2023]. I reconstruct household monthly per-capita consumption expenditure (MPCE) within each design, validate nominal MPCE against published survey totals, and then harmonize prices and district geography. The paper uses real consumption per person across rounds.

**NSS consumption definitions.** For each NSS Schedule 1.0 design, I retain its published consumption concept. The metadata record the household MPCE field, scale, survey-design variables used for weighting, and whether the published benchmark uses the mixed reference period (MRP) or modified mixed reference period (MMRP). Under MRP, less frequently purchased categories use a 365-day recall and most other spending uses 30 days; MMRP additionally uses a 7-day recall for selected food and related items [@nsoHCESReport2023]. Some releases report MPCE directly, while others require reconstruction from split household files. All 18 reconstructed rural and urban benchmark means across the nine surveys differ from their published counterparts by less than ₹0.50. A survey round enters the district series only after passing its national rural and urban MPCE benchmark.

**HCES three-visit reconstruction.** The modern HCES releases divide consumption across three questionnaires administered in separate visits: the Food Questionnaire (FDQ), Consumables and Services Questionnaire (CSQ), and Durable Goods Questionnaire (DGQ) [@nsoHCES2022; @nsoHCESReport2023]. Items use 7-, 30-, or 365-day reference periods. I convert each registered expenditure amount to a monthly equivalent and sum the three questionnaire components by household. Imputed house and garage rent is excluded in accordance with the published MPCE concept used for benchmarking. The 2023--24 release identifies visit order, so I use the third-visit multiplier prescribed by the official estimation procedure [@nsoHCESReport2023]. The 2022--23 household is retained when its three questionnaire records carry the same multiplier, which supplies the common household weight. A complete HCES consumption observation requires all three questionnaire components.

**Temporal price adjustment.** The *temporal price relative* measures price change across survey dates within a state and rural or urban sector. Before 2013, rural prices use the Labour Bureau's Consumer Price Index for Rural Labourers (CPI-RL), whose 1986--87-base series covers 20 states and all-India [@labourBureauCPIRL]. Urban prices use the Labour Bureau's 2001-base Consumer Price Index for Industrial Workers (CPI-IW), aggregated from centre series to state level [@labourBureauCPIIW]. From 2013 onward, I use state rural and urban CPI series with base 2012=100 from the national CPI system [@reserveBankIndiaDBIE]. Common state-sector months in 2013--14 link the earlier series to the 2012-base CPI. A state-sector link is available only when that overlap is observed. A separate overlap between the 2010- and 2012-base rural/urban CPI series places the July 2011--June 2012 reference-year index on the same scale.

**Spatial price adjustment.** The *spatial price relative* preserves cross-state and rural/urban price-level differences using the state-specific 2011--12 Tendulkar poverty lines [@planningCommissionPovertyEstimates2013]. Each state-sector poverty line is divided by the all-India rural reference of ₹816. A common denominator preserves the national rural-urban price difference. Direct state observations are used when available. For small states or union territories lacking a historical index, the price table permits documented donor or all-India substitutions and records the basis for each substitution.

**Household deflation and survey timing.** The household price deflator is

$$
P_{srt}=P^{\text{spatial}}_{sr}\times P^{\text{temporal}}_{srt},
$$

where $s$ denotes state, $r$ rural or urban sector, and $t$ survey time. Real MPCE is nominal MPCE divided by this deflator. For NSS rounds, I average monthly state-sector indices over the three months represented by the household's survey sub-round. For HCES, price assignment follows the household's three-month panel.

**District aggregation and support rules.** I attach each household's survey district to Census-2001 geography using @sec-app-lineage. Geographic assignment coverage, measured by represented person weight, ranges from 92.6 percent in 2000--01 to 100 percent in 2007--08, is about 97--98 percent in 2009--12, and is about 91.6 percent in each modern HCES round. HCES publishes consumption estimates at the national and state or union-territory levels [@nsoHCESReport2022]. I construct the district estimates used here from household microdata and retain district sample information with every estimate. District means are person-weighted by multiplying the household survey weight by household size.

The preferred district estimates must also satisfy predeclared sample-support rules. I record the number of sampled households and primary sampling units and calculate a Kish effective sample size, which summarizes effective information under unequal survey weights. Comparisons requiring a fixed district set use the intersection of districts satisfying the relevant support rules in every compared round or outcome. Geographic assignment coverage and statistical support are therefore recorded separately.

**Cross-round validation.** The adjacent modern HCES rounds provide a direct validation under a nearly unchanged three-visit survey design. On preferred common support, the 2022--23 and 2023--24 district estimates have Pearson correlations of 0.859 for real mean MPCE, 0.854 for mean log real MPCE, and 0.860 for weighted median real MPCE. These correlations show consistent district variation across adjacent rounds. The older NSS rounds require the survey-definition and price adjustments described above, while the national benchmark checks verify each reconstructed nominal MPCE series before those adjustments are applied.

![District-level consistency of reconstructed consumption across the 2022--23 and 2023--24 HCES rounds. Each panel uses districts satisfying the preferred support rule in both rounds; the dashed line is the 45-degree line. Sources: National Sample Survey Office, HCES 2022--23 and HCES 2023--24 [@nsoHCES2022; @nsoHCES2023].](../outputs/figures/appendix/appendix_consumption_hces_consistency.pdf){#fig-app-consumption-hces fig-pos="H" width="90%"}

## Additional Harmonized Outcome Panels {#sec-app-additional-data}

For the 2001--11 Census comparisons, I reconstruct migration, household human-capital, housing and asset, and worker-industry measures from published district counts [@officeoftheregistrargeneralandcensuscommissionerofindiaCensusIndia20012001; @censusIndia2011]. Eligible 2011 counts are pooled onto Census-2001 geography before rates are calculated, so numerators and denominators refer to the same places in both Census years.

For the Economic Census, I use SHRUG's district products for the 2005 and 2013--14 measures that are available comparably across rounds [@asherDevelopmentResearch2021]. The predetermined 2005 information-technology measure requires finer industry detail, so I read the official Fifth Economic Census establishment records directly [@centralstatisticsofficeIndiaFifthEconomic2008]. Those records identify district, activity status, four-digit NIC-2004 activity, and total workers. I retain major-activity nonagricultural establishments and define computer-related activity as NIC-2004 Division 72. This variable is used as a 2005 baseline characteristic because the available Sixth Economic Census files do not support the same four-digit NIC-2004 definition [@office2019a]. The 2009--10 NSS employment survey and 2017--18 PLFS provide labor-force participation and employment measures for resident workers [@nationalsamplesurveyofficeIndiaEmploymentUnemployment2010; @nationalstatisticalofficeINDIAPeriodicLabour2019]. Economic Census employment measures workers attached to establishments. These two measurement concepts are kept separate in the analysis.

# Migration {#sec-app-migration}

## Literature Review

Migration changes the interpretation of district outcomes. The schooling measures describe children observed in each district in 2007--08. Later consumption describes the district population many years afterward. Population turnover means changes in district consumption need not describe the later outcomes of the same children. I therefore interpret the long-run consumption results as district-level outcomes for changing populations. They do not estimate individual returns for the children observed in the earlier schooling data.

Earlier studies documented low interdistrict mobility in parts of India. @chakraborty2016a report low mobility in West Bengal, Haryana, and Punjab and use district of schooling to approximate later district of residence or employment. Using earlier NSS rounds, @topalovaTradeLiberalizationPoverty2005 similarly documents low decadal interdistrict mobility, especially in rural areas and for employment-related moves. Those estimates concern migration over relatively short durations. Lifetime-migrant measures accumulate moves over much longer exposure periods as respondents age [@mahapatroInternalMigrationEmerging2020].

Census-cohort evidence and railway passenger flows indicate substantially greater interdistrict movement after 2001 than the late-twentieth-century evidence implied [@divison2015a]. Recent work also links internal migration and skill acquisition to regional economic opportunities. @ghoseTradeInternalMigration2026 studies India's information-technology expansion in a model where education choices and internal migration respond jointly to regional opportunities. Migration also faces substantial social and geographic frictions. Caste-based rural insurance networks can discourage migration when departure weakens access to informal insurance [@munshiNetworksMisallocationInsurance2016], and state borders sharply reduce migration between neighboring districts [@koneInternalBordersMigration2018].

My migration measures come from the 2011 Census D-series tables [@censusIndia2011]. D-02 describes migrants by place of last residence and provides the interstate share among migrants. D-03 records reason for migration. D-07 identifies migrants who have been at the destination for at most nine years, report work or employment as the reason for migration, and previously resided elsewhere in India. I call these 0--9-year work migrants *recent work migrants*. I verify the D-07 total against the corresponding D-03 work-migrant subtotal before constructing the measures. I then pool the 2011 counts to Census-2001 districts using the reviewed district concordance described in @sec-app-lineage.

Overall migrant composition combines moves for marriage, education, household relocation, and work. The recent-work-migrant measure isolates people who migrated for employment and describes the skill composition of workers newly observed at each destination. It measures work in-migrants at the destination. It contains no corresponding measure of workers leaving that district.

## Empirical Results

@tbl-app-migration reports the registered state-fixed-effect regressions of migration outcomes on linguistic distance. Each coefficient compares districts within states and conditions on the predetermined Census controls used elsewhere in the paper. Panel A reports the interstate, work-reason, and education-reason composition of all migrants. Panel B reports geographic origin and educational composition among recent work migrants, together with the prespecified Hindi-belt restriction.

```{r}
#| label: tbl-app-migration
#| echo: false
#| results: asis
render_public_table("../outputs/tables/appendix/appendix_migration_summary.tex", "appendix_migration_summary")
```

Linguistic distance has little association with migrants' geographic origin or stated reason for migration in these regressions. A one-degree increase in linguistic distance is associated with a 0.23 percentage-point lower interstate share among all migrants (SE 1.66 percentage points), a 0.60 percentage-point higher work-reason share (SE 0.50), and a 0.04 percentage-point lower education-reason share (SE 0.10). Among recent work migrants, the outside-state share is 0.50 percentage points higher per degree (SE 2.66). The Holm-adjusted $p$-value is 1.000 for each of these four associations. Separate 2001--11 population-denominated migration changes are reported descriptively outside this registered family.

The clearest migration result concerns educational composition. A one-degree increase in speaker-weighted linguistic distance from Hindi is associated with a 1.29 percentage-point higher graduate-or-technical-degree share among recent work migrants (SE 0.44 percentage points; raw $p=0.0033$; Holm-adjusted $p=0.0266$). The dependent variable is the graduate-or-technical-degree share among recent work in-migrants observed at the destination. The pattern is consistent with systematic differences in the skill composition of work in-migration across destinations. The regression is descriptive and does not identify why that composition differs.

Within the Hindi-belt states, the corresponding estimate falls to 0.79 percentage points and is less precise (SE 0.57 percentage points, $p=0.171$).

Linguistic distance is therefore associated most clearly with the educational composition of recent work in-migrants. The corresponding associations with interstate origin and migration reason are small or imprecise. This compositional difference is one reason I interpret later consumption as a district-level economic outcome for a changing population.

# Selection into School Enrollment {#sec-app-selection}

The district comparisons in the main text show large social-group differences in enrollment, English-medium instruction, and private schooling within the same districts. I use the child-level enrollment model to examine the earlier participation margin: whether a child appears in the enrolled-child education record at all. I treat the estimates as descriptive associations and keep this analysis separate from the district-level identification design.

I use the NSS 64th Round *Participation and Expenditure in Education* survey and restrict the child roster to ages 5--19 [@nationalsamplesurveyoffice2008]. I define enrollment by whether a child in the household roster has a corresponding enrolled-child education record. The probit includes age, sex, household size, religion, social group, rural or urban residence, and distance to the nearest primary class. It also includes a household education measure constructed from the education of the male household head; when no male head is available, the construction uses a married son and then a father or father-in-law. The fitted specification also includes district averages of selected schooling costs and benefits among enrolled children. Those district measures combine local provision with the composition of enrolled children, so I retain them as adjustment variables and omit their coefficients from the reader-facing table.

The probit uses the NSS survey design, including household weights, sampled villages or urban blocks as primary sampling units, and strata formed from state, stratum, and substratum identifiers. I report quantities on the predicted-probability scale. Continuous covariates are summarized by average slopes. Categorical covariates are summarized by average discrete comparisons against the stated reference category, averaging over the observed estimation-sample covariate distribution [@hansenEconometrics2022; @arel-bundockHowInterpretStatistical2024].

```{r}
#| label: tbl-appendix-selection-ame
#| echo: false
#| results: asis
render_public_table("../outputs/tables/appendix/appendix_selection_ame.tex", "appendix_selection_ame")
```

Social identity is strongly associated with enrollment. Conditional predicted enrollment is 12.6 percentage points lower for Muslim children than for Hindu children and 4.3 points lower for Christian children. The Muslim difference is consistent with evidence of persistent educational disadvantage among Indian Muslims [@borooah2021a]. Relative to children in the residual "Other" social-group category, predicted enrollment is 10.2 points lower for Scheduled Tribe children, 7.5 points lower for Scheduled Caste children, and 2.9 points lower for Other Backward Class children. These are conditional associations. The model does not identify causal effects of religion or caste. Social institutions and caste networks can shape schooling choices and later occupational opportunities [@munshi2006a].

Male household education has the largest ordered association in the model. Relative to an illiterate designated male household member, predicted enrollment is 9.7 percentage points higher for the literate-without-schooling category, 14.2 points higher at primary education, 19.5 points higher at upper primary, 21.2 points higher at secondary, and 24.7 points higher at higher secondary. The postsecondary comparison is 24.5 points. This measure is correlated with household resources, information, and intergenerational educational advantage. The model does not separate those channels.

Within the fitted specification, one additional year of age is associated with a 2.8 percentage-point lower predicted enrollment probability on average. Girls have a 4.4-point higher predicted probability than boys conditional on the other covariates. Each additional household member is associated with a 0.37-point lower predicted probability. These estimates describe the conditional pattern in the NSS sample.

Nearby school access also has a sizable association with enrollment. Children whose nearest primary class is reported one to two kilometers away have a predicted enrollment probability 7.7 percentage points lower than children with a primary class within one kilometer. The estimates for the more distant categories are imprecise and do not form a monotone pattern. The data therefore support a distinct one-to-two-kilometer enrollment difference without a stable distance-response relation across the full set of categories.

The fitted model includes district averages of free schooling, fee waivers, scholarships, textbooks, stationery, mid-day meals, and enrollment cost among enrolled children. These terms remain adjustment variables because each combines local schooling conditions with selection into enrollment. Their coefficients are omitted from the reader-facing table.

The child roster contains 127,181 observations. The complete-case estimation sample contains 114,898 children, leaving 12,283 observations with at least one required covariate missing. District-average enrollment cost is missing for 5.6 percent of the roster, male household education for 4.1 percent, and distance to the nearest primary class for 0.2 percent. Observed covariates explain considerably more of missingness in district-average enrollment cost and school distance than of missing male household education: the corresponding missingness-logit pseudo-$R^2$ values are 0.310, 0.246, and 0.023. Complete cases therefore differ systematically from omitted observations on measured characteristics. No missing-value imputation is used.

```{r}
#| label: tbl-appendix-selection-missingness
#| echo: false
#| results: asis
render_public_table("../outputs/tables/appendix/appendix_selection_missingness.tex", "appendix_selection_missingness")
```

The child-level estimates show that unequal educational participation appears before the choice of school medium. Religion, social group, and especially male household education are strongly associated with enrollment itself, adding an earlier participation margin to the within-district inequalities in English-medium and private schooling documented in the main text.

\newpage

# Linguistic Distance as an Instrumental Variable {#sec-app-identification}

The main text uses linguistic distance as a measure of inherited linguistic conditions. I next evaluate it as a candidate instrument for English-medium schooling. The design requires linguistic distance to predict EMI after the preferred geographic and baseline adjustment and to affect later consumption only through the schooling channel. The first requirement is weak within states, and the exclusion restriction remains substantively demanding. I therefore use this appendix to assess first-stage relevance, alternative linguistic and schooling measures, weak-identification-robust inference, and sensitivity to direct effects of linguistic distance. Weak instruments can invalidate conventional 2SLS inference even when familiar first-stage rules appear reassuring [@keanePracticalGuideWeak2024].

## First-Stage Relevance

Linguistic distance strongly predicts EMI in the unadjusted national comparison. On the common sample used for the registered first-stage comparison, the excluded-instrument $F$ statistic is 11.73 without geographic fixed effects or controls. It falls to 3.24 with region fixed effects and the main Census-2001 controls and to 0.75 with state fixed effects and the same controls. @fig-first-stage-absorption shows the corresponding partial relationships after applying each adjustment, and @tbl-app-iv-relevance reports the same progression across alternative linguistic measures. The state-fixed-effects partial $R^2$ is 0.0039, so linguistic distance accounts for less than half of one percent of the residual variation in all-child EMI exposure under the preferred adjustment.

![Partial relationship between linguistic distance and all-child EMI exposure under progressively finer geographic adjustment. Points are equal-frequency bin means on a common sample; lines are OLS first-stage fits. The region and state panels residualize both axes using the corresponding fixed effects and the main Census-2001 controls. Panel labels report state-clustered HC1 $F$ statistics. Sources: Census of India 2001 and NSS 64th Round, 2007--08 [@officeoftheregistrargeneralandcensuscommissionerofindiaCensusIndia20012001; @nationalsamplesurveyoffice2008; @asherDevelopmentResearch2021].](../outputs/figures/main/poster_first_stage_specs.pdf){#fig-first-stage-absorption fig-pos="H" width="100%"}

The conventional first-stage $F$ statistic describes how relevance changes as geographic adjustment is added. For the preferred IV specifications, I assess strength with the Montiel Olea--Pflueger effective $F$, which is designed for weak-instrument assessment under heteroskedasticity and clustering [@montieloleaRobustTestWeak2013]. A fixed rule based on the conventional $F$ is insufficient for this purpose because the size distortions of conventional 2SLS inference depend on the design [@keanePracticalGuideWeak2024].

The attenuation identifies the geographic scale at which the linguistic-distance/EMI association is concentrated. Linguistic distance and EMI are strongly related across broad Indian geography. The association becomes much smaller when districts are compared within the same state. This pattern is consistent with the state-level organization of English-medium schooling documented in the main text, especially in government schools. The remaining within-state variation in linguistic distance therefore provides little predictive content for EMI exposure.

## Spatial Dependence

EMI exposure, linguistic distance, consumption growth, and changes in consumption inequality are geographically concentrated, so I test the preferred first- and second-stage residuals for positive spatial autocorrelation. I construct row-standardized contiguity weights from the Census-2001 district polygons and calculate Moran's $I$, a global measure of whether neighboring districts have similar values [@anselin2001]. Under rook contiguity, English-medium exposure has $I=`r report_value("moran_emi_i")`$, linguistic distance has $I=`r report_value("moran_linguistic_distance_i")`$, real consumption growth has $I=`r report_value("moran_consumption_growth_i")`$, and the change in the Gini coefficient has $I=`r report_value("moran_gini_change_i")`$; each test rejects the null of no positive spatial autocorrelation at $p<0.001$.

For the preferred consumption specification, the second-stage residuals have Moran's $I=`r report_value("moran_iv_residual_i")`$ ($p=`r report_value("moran_iv_residual_p")`$), and the first-stage residuals have $I=`r report_value("moran_first_stage_residual_i")`$ ($p=`r report_value("moran_first_stage_residual_p")`$). Neither test rejects the null of no positive residual spatial autocorrelation. Queen contiguity gives the same result: the second-stage residual test has $p=`r report_value("moran_iv_residual_p_queen")`$ and the first-stage residual test has $p=`r report_value("moran_first_stage_residual_p_queen")`$. Lakshadweep, South Andaman, and Nicobars have no contiguous district neighbor and enter with zero spatial lag. The raw variables are strongly spatially clustered, while the preferred residual tests provide no evidence of positive spatial autocorrelation under either contiguity definition.

## Alternative Measures and Historical Constructions

The same weak within-state relevance appears across alternative linguistic measures. @tbl-app-iv-relevance compares the preferred speaker-weighted measure with scalar alternatives based on the largest languages, genealogical distance, lexical noncognacy, and the population share speaking languages distant from Hindi. It also reports a five-component language-distance composition and the independently reconstructed 1991 measure. For each construction, I report the excluded-instrument $F$ statistic without geographic fixed effects, with region fixed effects, and with state fixed effects. The final column gives the state-fixed-effects partial $R^2$.

```{r}
#| label: tbl-app-iv-relevance
#| echo: false
#| results: asis
render_public_table("../outputs/tables/appendix/appendix_iv_relevance_summary.tex", "appendix_iv_relevance_summary")
```

Each scalar alternative loses most of its first-stage relevance under state fixed effects. The top-three measure falls from an $F$ statistic of 16.30 to 1.96. The genealogical and lexical measures fall from 10.43 and 5.29 to 0.28 and 0.15, and the distant-language share falls from 4.08 to 0.59. The five-component specification retains more within-state predictive content than the scalar measures: its joint state-FE $F$ is 3.73 and its partial $R^2$ is 0.041. These values remain small for an overidentified IV specification with five excluded instruments.

The independently reconstructed 1991 measure is even weaker under geographic adjustment. On its validated common support, the 1991 speaker-weighted distance has an $F$ statistic of 9.98 without geographic fixed effects, 0.31 with region fixed effects and Census controls, and 0.03 with state fixed effects and the same controls. Measuring the linguistic environment in 1991 gives it clear temporal precedence over the 2007--08 schooling measure. That timing does not establish exclusion or supply a strong first stage. Pretreatment balance and negative-control exercises provide additional information about the design, while their interpretation depends on the specification and neither test establishes the exclusion restriction by itself [@danieliNegativeControl2026].

The NSS--DISE comparison shows that weak state-FE relevance is also present with an independently reconstructed schooling measure. The two EMI measures correlate at about 0.896 in district levels and 0.607 after removing state means (@fig-nss-dise-agreement). EMI measurement therefore retains substantial within-state agreement across the household and school data even though linguistic distance predicts that variation weakly. Alternative language classifications and distance constructions lead to the same state-FE conclusion.

The richer language-share specifications also remain weak under state fixed effects. Their constituent first stages require caution, and all three registered state-FE falsification-adaptive sets include zero. A falsification-adaptive set reports structural values that remain admissible under the smallest relaxations needed to reconcile the overidentifying restrictions with the data [@mastenSalvagingFalsified2021]. In these specifications, the sets provide no sign-stable structural estimate.

## Weak-IV-Robust Inference

The weak first stage changes the appropriate inference for the second stage. I report conventional 2SLS coefficients for completeness and base inference on methods that remain valid under weak identification. The preferred assessment uses the Montiel Olea--Pflueger effective $F$ statistic and Anderson--Rubin (AR) tests. The AR test evaluates candidate structural coefficients without requiring a strong first stage [@keanePracticalGuideWeak2024].

@tbl-app-iv-weak reports the two long-change specifications used in the main identification discussion. The table places the conventional 2SLS coefficients beside the weak-identification statistics and omits conventional significance stars.

```{r}
#| label: tbl-app-iv-weak
#| echo: false
#| results: asis
render_public_table("../outputs/tables/appendix/appendix_iv_weak_inference.tex", "appendix_iv_weak_inference")
```

For 2004--05 to 2022--23, the 2SLS coefficient is 0.117 with a state-clustered standard error of 0.152 and an effective $F$ statistic of 0.69. For the 2023--24 endpoint, the coefficient is 0.137 with a standard error of 0.234 and an effective $F$ of 0.41. These effective-$F$ values indicate extremely weak identification.

The AR results make the inferential consequence explicit. At zero, the AR $p$-values are 0.001 and 0.014. Inverting the test over the registered grid produces disconnected 95% accepted sets: the 2022 set includes values at or below approximately $-0.071$ and at or above $0.025$, and the 2023 set includes values at or below approximately $-0.051$ and at or above $0.015$. Both sets reach the grid boundaries. The rejection of zero at the point test therefore does not identify the sign of the structural effect; negative and positive values both remain in the accepted set on the evaluated grid.

The registered robustness families reach the same conclusion. They vary the linguistic measure, EMI treatment, welfare outcome, controls, parameterization, and historical adjustment. Together they contain 408 model specifications across seven families, and none satisfies the registered first-stage strength criterion. I retain the model-level results in the replication outputs and summarize the family-level evidence here.

I also evaluate sensitivity to violations of the exclusion restriction. I use a weak-identification-robust bounded-direct-effect analysis in the spirit of plausibly exogenous instruments, allowing linguistic distance to have a direct effect on the outcome and asking how large the permitted effect must be before $\beta=0$ enters the AR confidence set [@conleyPlausiblyExogenous2012]. For the 2022 long change, the minimum bound is about 0.0267 log points, equal to 41.7% of the observed reduced form. For 2023, the corresponding threshold is about 0.0112 log points, or 19.9% of the reduced form. These numbers are sensitivity thresholds. They do not estimate the direct effect of linguistic distance. For 2023, allowing a direct effect equal to about one-fifth of the observed reduced form is sufficient for the AR set to include zero; the corresponding share for 2022 is about 42%. The first stage remains weak in both specifications.

I therefore do not use linguistic distance to estimate a causal effect of English-medium schooling. The evidence supports using linguistic distance as a measure of inherited linguistic conditions. Within-state relevance for EMI is extremely weak, and linguistic distance is also associated with several other local characteristics, so exact exclusion requires assumptions that the data cannot verify. I therefore interpret the schooling--welfare estimates in the main text as descriptive district associations.

# References {-}
