---
title: "Inequality in a Potential Equalizer of Opportunity"
subtitle: "Variation in the Accessibility and Net Benefits of English-Medium Instruction in India"
author: "Rishav Roy"
thanks: "Acknowledgments: I would like to thank Laura Schechter, Corina Mommaerts, and Agustín Gutiérrez, for their feedback on earlier stages of this project. The Internet Archive also proved indispensable for this project; a plurality of the data sources I use were only accessible because of their preservation work. GPT models 4.5, 5, 5.4, 5.5, and 5.6 were used for troubleshooting, proofreading, and generating parts of this paper and its code from detailed outlines and pseudocode. All ideas, analytical work, and errors are my own."
format:
  pdf:
    pdf-engine: xelatex
bibliography: references.bib
execute:
  echo: false
  warning: false
  message: false
abstract: "English-medium instruction (EMI), or the teaching of school subjects in English, is often viewed as a potential palliative for the integenerational rigidty of socioeconomic status in India, where English skills command substantial wage premia. I study whether the conditions governing households’ access to and potential net gains from EMI are themselves unequally distributed. I organize my analysis around three such conditions: inherited linguistic costs of acquiring English, access to English-medium schooling, and the local economic setting in which English skills may be rewarded. By harmonizing district data on languages, schooling, consumption, migration, human capital, and sectoral employment, I find inequality across all three. Inherited linguistic conditions are highly persistent, but their effects on language acquisition vary across insitutional settings. EMI exposure and provision are strongly organized by schooling system and sharply stratified along socioeconomic lines, even within the same local education market. Consumption growth is more associated with EMI enrollment than enrollment alone, and their strength increases in pre-existing human capital and urbanization. Though linguistic conditions predict more human capital, financial access, and structural change, they do not provide persuasive within-state variation in EMI and I interpret all associations descriptively. In sum, a high-return educational opportunity need not equalize opportunity when access to it and the resources required to benefit from it are themselves unequal."
geometry: "left=0.85in, right=0.85in, top=0.9in, bottom=0.9in"
header-includes:
  - \usepackage{setspace}\doublespacing
  - \usepackage{mathtools}
  - \usepackage{longtable}
  - \usepackage{booktabs}
  - \usepackage{array}
  - \usepackage{xcolor}
  - \definecolor{gray35}{gray}{0.35}
  - \usepackage{colortbl}
  - \usepackage{threeparttablex}
  - \usepackage{makecell}
  - \usepackage{tabularray}
  - \usepackage{float}
  - \usepackage{threeparttable}
  - \usepackage{etoolbox}
  - \AtBeginEnvironment{CSLReferences}{\setstretch{1}\setlength{\parskip}{0pt}\setlength{\itemsep}{0pt}}
  - \UseTblrLibrary{booktabs}
  - \UseTblrLibrary{siunitx}
  - \usepackage{caption}
  - \usepackage{subcaption}
  - \captionsetup{width=.95\linewidth,justification=centering}
  - \captionsetup[subtable]{labelformat=empty}
link-citations: true
cite-method: citeproc
---

<!--
WORKING PAPER: paper-new.qmd is the manuscript-facing draft built around the
registered final exhibits. paper.qmd remains the legacy compatibility document
until application-sample excerpts are migrated separately.
-->

```{r public-output-table-helper}
#| include: false
source_public_qmd_helpers <- function() {
  candidates <- c(
    "R/output/public_qmd_helpers.R",
    "../R/output/public_qmd_helpers.R",
    "../../R/output/public_qmd_helpers.R"
  )
  hit <- candidates[file.exists(candidates)]
  if (!length(hit)) stop("Could not locate R/output/public_qmd_helpers.R", call. = FALSE)
  source(hit[[1]], local = knitr::knit_global())
}
source_public_qmd_helpers()
initialize_public_qmd_helpers(knitr::knit_global())
```

# Introduction {#sec-intro}

In an Indian economy characterized by rapid GDP growth alongside wage stagnation, profuse informality, and high rates of both youth unemployment and underemployment [@organization2024a], what can a household do to improve the future prospects of their children? The history of economics highlights *education* as one answer [@mincerHumanCapitalEarnings1974; @card1999; @bjorklundEstimatingReturnInvestments2002; @bhandariIncomeDifferentialsReturns2006]. The history of India highlights another: *English*. In a context of extraordinary linguistic diversity [@gurevichDatasetLinguisticConnectivity2025] and thus extraordinary communication and coordination costs [@nakagawa2021a], English has come to serve as a primary domestic lingua franca [@clingingsmithIndustrializationBilingualismIndia2014]. In its capacity as a global lingua franca [@melitzEnglishGlobalLanguage2016], meanwhile, English opens pathways both internationally---e.g., to education at prestigious universities or employment at multinational firms [@ramanathanTeacherKnowledgeBase2014; @bediEnglishLanguageIndia2019, pp. 32-33]---and at home. The domestic expansion of internationally traded services and information technology, in particular, has led to a flurry of jobs whose higher socioeconomic returns [@chamarbagwalaIndustrialDelicensingTrade2011; @smokotin2014a] are largely accessible only to those who can use English [@mankiwPoliticsEconomicsOffshore2006; @shastry2012a].

Empirical study corroborates the notion that English expands economic mobility [@itaniMeaningLanguageSkills2015]. Using nationally representative household data, @azam2013a find that English fluency was, at least up until 2005, associated with a 34% hourly wage premium for men and a 22% premium for women; even limited English proficiency was associated with premia of 13% and 10%, respectively. @refeque2022a find similar differentials extending through 2011--12. Evidence for causality in this association comes from @chakraborty2016a, who interpret West Bengal's 1983--2004 ban on English teaching in public primary schools as a natural experiment where cohorts more *exposed* to the policy earned lower subsequent wages. "Language skills are human capital," after all [@chiswick1995a, p. 4], just as "English-language skills [specifically] are a form of human capital" [@azam2013a, p. 344].

It is thus by combining our two answers---education and English---that we can begin to understand the demand for *English-medium instruction (EMI)*, or for the teaching of all school subjects in English, currently seen in India [@gooptu2023a; @lahoti2019a; @chattopadhayLowFeePrivateSchools2017]. For its potential to expand equality of opportunity, EMI has been touted as a tool of social development [@commission2009a] and described as a route to escape caste [@mathur2013a; @mathew2011a; @vij2020a] alongside the rigid wage, employment, and occupation distributions it enforces across generations [@madheswaranSocialExclusionCaste2016; @munshi2006a].

But it is ultimately households who decide whether or not a child attends EMI, and the determinants of their choice---the costs they face, the benefits they expect, and the forces constraining the choice set in front of them---are affected by factors which are not just national and historical, but also specific and local. My focus is on whether such factors are associated with the same inequalities that English skills are often expected to help households escape. I study inequalities which exist along dimensions such as social group, religion, parental education, household resources, and urban or rural residence. If the ability to access or benefit from EMI is systematically related to these characteristics, even within the same local education market, then we have evidence that a high economic payoff to English skills alone is not enough for EMI to become an equalizer of opportunity.

Though I never estimate a structural model of household choice, their investment problem still provides a helpful organizing framework. I will first study a type of cost whose variation is predetermined relative to the institutional and economic factors that otherwise affect the household's problem: **inherited linguistic costs to learning English**. These will reflect historically persistent features of the local linguistic environment [see @roweLanguageDevelopmentContext2020] which affect the opportunity cost of learning English over other domestic bridge languages, namely Hindi. This work, detailed in @sec-language, owes much credit to @shastry2012a; in line with her methodology, I create several different measures of local linguistic distance from *Hindi* to capture this opportunity cost. In @sec-access I analyze the **accessibility of English-medium instruction** through enrollment, several measures of EMI exposure, and school management. Private schooling is a distinct household choice with its own costs and selection patterns, so I do not treat private management as a measure of instructional quality [@endow2019a; @muralidharan2015a]. Finally, in @sec-conversion I study potential determinants of **expected returns to English skills**. My goal here is to find evidence on the value of English skills in the local labor market, with the assumption that households incorporate the local history of such returns into their beliefs regarding English, akin to households having adaptive expectations. I ask whether EMI exposure predicts later district consumption and whether that association is stronger where pre-existing human capital and urbanization are greater. I then examine whether inherited linguistic conditions accompany human-capital growth and skilled in-migration, as well as changes in financial access and economic structure.

The local linguistic environment, the local education market, and the local labor market---all three of the factors I focus on are studied locally. These choices are justifiable: the linguistic conditions which households pass down to their children are highly persistent at the local level, local schooling conditions constrain the menu of educational opportunities households choose from, and local economic conditions plausibly cause households to update their beliefs on the economic returns to learning English. I will use a household's district to proxy for their 'local' environment.^[A district in India is a political unit akin to a county in the United States, albeit with an average population of around 2 million. Table \ref{tbl-paper-core-summary} contains summary statistics derived from the distribution of district populations.] Though the district is the most granular unit at which I can harmonize the longitudinal data sources I rely on, it must be noted that interdistrict migration would confound and attenuate any association between the characteristics of one district's education market and its future economic outcomes. Older studies found interdistrict migration rates (and particularly rates of economically-motivated interdistrict migration) to be so low even after the expansion of IT and services in the 1990s [@topalovaTradeLiberalizationPoverty2005; @office2021a] that they "assume[d] that district of current residence (or of employment) of an individual is approximately the same as the schooling district" [@chakraborty2016a, p. 6]. More recent work, however, complicates [@mahapatroInternalMigrationEmerging2020; @rajanCOVID19PandemicInternal2020] and even outright repudiates [@lucasCountrySpecificMagnitudesMigration2021] this notion. These studies emphasize the role of the IT expansion in driving interdistrict migration [@divison2015a; @ghoseTradeInternalMigration2026], and my own results associate potential determinants of household investment into EMI with the skill composition of recent work migrants. I return to those results in @sec-app-migration. Regardless, it is evident that district outcomes are not an aggregation of individual returns to EMI, but a function of places and their changing populations.

The results show that the conditions surrounding English-skill investment are themselves unequally distributed. The relative cost of acquiring a bridge language is partly inherited, and language behavior varies with these inherited linguistic conditions. Formal English-medium schooling is strongly organized by state boundaries, particularly in the public sector. Within the same district, social-group differences remain large in English-medium and private-school participation, with larger differences in urban schooling markets. EMI exposure is more strongly associated with later local consumption than enrollment alone, and this association is stronger in districts with higher initial human capital and urbanization. The inherited linguistic gradient is also associated with later changes in human capital, the skill composition of recent work migrants, financial access, and sectoral composition. These broader associations are selective rather than evidence of general employment expansion.

Taken together, the evidence limits the equalizing promise of a high-return educational opportunity. A child may face high future payoffs to English skills while the household inherits a less favorable linguistic starting point, has less access to EMI, or lives in a district with fewer complementary resources through which English skills can be converted into economic opportunity. English skills can be valuable while access to English-medium schooling remains unequal. A high-return educational opportunity does not automatically equalize opportunity when entry into it, and the resources needed to benefit from it, are themselves unequally distributed.

# Data and Variables {#sec-data}

I construct the district datasets used in the paper by reconciling changing district boundaries, reconstructing archived schooling records, linking Census mother tongues to several linguistic classifications, and harmonizing household consumption surveys across major changes in survey design. I place survey, Census, and school records from different boundary regimes onto Census-2001 district geography using a reviewed district-lineage system informed by official identifiers, district histories, and published concordances [@kumarCreatingLongPanels2016; @ministryPanchayatiRajLocalGovernmentDirectory2026; @asherDevelopmentResearch2021]. I reconstruct district-level EMI and school-system measures from archived DISE workbooks and report cards whose formats and medium-of-instruction fields vary across releases [@mehtaUDISEPublications2005062022; @diseDistrictReportCardsArchive2017]. I combine Census mother-tongue populations with Shastry's Hindi-distance classification, reviewed Census-language crosswalks, Glottolog genealogy, and Dyen lexical data [@shastry2012a; @officeRegistrarLanguageAtlasIndia2004; @hammarstromGlottologGlottologGlottolog2026; @dyenIndoeuropeanClassification1992; @dyenComparativeIndoeuropeanDatabase1997]. I construct comparable district consumption measures from nine NSS/HCES survey designs spanning 2000--01 through 2023--24, placing households on Census-2001 district geography and expressing consumption in 2011--12 prices. Five survey designs provide the baseline and later outcomes used in the main longitudinal analysis [@nssHouseholdConsumerExpenditure2004; @nssHouseholdConsumerExpenditure2009Type2; @nssHouseholdConsumerExpenditure2011Type2; @nsoHCES2022; @nsoHCES2023]. Construction details appear in @sec-app-data, including the additional Census and Economic Census panels used for local-development outcomes. Identification checks are reported in @sec-app-identification.

**Geography and coverage.** District boundaries change often enough that a district name cannot serve as a stable longitudinal identifier. I use Census-2001 districts as the reference geography and reconcile later and earlier units using reviewed district-lineage evidence. The lineage system draws on Census materials and independent district-harmonization resources, including the India State and District Evolution Database, the Indian Census Data Collection, the India District Changes Tracker, and SHRUG [@joladIndiaStateDistrict2026; @joladIndianCensusData2026; @jaacksIndiaDistrictChanges2020; @asherDevelopmentResearch2021]. The reference geography contains 593 Census-2001 districts. The 2007--08 schooling survey reaches 587 of them under the reviewed lineage, while the primary 2017--18 lineage reaches 578; requiring support in both schooling rounds yields a 573-district panel. Because the household surveys are repeated cross-sections rather than repeated observations of the same respondents, I aggregate the relevant measures to harmonized districts.

The principal schooling measures come from the 2007--08 National Sample Survey (NSS) 64th Round, *Participation and Expenditure in Education* [@nationalsamplesurveyoffice2008]. I use survey weights when forming district aggregates. DISE provides a separate school-reported measure of medium of instruction, management, enrollment, and school characteristics. I reconstruct the DISE district series from report cards covering 2005--06 through 2007--08 that were preserved online after the original publication pages disappeared [@mehtaUDISEPublications2005062022; @diseDistrictReportCardsArchive2017]. Census-2001 language tables provide mother-tongue populations and language behavior [@officeoftheregistrargeneralandcensuscommissionerofindiaCensusIndia20012001]. Later sections also use Census-2001 and Census-2011 migration, household, housing, and worker tables. I use the 2005 and 2013--14 Economic Censuses for changes in local economic structure [@centralstatisticsofficeIndiaFifthEconomic2008; @office2019a]. Labor outcomes come from the 2009--10 NSS employment survey and the 2017--18 Periodic Labour Force Survey [@nationalsamplesurveyofficeIndiaEmploymentUnemployment2010; @nationalstatisticalofficeINDIAPeriodicLabour2019].

**Access to English-medium instruction.** The NSS schooling variables separate school participation from medium of instruction. Let $N_d$ be the survey-weighted number of children ages 5--19 in district $d$, $E_d$ the number enrolled, $K_d$ the number enrolled for whom medium of instruction is observed, and $M_d$ the number in $K_d$ enrolled in English-medium instruction (EMI). I call

$$
\mathrm{Enrollment}_d = 100\frac{E_d}{N_d}
$$ {#eq-enrollment}

the *enrollment rate*. This is the extensive margin: whether a child is enrolled at all. Among enrolled children whose medium of instruction is observed, I call

$$
\mathrm{EnglishMediumShare}_d = 100\frac{M_d}{K_d}
$$ {#eq-emi-enrolled}

the *English-medium share among enrolled children*. This is the conditional medium-of-instruction margin among children with a reported medium. Finally,

$$
\mathrm{EnglishMediumExposure}_d = 100\frac{M_d}{N_d}
$$ {#eq-emie}

is *English-medium exposure among all children*. It incorporates enrollment, whether medium is reported among enrolled children, and EMI conditional on a reported medium. Enrolled children with unreported medium remain in an explicit unclassified category. The all-child exposure measure is therefore lower than the conditional English-medium share whenever some children are unenrolled or some enrolled children have unreported medium. I also distinguish *public-school English-medium exposure* and *private-school English-medium exposure*, each defined as the share of all children observed in English-medium schooling in the relevant management sector. Children whose medium or school management is unreported remain outside those sector-medium cells. *Private-school enrollment* is the share of enrolled children with observed school management who attend private aided or unaided schools. These measures describe observed formal-school enrollment and reported medium of instruction. NSS 64 excludes classes taken from private tutors and therefore does not measure supplemental English instruction [@nationalsamplesurveyoffice2008].

**Inherited linguistic conditions.** The primary language measure begins with Shastry's 0--5 ordinal scale of separation from Hindi, developed with linguist Jay Jasanoff [@shastry2012a]. Shastry uses linguistic distance from Hindi to capture variation in the relative cost of acquiring English rather than Hindi as a lingua franca. For each mapped, non-English mother tongue $l$ in district $d$, let $S_{ld}$ denote its Census-2001 speaker population and $D_l$ its Shastry distance from Hindi. The *speaker-weighted distance from Hindi* is

$$
\overline{D}_d =
\frac{\sum_l S_{ld}D_l\,\mathbf{1}\{D_l>0\}}
     {\sum_l S_{ld}\,\mathbf{1}\{D_l>0\}}.
$$ {#eq-speaker-weighted-distance}

The positive-distance restriction excludes Hindi and Urdu, the zero-distance reference languages, from both the numerator and denominator. The resulting mean describes the linguistic composition of speakers whose mapped mother tongue has positive distance from Hindi. I use *linguistic distance from Hindi* as a shorter name for this conditional speaker-weighted measure in the main text. A second measure, the *share speaking languages distant from Hindi*, is the percentage of all mother-tongue speakers in the district whose mapped language has Shastry distance at least three. The distant-language share uses all mother-tongue speakers in its denominator, so the size of the distant-language population enters the measure directly.

I use several alternatives when testing whether results depend on one linguistic classification. The *top-three-language distance from Hindi* is the speaker-weighted Shastry distance among the three largest mapped mother tongues in a district, corresponding to the older construction used in the original analysis. *Genealogical distance from Hindi* is the unweighted number of ancestry-tree edges separating each language from Hindi in Glottolog [@hammarstromGlottologGlottologGlottolog2026]. *Lexical noncognacy with Hindi* is one hundred minus the percentage of cognate items shared with Hindi in the Dyen lexicostatistical data [@dyenIndoeuropeanClassification1992; @dyenComparativeIndoeuropeanDatabase1997]. I also reconstruct the speaker-weighted and distant-language measures using the 1991 Language Atlas for historical validation [@officeRegistrarLanguageAtlasIndia2004]. The crosswalk and adjudication rules for Census mother tongues are documented in @sec-app-data.

**Language-behavior controls.** The Census language-acquisition regressions use state-by-native-language cells rather than districts, so their controls differ from the district covariates used elsewhere. Within each state I condition on the native language's share of speakers and an indicator for the state's most common native language. The relevant continuous and threshold distance specifications also include an indicator for Hindi/Urdu, the zero-distance reference category. I refer to the first two variables together as *within-state language-prevalence controls*. Their summary statistics appear separately in Table \ref{tbl-paper-core-summary} because their unit of observation is a state-language cell rather than a district.

**Economic outcomes and baseline covariates.** I use household consumption as the principal measure of local material living standards. At the household level, monthly per-capita consumption expenditure (MPCE) is consumption expenditure divided by household size. I express MPCE in 2011--12 prices and estimate district means using the survey design and person weights, so the district statistic is the person-level average of real MPCE implied by the survey weights. The baseline comes from the 2004--05 NSS 61st Round [@nssHouseholdConsumerExpenditure2004]. The 2009--10 and 2011--12 NSS rounds provide intermediate endpoints [@nssHouseholdConsumerExpenditure2009Type2; @nssHouseholdConsumerExpenditure2011Type2], and HCES 2022--23 and 2023--24 provide the long-run endpoints [@nsoHCES2022; @nsoHCES2023]. I reconstruct each survey's household consumption measure under its documented design and express the resulting MPCE series in 2011--12 prices using official price series, including the Reserve Bank of India and Labour Bureau series [@reserveBankIndiaDBIE; @labourBureauCPIIW]. HCES 2022--23 and 2023--24 divide consumption across three questionnaires administered to each selected household in three consecutive monthly visits [@nsoHCES2022; @nsoHCES2023]. I reconstruct household MPCE from those questionnaire components before district estimation; @sec-app-data gives the details.

The preferred district regressions compare districts within states and adjust for covariates measured in 2001, before the 2007--08 schooling outcomes. I call these the *baseline Census-2001 covariates*. The registered covariates cover six groups: scale and geography, social composition, human capital, demography, economic structure, and basic development. They include log population, urban share, and log population density; Scheduled Caste, Scheduled Tribe, and Muslim shares; secondary-or-higher attainment; the dependency ratio; the agricultural share of workers; and household electricity access [@officeoftheregistrargeneralandcensuscommissionerofindiaCensusIndia20012001; @asherDevelopmentResearch2021]. These covariates adjust the district regressions for observed differences in baseline conditions. In the heterogeneity analysis, baseline secondary-or-higher attainment and urban share enter as moderators and are omitted from the baseline covariate vector.

Table \ref{tbl-paper-core-summary} reports the scale and dispersion of the principal variables and both covariate families. Enrollment averages about 74 percent of children ages 5--19, while English-medium exposure among all children averages about 15 percent. Private-school English-medium exposure averages about 9 percent and public-school English-medium exposure about 6 percent. The speaker-weighted distance from Hindi averages 3.05 Shastry degrees across districts, and the district mean of the share speaking languages at distance three or greater is 42 percent. The table also reports all ten baseline district covariates and, in a separate state-language panel, the variables used in the Census language-behavior regressions. Mean real consumption in the available district samples is about Rs. 928 per month in 2004--05, Rs. 2,152 in 2022--23, and Rs. 2,242 in 2023--24, all expressed in 2011--12 prices.

::: {.landscape}

```{r}
#| label: render-main-core-summary
#| echo: false
#| results: asis

render_public_table("../outputs/tables/main/paper_core_summary.tex", "paper_core_summary")
```

:::

# Inherited Linguistic Conditions {#sec-language}

## Proxies and Persistence of Linguistic Conditions {#sec-language-persistence}

Why measure distance from Hindi? India's linguistic diversity gives languages of wider communication an important role, and both Hindi and English have served that role across groups [@clingingsmithIndustrializationBilingualismIndia2014; @shastry2012a]. Shastry's argument is comparative: linguistic diversity creates incentives to learn English or Hindi, and greater distance between a mother tongue and Hindi raises the cost of learning Hindi relative to English [@shastry2012a]. The speaker-weighted measure defined in @sec-data summarizes a persistent linguistic component of this relative cost.

Most district variation in the measure occurs between states. In the 573-district schooling panel used later in the paper, state indicators alone explain 84.4% of its cross-district variation. The preferred district specifications therefore compare districts within states and adjust for the baseline Census-2001 covariates. @fig-linguistic-geography shows the observed measure and the residual after removing state indicators and the linear contribution of those covariates.

::: {#fig-linguistic-geography layout-ncol=2}
![Raw linguistic distance.](../outputs/figures/main/map_linguistic_distance.pdf)

![Residual linguistic distance: state fixed effects + baseline Census-2001 covariates.](../outputs/figures/main/map_residual_linguistic_distance_state_main.pdf)

District linguistic distance before and after the preferred state-level adjustment. Sources: Census of India 2001 language and district characteristics [@officeoftheregistrargeneralandcensuscommissionerofindiaCensusIndia20012001; @asherDevelopmentResearch2021]. `r paper_map_boundary_note()` [@datameetCensus2001Districts; @naturalearthDisputedBoundaries].
:::

I call these conditions *inherited* because the district pattern is highly persistent across the independently reconstructed 1991 and 2001 measures. Among the 91 Census-2001 districts meeting the historical reconstruction requirements in both years, the correlation between the 1991 and 2001 speaker-weighted distance measures is 0.898; weighting districts by their 1991 populations raises the correlation to 0.928. The population-weighted correlation for the share of speakers whose mother tongues are at least three Shastry degrees from Hindi is 0.919. @fig-linguistic-persistence shows the district comparison. These estimates agree with Shastry's finding that district linguistic distance is highly correlated across earlier Census years [@shastry2012a].

![Persistence of speaker-weighted linguistic distance from Hindi from 1991 to 2001. Each point is a Census-2001 district meeting the historical language-reconstruction requirements in both years; the dashed line is the 45-degree line. Sources: Census of India 1991 Language Atlas [@officeRegistrarLanguageAtlasIndia2004] and Census of India 2001 language tables [@officeoftheregistrargeneralandcensuscommissionerofindiaCensusIndia20012001].](../outputs/figures/appendix/appendix_historical_language_persistence.pdf){#fig-linguistic-persistence fig-pos="H" width="82%"}

An independent reconstruction provides a second validation of the 1991 series. Helms and Lim's replication data include a Shastry-derived 1991 district linguistic-distance measure [@helmsReplicationDataGlobalization2025]. Across the 148 districts shared by their extract and my reconstruction, the two measures correlate at 0.989. The agreement shows that the independently assembled series contain nearly identical cross-district variation.

The measure captures inherited linguistic distance from Hindi. Family members, peers, schools, media, and other social settings also shape the language input children receive [@roweLanguageDevelopmentContext2020].

## Linguistic Conditions and Reported Language Choice {#sec-language-behavior}

I next test whether the Shastry relative-cost measure is associated with the subsidiary languages people report. The Census-2001 bilingualism and trilingualism tables record native-language populations and reported subsidiary languages [@officeoftheregistrargeneralandcensuscommissionerofindiaCensusIndia20012001]. For English and Hindi, the outcome is the share of multilingual native speakers who report the language as a first subsidiary language or, where applicable, a second subsidiary language. The multilingual outcome is the share of native speakers who report any subsidiary language. Each regression observation is a state-by-native-language cell, weighted by its native-speaker population. The specifications compare language groups within states and include the within-state language-prevalence controls defined in @sec-data; the registered continuous and threshold specifications also include the Hindi/Urdu reference indicator.

::: {.landscape}

```{r}
#| label: tbl-main-language-behavior
#| echo: false
#| results: asis

render_public_table("../outputs/tables/main/paper_language_behavior.tex", "paper_language_behavior")
```

:::

The continuous English estimate is 2.23 percentage points per Shastry distance degree, with a standard error of 1.45 ($p=0.123$). The threshold specification is much larger: speakers of languages at least three degrees from Hindi report English at a rate 16.73 percentage points higher than speakers of the remaining classified mother tongues (SE 4.39; $p<0.001$). This specification parallels Shastry's district measure based on the share of speakers whose languages are at least three degrees from Hindi [@shastry2012a].

The largest continuous coefficient in the table is for reported Hindi. Each additional Shastry degree is associated with 9.69 percentage points less reported Hindi among multilingual speakers (SE 2.57; $p<0.001$). Greater distance is also associated with a 3.58 percentage-point decline in the share of native speakers who report any subsidiary language (SE 0.92; $p<0.001$). The pattern therefore includes both the reported subsidiary language and the prevalence of multilingualism.

The alternative linguistic measures produce different English estimates. Glottolog genealogical distance yields a coefficient of 0.21 percentage points (SE 0.43; $p=0.627$), and Dyen lexical noncognacy yields 0.11 percentage points (SE 0.08; $p=0.167$) in the corresponding continuous specifications [@hammarstromGlottologGlottologGlottolog2026; @dyenIndoeuropeanClassification1992; @dyenComparativeIndoeuropeanDatabase1997]. The Shastry threshold specification yields the large English association.

## Context Dependence of Reported Language Choice {#sec-language-context}

Within the predeclared Hindi-belt sample, the continuous English coefficient is $-0.04$ percentage points per Shastry degree (SE 2.94; $p=0.990$). The threshold estimate is 10.16 percentage points (SE 10.81; $p=0.347$). Reported Hindi remains negatively associated with distance at 9.02 percentage points per degree (SE 5.32; $p=0.091$).

The 1991 Scheduled Tribe language data show a different allocation across reported languages [@officeRegistrarLanguageAtlasIndia2004]. In the validated all-state sample, each Shastry degree is associated with 3.38 percentage points more reported Hindi and 0.11 percentage points less reported English. The corresponding bilingualism estimate is 3.64 percentage points. In the Hindi-belt restriction, the Hindi estimate is 5.69 percentage points per degree. The English estimate is $-0.17$ percentage points, and the bilingualism estimate is 5.82 percentage points. That restriction contains four state clusters, so I use it as descriptive evidence. Clingingsmith's historical analysis provides independent evidence that changing economic incentives increased second-language learning in India [@clingingsmithIndustrializationBilingualismIndia2014].

The sex-specific estimates are similar. In the Census-2001 language data [@officeoftheregistrargeneralandcensuscommissionerofindiaCensusIndia20012001], the continuous English coefficients are 2.33 percentage points for men and 2.01 percentage points for women per Shastry distance degree. Geographic and historical comparisons produce much larger differences in the relationship between linguistic distance and reported language.

Across these samples, linguistic distance from Hindi captures persistent variation in the relative cost of learning Hindi and English. The reported subsidiary language associated with greater distance differs across settings. In the national 2001 estimates, greater distance is associated with less Hindi, while the threshold specification shows more English. The English relationship is weak within the Hindi belt. In the 1991 Scheduled Tribe data, greater distance is associated primarily with more Hindi. Reported bridge-language choice therefore varies with setting as well as inherited linguistic conditions. The next section examines formal schooling as one institutional setting in which access to English differs.

# Access to English-Medium Instruction {#sec-access}

## Measuring Access to English-Medium Schooling {#sec-schooling-menu}

Inherited linguistic conditions can change the relative cost of acquiring English, but households can invest only through schooling options they can reach. I therefore use the measures defined in @sec-data to describe realized access to English-medium schooling. The NSS 64th Round records household choices in 2007--08: whether children ages 5--19 are enrolled, the reported medium of instruction among enrolled children, and school management among enrolled children with reported management [@nationalsamplesurveyoffice2008]. These outcomes reflect available schools, household preferences, and household resources. They are not direct measures of school supply.

The schooling measures separate distinct margins of realized access. Enrollment records whether an age-eligible child attends school. EMI among enrolled children records medium of instruction among children already in school, and all-child EMI exposure incorporates both enrollment and reported medium. I also separate public- and private-school EMI exposure and private-school enrollment. School management and medium of instruction identify different schooling choices: households choosing private schools often cite perceived learning quality or school environment more often than English medium itself [@joshiUnderstandingConsumerDemographics2017]. I therefore keep enrollment, EMI among enrolled children, all-child EMI exposure, public EMI, private EMI, and private enrollment separate.

Reported medium of instruction is also distinct from English proficiency and instructional quality. In the rural sample studied by @lahoti2019a, 57% of children described as attending English-medium schools were actually taught in the dominant regional language, while another 18% used English books with explanation in the regional language. I therefore interpret the NSS and DISE measures as classifications of reported school medium rather than direct measures of English acquisition or the amount of English used in class. The NSS also does not observe private tutoring or other supplemental English instruction.

DISE provides a school-reported perspective. Its district aggregates describe the composition of the local school system, including enrollment by medium and management, rather than household survey choices. Enrollment shares still combine available provision with family demand, so I use DISE as a measure of the reported school system rather than a pure supply measure. The agreement between NSS and DISE later in this section is informative because the two datasets measure the same schooling margin from different respondents [@mehtaUDISEPublications2005062022; @diseDistrictReportCardsArchive2017].

@fig-schooling-geography compares the four EMI measures directly. EMI among enrolled children isolates the medium-of-instruction margin among children already in school, while all-child EMI also incorporates enrollment. Splitting all-child exposure by school management shows that public-school EMI is concentrated especially in northern and northeastern districts, whereas private-school EMI is more geographically dispersed. The four panels use the same percentage bins so that differences in both levels and geography can be compared directly.

::: {#fig-schooling-geography layout-ncol=2}
![EMI share among enrolled children.](../outputs/figures/main/map_emi_share_enrolled.pdf)

![EMI exposure among all children.](../outputs/figures/main/map_emi_exposure.pdf)

![Public-school EMI exposure among all children.](../outputs/figures/main/map_public_emi_exposure.pdf)

![Private-school EMI exposure among all children.](../outputs/figures/main/map_private_emi_exposure.pdf)

Geography of English-medium instruction on the common Census-2001 district geography. All panels use a common percentage scale. Source: NSS 64th Round, 2007--08 [@nationalsamplesurveyoffice2008]. `r paper_map_boundary_note()` [@datameetCensus2001Districts; @naturalearthDisputedBoundaries].
:::

## EMI Access as a Function of Institutional Factors {#sec-schooling-market}

The schooling measures are organized very differently across states. Table \ref{tbl-paper-schooling-market-geography} presents the same three geographic specifications for each schooling outcome: an unadjusted association, a specification with region fixed effects and predetermined Census-2001 controls, and a specification with state fixed effects and the same controls. The table reports standardized coefficients so that attenuation can be compared across outcomes, with state-clustered standard errors beneath each estimate. Nationally, linguistic distance is positively associated with every English-medium schooling margin in the table. The association attenuates sharply within states: the state-adjusted standardized coefficient is about 0.12 for enrollment, 0.03 for EMI among enrolled children, 0.06 for all-child EMI, 0.05 for public EMI, 0.09 for private EMI, and 0.05 for DISE English-medium enrollment.

The state-membership $R^2$ rows ask a separate descriptive question: how much district variation is associated with state membership? State indicators explain roughly 84% of linguistic-distance variation. Across the schooling measures, they explain about 83--84% of all-child and conditional EMI, about 83% of DISE English-medium enrollment, and nearly 90% of public EMI, but only about 57% of private EMI and 36% of enrollment. These statistics locate the variation; they do not identify a causal effect of state institutions. Still, the public/private contrast is consistent with the institutional importance of state language policy. Indian state boundaries and schooling systems have historically been organized around language, and @jain2017a shows that aligning official instructional language with local language after linguistic state reorganization substantially improved educational attainment.

```{r}
#| label: render-main-schooling-market-geography
#| echo: false
#| results: asis

render_public_table("../outputs/tables/main/paper_schooling_market_geography.tex", "paper_schooling_market_geography")
```

The NSS pattern is not an idiosyncrasy of household reporting. DISE English-medium enrollment and NSS EMI among enrolled children have a raw district correlation of 0.896, while DISE and all-child NSS EMI exposure correlate at 0.882. After I subtract each state's mean from both measures, those correlations fall to 0.607 and 0.562, respectively. The decline is informative. A raw district correlation combines agreement across states with agreement among districts within the same state; state residualization removes the between-state component and asks only whether districts that are high relative to their own state are also high in the other source. Because the schooling geography above is strongly organized along state boundaries, removing that common state component should reduce the correlations if part of the cross-source agreement reflects broad state-level schooling systems. The remaining within-state correlations are still substantial, which makes simple NSS measurement error an implausible explanation for the weak within-state relationship between linguistic distance and EMI. The corresponding district-level NSS--DISE comparison is shown in @fig-nss-dise-agreement.

Private schooling has a distinct household cost. Even schools described as low-fee can absorb a substantial share of poor households' income [@endow2019a], and private enrollment is strongly selected on household characteristics [@muralidharan2008a; @rajuPrivateSchoolingIndia2023]. Raw public-private achievement differences therefore do not identify differences in school effectiveness. In a randomized voucher experiment in Andhra Pradesh, @muralidharan2015a find no statistically significant difference between lottery winners and losers in Telugu, mathematics, English, or science and social studies after two and four years, despite large cross-sectional differences between public- and private-school students.

Private management is not a measure of instructional quality. The project's DISE indicators of pupil-teacher ratios, single-teacher schools, sanitation, private-sector composition, and EMI do not form a single quality dimension. I therefore use management to distinguish the schooling options households attend, without ranking public and private schools by quality.

## EMI Access as a Function of Social Factors {#sec-social-access}

The existence of options within a district does not imply equal access to them. @fig-unequal-schooling-access compares Scheduled Tribe (ST), Scheduled Caste (SC), and Other Backward Class (OBC) children with children in the residual Other category within common districts. The social-group differences are present for enrollment but substantially larger for English-medium and private schooling. Relative to Other children, mean district enrollment is lower by about 13.6 percentage points for ST children, 9.8 points for SC children, and 6.4 points for OBC children. EMI among enrolled children is lower by about 13.3, 17.2, and 13.2 points respectively. Private enrollment is lower by about 18.1 points for ST children, 23.5 for SC children, and 14.5 for OBC children. Household resources and social institutions can shape these schooling choices [@endow2019a; @rajuPrivateSchoolingIndia2023]. In Mumbai, @munshi2006a show that caste networks continued to direct lower-caste boys toward local-language schools associated with traditional occupations as returns to nontraditional white-collar work rose, while lower-caste girls shifted more rapidly toward English schools.

![Social-group differences in access to English-medium and private schooling. Panel A reports mean within-district percentage-point differences for OBC, SC, and ST children relative to the Other category in enrollment, EMI among enrolled children, private enrollment, and all-child private EMI exposure. Panel B separates rural and urban differences for EMI among enrolled children and private EMI exposure. Source: NSS 64th Round, 2007--08 [@nationalsamplesurveyoffice2008].](../outputs/figures/main/paper_unequal_schooling_access.pdf){#fig-unequal-schooling-access fig-pos="H" width="100%"}

The rural/urban comparison makes the social stratification more pronounced. For SC children, for example, the difference in EMI among enrolled children is about 8.4 percentage points in rural districts and 24.6 points in urban districts; the corresponding difference in private EMI exposure is about 7.5 and 21.4 points. OBC and ST children show the same qualitative pattern. A plausible interpretation is that more differentiated urban schooling markets offer a wider range of formal options while household resources and social position still condition which options are reachable. I treat that as a descriptive sorting pattern rather than evidence that urbanization itself causes the larger differences.

The corresponding decompositions for boys and girls are directionally similar. Government programs operating before the NSS round explicitly targeted girls' education [@pressinformationbureauInitiativesHumanResource2008], so I also examine the sex split directly. In this national cross-section, social group and rural/urban setting distinguish access more strongly than sex.

The result is a constraint on how the schooling menu should be interpreted. A district can contain English-medium and private schools without placing those options within equal reach of every household. The next section asks whether these observed EMI choices are associated with later economic outcomes that might help explain household demand for them.

# Local Economic Outcomes Associated with English-Medium Instruction {#sec-conversion}

## Later Consumption and English-Medium Instruction

English-language skills are a form of human capital [@chiswick1995a; @azam2013a], and evidence from India gives households an economic reason to value them. Using nationally representative 2005 household data, @azam2013a find English fluency associated with a 34% hourly wage premium for men and a 22% premium for women; workers with a little English receive smaller positive premia, and the returns are greater for more educated workers. @refeque2022a later documents substantial English-skill earnings associations through 2011--12. The strongest evidence for a causal wage effect comes from @chakraborty2016a, who exploit West Bengal's withdrawal of English from primary schools and estimate substantially lower later wages for cohorts more exposed to the policy.

Households making schooling decisions in 2007--08 could observe contemporary labor-market rewards to English, including the wage differences documented in this literature. I use later district consumption to describe the economic outcomes eventually associated with the schooling choices observed in 2007--08. These later outcomes do not measure parents' expectations at the time of enrollment.

Later district consumption varies substantially across India. @fig-welfare-geography shows real consumption per person in 2022--23 and the long change from 2004--05 to 2022--23 among districts satisfying the preferred support criteria for both rounds. The figure shows the raw geographic distribution. The regressions below condition on state fixed effects and baseline Census-2001 covariates.

::: {#fig-welfare-geography layout-ncol=2}
![Real consumption per person, 2022--23.](../outputs/figures/main/map_paper_real_consumption_2022_23.pdf)

![Log change in real consumption per person, 2004--05 to 2022--23.](../outputs/figures/main/map_paper_real_consumption_change.pdf)

Later district consumption and long-run consumption growth on common preferred support. Real consumption is expressed in 2011--12 rupees. Sources: NSS 61st Round, 2004--05, and HCES 2022--23 [@nssHouseholdConsumerExpenditure2004; @nsoHCES2022]. `r paper_map_boundary_note()` [@datameetCensus2001Districts; @naturalearthDisputedBoundaries].
:::

Table \ref{tbl-paper-economic-conversion} compares five schooling margins on common district support. Each entry comes from a separate district regression with state fixed effects and baseline Census-2001 covariates. The table places the coefficients on the same scale for comparison. In the 2022--23 ANCOVA specification, a 10 percentage-point increase in enrollment is associated with 0.7% higher real mean consumption (SE 1.70). The corresponding estimates are 3.9% for EMI among enrolled children, 5.1% for all-child EMI, 4.1% for public EMI, and 5.3% for private EMI. The EMI-among-enrolled, all-child EMI, and private-EMI estimates survive the predeclared family correction. In the 2004--05 to 2022--23 long-change specification, all-child EMI is associated with 3.9% higher real mean consumption and private EMI with 5.4%; both survive family correction. The analogous 2023--24 estimates are smaller, and none survives the predeclared family correction.

```{r}
#| label: render-main-economic-conversion
#| echo: false
#| results: asis

render_public_table("../outputs/tables/main/paper_economic_conversion.tex", "paper_economic_conversion")
```

These are conditional district associations. English-medium and private schooling are socially selected, so the coefficients cannot be interpreted as individual returns to attending those schools. The district-consumption coefficients and individual wage returns are different estimands. This distinction is especially important for private schooling: in @muralidharan2015a's voucher experiment, randomized school access yielded similar scores in Telugu, math, English, and science/social studies despite large cross-sectional public--private achievement differences. For the demand question, the relevant pattern is that EMI margins are much more strongly associated with later consumption than enrollment alone. This pattern is also consistent with @azam2013a's evidence that English skills and education are complementary in the labor market.

## Predetermined Complements to English-Medium Instruction

The same Table \ref{tbl-paper-economic-conversion} asks whether the 2004--05 to 2022--23 association varies with three characteristics measured before the schooling round: baseline human capital, urbanization, and ST concentration. The interaction rows standardize each characteristic so that they report how much the schooling association differs with a one-standard-deviation increase. For all-child EMI, the schooling--consumption association is about 1.5 percentage points larger for a one-standard-deviation increase in baseline human capital and about 1.8 points larger for the same increase in urbanization. Both interactions survive the predeclared family correction. The ST-concentration interaction is near zero. The private-EMI interaction estimates are also positive: 1.47 points for baseline human capital and 2.00 for urbanization. They do not survive family correction.

I use *complement* descriptively for a baseline characteristic associated with a larger schooling--consumption slope. Baseline human capital and urbanization fit that description for all-child EMI. The predeclared IT-employment interaction provides no comparable evidence for either all-child EMI or linguistic distance.

The EMI margins most strongly associated with later district consumption are also the schooling margins that are sharply stratified by social group. The all-child EMI association with consumption growth is larger where baseline human capital and urbanization are greater. The inequality argument therefore concerns both access to EMI and variation in the district characteristics associated with a stronger consumption relationship.

## Economic Outcomes as a Function of Linguistic Conditions {#sec-development}

The final part of the local-outcomes analysis examines whether inherited linguistic conditions are associated with later economic change beyond their weak within-state relationship with formal EMI.

The consumption association appears in the long horizon. @fig-consumption-dynamics follows the coefficient on linguistic distance across the registered later-consumption endpoints. The 2009--10 and 2011--12 estimates are small relative to their uncertainty. The 2004--05 to 2022--23 long-change estimate is positive and precisely estimated, and the 2023--24 estimate remains positive with greater uncertainty. The local-development outcomes below help characterize which economic margins accompany this later consumption association. The timing alone does not identify a mechanism.

![Association between linguistic distance and real mean household consumption across later horizons. Points show ANCOVA and long-change coefficients with 95% state-clustered Wald confidence intervals; all specifications include state fixed effects and predetermined Census-2001 controls. Sources: Census of India 2001; NSS 61st, 66th, and 68th Rounds; HCES 2022--23 and 2023--24 [@officeoftheregistrargeneralandcensuscommissionerofindiaCensusIndia20012001; @nssHouseholdConsumerExpenditure2004; @nssHouseholdConsumerExpenditure2009Type2; @nssHouseholdConsumerExpenditure2011Type2; @nsoHCES2022; @nsoHCES2023].](../outputs/figures/main/consumption_iv_dynamics.pdf){#fig-consumption-dynamics fig-pos="H" width="90%"}

Table \ref{tbl-paper-local-development} reports the predeclared local-development outcomes spanning human capital, migration, household financial access, sectoral composition, and employment. Greater linguistic distance is associated with deeper household literacy and greater graduate access. It is also associated with a higher skilled share among recent work in-migrants and larger gains in household banking access. Employment composition shifts toward services and away from manufacturing. Household television ownership also increases. The migration outcome measures the skill composition of recent work in-migrants at their destination. The interstate share among migrants is close to zero in the preferred specification.

```{r}
#| label: render-main-local-development
#| echo: false
#| results: asis

render_public_table("../outputs/tables/main/paper_local_development.tex", "paper_local_development")
```

The local-development pattern is concentrated in human capital, skilled in-migration, financial access, and sectoral composition. Nonfarm employment, labor-force participation, and the employment rate show no corresponding broad increase. The strongest welfare evidence is for real mean MPCE; the alternative welfare measures do not show a common pattern. The Economic Census results are also compositional. Greater linguistic distance is associated with a larger increase in the services employment share and a larger decline in the manufacturing share. Total nonfarm employment shows no comparable association.^[The sector-composition results are strongest with region fixed effects: the raw state-clustered $p$-values are 0.002 for services and 0.004 for manufacturing. Under state fixed effects the corresponding raw $p$-values are 0.031 and 0.027, and both adjust to about 0.110 under the predeclared family correction.]

The sectoral result is consistent with the service- and skill-intensive pattern documented in India's broader development literature. @kochharIndiasPatternsDevelopment2006 emphasize the country's service- and skill-intensive pattern relative to the manufacturing-led path followed by several other fast-growing Asian economies. More recent work stresses how tradeable services can become increasingly important to development as technology changes the tradability of service work [@baldwinGloboticsDevelopmentWhen2023]. These descriptive estimates associate inherited linguistic conditions with human-capital deepening, skilled in-migration, financial access, and sectoral reallocation. They do not identify a causal effect of linguistic distance on structural transformation.

The Census household and housing outcomes cover 2001--11, while the Economic Census changes cover 2005--13; both windows overlap the 2007--08 schooling measurement. These outcomes therefore cannot be ordered as downstream mediators of measured EMI. I treat them as contemporaneous or co-evolving local conditions. The results identify two district characteristics associated with a stronger schooling--consumption relationship: pre-existing human capital and urbanization. Inherited linguistic conditions are also associated with human capital, skilled in-migration, financial access, and sectoral composition. Together these results show substantial variation in the local economic conditions surrounding EMI.

# Conclusion {#sec-discussion}

English skills are often presented as a way for households to loosen the connection between children's prospects and the circumstances into which they are born [@itaniMeaningLanguageSkills2015; @bediEnglishLanguageIndia2019]. The premise is economically plausible: English proficiency carries substantial labor-market rewards in India [@azam2013a; @refeque2022a], and English schooling has been associated with access to occupations outside traditional caste-linked networks [@munshi2006a]. I began from a narrower question. If English-medium schooling is to expand access to valuable English skills, what conditions surround a household's access to it, and are those conditions themselves related to inherited and social inequality?

The evidence points to inequality in each of the three conditions studied in this paper. The linguistic conditions I call inherited are highly persistent across the 1991 and 2001 reconstructions. Their association with reported English and Hindi differs substantially across settings. State schooling systems account for most of the geographic variation in EMI exposure and provision: state membership explains about 84% of the variation in EMI among enrolled children and about 90% of government-school EMI exposure. Within the same districts, realized access to English-medium and private schooling differs sharply across social groups.

The later consumption results show a corresponding economic pattern. In the 2022--23 ANCOVA comparison, a 10 percentage-point increase in all-child EMI is associated with 5.1% higher real mean MPCE, compared with 0.7% for enrollment. The EMI-among-enrolled, all-child EMI, and private-EMI estimates survive the predeclared family correction. In the long-change specification, the all-child and private-EMI estimates also survive that correction. The schooling--consumption association is larger in districts with greater baseline human capital and urbanization. Greater linguistic distance is also associated with deeper human capital, a more skilled composition of recent work in-migration, greater banking access, and a shift in employment toward services. Nonfarm employment, labor-force participation, and the employment rate show no corresponding broad increase. India's service- and skill-intensive development path provides context for the sectoral pattern [@kochharIndiasPatternsDevelopment2006; @baldwinGloboticsDevelopmentWhen2023]. These estimates describe conditional district relationships. They do not identify individual returns or a causal sequence from linguistic conditions through schooling to later development.

The identification analysis reaches a clear limit. Under the preferred state-adjusted specification, linguistic distance has a partial $R^2$ of about 0.004 for all-child EMI, and the registered IV variants fail the effective-$F$ strength criterion. Alternative linguistic-distance measures also remain weak after state adjustment. I therefore cannot use linguistic distance to identify the causal effect of English-medium instruction. This result also reinforces the substantive finding that state schooling systems account for most of the observed geographic variation in EMI.

English skills carry economic returns. The language used to teach school subjects also affects learning. Evidence from India documents educational costs when students are taught through an unfamiliar language [@jain2017a; @nair2015a; @mohanty2022a]. This paper does not estimate the effect of EMI on learning or follow the same children from schooling into later economic outcomes. Migration also changes the population represented by later district outcomes.

Households therefore begin investment in English skills from unequal positions. Children face different inherited linguistic conditions across districts, and realized access to EMI differs sharply by social group within the same districts. The later consumption relationship is also larger where baseline human capital and urbanization are higher. A high-return educational opportunity can coexist with substantial inequality when inherited linguistic conditions, realized access to EMI, and the local conditions associated with its later economic returns are themselves unequally distributed.

# Technical Note for Replication {#sec-replication}

All reported tables and figures are generated through the project's `{targets}` build. Exact R package versions are recorded in `renv.lock`. Required raw files are declared in `data/metadata/file_manifest.csv` with expected paths and sizes; the manifest also records SHA-256 identities for verified inputs. `DATA_AVAILABILITY.md` records availability and redistribution constraints for each required dataset. The appendices document data construction (@sec-app-data), migration (@sec-app-migration), selection (@sec-app-selection), and weak identification (@sec-app-identification). The repository also provides machine-readable results for the larger validation and local-development analyses. `REPLICATION.md` gives the reproduction procedure and system requirements. The full build checks package synchronization and parses the active R and Quarto files before running the test suite. It then renders the manuscript, verifies cross-references and required outputs, and checks reproduction from the tracked processed district inputs.

\newpage

```{=latex}
\appendix
```

# Major Data Contributions {#sec-app-data}

Several of the datasets used in this paper are not directly comparable across time or collection. District boundaries change, survey district codes follow different vintages, archived school records use their own district labels, and consumption surveys span multiple geographic classifications. I therefore treat geographic and measurement harmonization as part of the empirical work rather than as a preliminary name-matching exercise. I focus first on the four constructions central to measurement in the paper: district lineage, DISE, linguistic measures, and consumption. A final subsection records the additional Census and Economic Census harmonizations used for the local-development outcomes.

## District Lineage Harmonization {#sec-app-lineage}

District names are not longitudinal identifiers in India. Between the 1961 and 2001 Censuses the number of districts rose from 339 to 593, through renamings, partitions, mergers, and boundary changes [@kumarCreatingLongPanels2016]. The 2011 Census contained 640 districts after another 47 were created during the intervening decade [@censusIndiaAdministrativeAtlas2011]. Later surveys and the 2022--24 Household Consumption Expenditure Surveys therefore cannot simply be joined to the 2001 Census by district name. The difficulty is substantive: if a later district contains territory from several earlier districts, assigning it to whichever earlier district has the most similar name changes the population represented by the observation.

**Reference geography and source evidence.** I use the 593 Census-2001 districts as the reference geography. Historical boundary evidence extends back to 1961, while source-specific links carry later surveys through the 2023--24 consumption round. This choice keeps the linguistic measures and predetermined Census controls on the geography in which they are observed and gives every later source a common target. Kumar and Somanathan's district histories provide the benchmark for pre-2001 partitions and population transfers [@kumarCreatingLongPanels2016]. For later changes, Census records and the Ministry of Panchayati Raj's Local Government Directory (LGD) provide the principal evidence on official district identity, creation, renaming, and code continuity [@ministryPanchayatiRajLocalGovernmentDirectory2026]. The India State and District Evolution Database, the India District Changes Tracker, and published survey concordances provide independent checks on district histories and survey-to-Census links [@joladIndiaStateDistrict2026; @jaacksIndiaDistrictChanges2020; @deshpandeIndianEnigma2024]. SHRUG locality keys provide population-based overlap between 2001 and 2011 places when a later district crosses 2001 boundaries [@asherDevelopmentResearch2021]. These sources serve distinct evidentiary roles rather than being combined by majority rule.

**Establishing district identity.** The current procedure supersedes the fuzzy-join method described in the original draft. I standardize spelling and names to generate possible matches, but string similarity cannot by itself establish district identity. Each survey district must first be tied to a registered Census or LGD unit using its code, an exact or documented renamed identity, or a reviewed source record. Cases requiring judgment enter a review ledger that records the decision and its supporting evidence. Every accepted decision must cite a registered source; in the current build all 1,254 accepted source identities satisfy that requirement. Reorganizations are stored as dated predecessor--successor relationships. A split links one predecessor to several successors, a merger links several predecessors to one successor, and a documented rename or unchanged district preserves identity.

**Assigning source districts to Census 2001.** I trace each accepted source unit to its Census-2001 predecessor. An unchanged district or documented rename receives a unit weight on the corresponding 2001 district. A later district created wholly from one 2001 district can also be assigned to that predecessor when the district history establishes a nested, single-parent relationship. A later district that overlaps more than one 2001 district is excluded from the preferred panel. For sensitivity analysis only, SHRUG locality populations determine the share attributable to each 2001 predecessor; accepted fractional weights must be nonnegative and sum to one within the source district. This follows the population-weighting principle used by @kumarCreatingLongPanels2016 for historical district partitions: aggregation should represent territory and population rather than similarity between district names.

**Preferred and sensitivity samples.** The *conservative* lineage variant admits only direct identities, documented renames, and deterministic one-parent transitions. The *primary* variant additionally admits reviewed single-parent cases supported either by official Census/LGD code continuity or by independent district-history evidence corroborated with SHRUG locality overlap; for the latter cases, at least 99 percent of the source district's population must be represented in the locality crosswalk. The *full reviewed* variant also admits accepted fractional allocations across multiple 2001 predecessors as a sensitivity sample. For the two schooling waves, these rules yield 427 common districts in the conservative panel, 573 in the primary panel used in the main analysis, and 587 in the full reviewed panel. Multi-parent cases remain outside the primary panel even when a fractional sensitivity allocation is available.

**Aggregating district measures.** Counts are summed across source units assigned to a common 2001 district. Rates, shares, and means are reconstructed from pooled numerators and denominators rather than averaged across successor districts. Distributional measures that cannot be recovered from sufficient statistics, such as the consumption Gini, are reconstructed from pooled household records where the underlying microdata permit it. This prevents a small successor district from receiving the same weight as a much larger one merely because both descend from the same reference district.

**Geographic boundaries.** The main geographic files use Census-2001 district boundaries. Where the available 2001 shapefile omits a district whose boundary is demonstrably unchanged, I carry back a 2011 polygon only when the Census administrative atlases establish exact continuity [@censusIndiaAdministrativeAtlas2001; @censusIndiaAdministrativeAtlas2011]; the current build uses this rule for eleven districts. I do not infer missing boundaries by reallocating area between polygons for these repairs. The statistical crosswalk and the geographic display therefore use the same district identities without allowing the availability of a convenient polygon to determine how observations are allocated.

The purpose of these rules is not to maximize the number of matched districts. It is to make the meaning of a district observation stable enough that differences across years reflect changes in measured outcomes rather than undocumented changes in territory. The three lineage variants make the remaining coverage trade-off explicit. The validation appendix reports the corresponding sample coverage and shows that the paper's identification conclusion is not a consequence of choosing the primary rather than the conservative or full reviewed lineage.

## DISE Data Reconstruction {#sec-app-dise}

**Recovering the source files.** The District Information System for Education (DISE) is valuable here because it records schools rather than household reports of schooling. The historical files are also difficult to use directly. The original publication site is no longer maintained, file formats change across years, and the raw enrollment workbooks encode medium of instruction in numbered slots whose language meaning can differ across districts. The surviving publication index maintained by Arun Mehta identifies the district report cards and raw files released by NIEPA/NUEPA and notes that 2005--06 was the first year in which DISE covered the entire country using a common data-capture format [@mehtaUDISEPublications2005062022]. I used that index to locate an Internet Archive snapshot of the former DISE district-report-card site and recovered the corresponding workbooks and published report cards [@diseDistrictReportCardsArchive2017]. The local archive spans 2002--03 through 2015--16; the 2005--06 through 2007--08 files provide the baseline measures used in the paper, while later years are retained for longitudinal checks.

**Reading year-specific files.** I record each academic year in a source registry that specifies the archived raw file, any file contained inside a ZIP archive, the enrollment, school, and teacher sheets to read, the corresponding report-card volumes, and any known source problem. This is necessary because neither the workbook layout nor the file extension reliably identifies where each measure is stored. Each year is therefore read using its registered sheet structure rather than a single layout assumed across all vintages. Earlier 2002--05 files are retained as inventory; the baseline construction begins in 2005--06, when nationwide coverage becomes comparable [@mehtaUDISEPublications2005062022].

**Decoding medium of instruction.** The most consequential reconstruction concerns medium of instruction. In the 2005--08 raw enrollment sheets, up to five medium slots contain grade-specific pupil counts, but the slot number is not itself a language code. I sum the grade cells within each slot and identify the slot's language from the corresponding published district report card. The resulting crosswalk contains 7,211 district-year-slot identities, each tied to a report-card PDF and page. English enrollment is the sum of the slots identified as English; Hindi enrollment is constructed in the same way. I do not turn an unidentified positive slot into zero English enrollment. If the English slot itself can be identified, its observed pupils remain usable; if a positive unidentified slot means that the absence of English cannot be established, the English count is left missing. This rule prevents uncertainty about a report label from being converted mechanically into zero exposure.

**Reconstructing enrollment and school characteristics.** The denominator is reconstructed separately from the medium slots. For the baseline years, I use total enrollment in Classes I--VIII obtained from the grade counts when available, followed by the workbook's direct enrollment total and then the sum of government and private enrollment when the earlier measures are unavailable. Medium-classified enrollment is used only as a consistency check; it is never substituted for total elementary enrollment. English-medium enrollment share is therefore English-medium pupils divided by total elementary enrollment, not the share of pupils whose medium happened to be classified in the source file. I also sum government and private management categories separately to construct private enrollment and private-school shares. School and teacher files supply the corresponding school counts, teacher counts, single-teacher-school counts, and girls'-toilet counts; ratios are computed only after these underlying counts have been assembled.

**School-age population denominators.** For measures scaled by the school-age population rather than by observed enrollment, I use Census counts of children ages 6--13. I first place the 2011 age counts on the Census-2001 district geography using only deterministic district links, then combine the 2001 and 2011 anchors to estimate each district's annual log population growth. Population in an intervening academic year is obtained by log-linear interpolation between those anchors. The English-medium gross enrollment ratio divides English-medium pupils by this interpolated age-6--13 population. Post-2011 values used only in longitudinal checks extend the same growth rate beyond 2011; the paper's 2005--08 baseline measures rely on interpolation rather than extrapolation.

**District harmonization.** District harmonization follows the lineage rules described in @sec-app-lineage, but DISE school counts impose a stricter restriction. A DISE district is carried to the Census-2001 geography only when it has an exact identity or a reviewed one-to-one lineage link with weight one. Fractional allocations across several predecessors are excluded because splitting reported school or pupil counts by population weights would create fractional source counts that were never observed. When several source districts correspond deterministically to one Census-2001 district, I sum the pupil, school, and teacher counts first and recompute shares and ratios from the pooled numerators and denominators.

**Checks against published DISE reports.** Published report cards are used both to decode the raw files and to check selected reconstructed cells. As a direct parser check, three registered 2005--06 medium-of-instruction cells for Kupwara reproduce the published report exactly, and the build fails if those cells cease to match. For later years, published report-card pages provide English and Hindi pupil counts when the workbook no longer supplies stable slot identities. The 2009--10 archive is a special case: the second report-card volume was not preserved, so I use the archived complete *District Elementary Education Report Card: 2009--10* where the missing volume is needed [@nuepaDistrictElementaryEducation2009]. In 2010--11 the raw enrollment workbook contains a documented header/data alignment problem; I therefore use the published current-year district total when it can be matched and leave the denominator unavailable otherwise rather than fall back to the corrupted raw total. These source-specific rules are registered by year instead of being inferred after looking at regression results.

**Independent comparison with NSS.** The independent NSS comparison provides a useful check on the resulting baseline measure. @fig-nss-dise-agreement plots 2007--08 DISE English-medium enrollment against NSS English-medium instruction among enrolled children. The raw district correlation is 0.896. After subtracting state means from both series, the correlation remains 0.607. The decline is expected because both schooling measures have a strong state component, while the remaining association shows that the two independent sources agree materially even within states.

![Independent agreement between NSS and DISE English-medium measures. The left panel compares district levels; the right panel subtracts each source's state mean before comparison. Sources: NSS 64th Round, 2007--08 [@nationalsamplesurveyoffice2008] and DISE District Report Cards, 2007--08 [@diseDistrictReportCardsArchive2017].](../outputs/figures/appendix/appendix_nss_dise_agreement.pdf){#fig-nss-dise-agreement fig-pos="H" width="90%"}

## Linguistic Measures {#sec-app-linguistic-measures}

The linguistic measures are intended to capture one specific economic margin: how costly it is to acquire Hindi, relative to English, as a language of wider communication. They are not measures of English proficiency, school quality, or a child's complete language environment. I follow @shastry2012a in using linguistic distance from Hindi because her argument is comparative. As the distance between one's mother tongue and Hindi increases, I expect the opportunity cost of learning English over Hindi to decrease. The primary district measure therefore begins with language-level distances from Hindi and then combines them with the local distribution of mother-tongue speakers.

**Census speaker population.** The speaker distribution comes from the mutually exclusive mother-tongue rows of Census 2001 Table C-16 [@officeoftheregistrargeneralandcensuscommissionerofindiaCensusIndia20012001]. Census language-group subtotals are not treated as additional speakers. This matters particularly for the Census Hindi group, which contains distinct mother tongues such as Bhojpuri and Rajasthani varieties. I assign distance at the mother-tongue level first; a distinct mother tongue cannot inherit Hindi's zero merely because the Census places it beneath the Hindi group.

**Shastry distance measures.** The preferred language scores use Shastry's ordinal 0--5 classification [@shastry2012a]. Hindi and Urdu are the zero-distance reference. Published Indo-European languages retain Shastry's reported degree, while languages outside Indo-European receive degree five only when their family identity has been reviewed. Native English is treated separately: it is neither a zero-distance Hindi language nor part of the weighted-distance mean. The primary *speaker-weighted distance from Hindi* is the Census-speaker-weighted mean among mapped mother tongues with positive Shastry distance. This differs from the older *top-three-language distance*, which retains only the three largest mapped mother tongues in each district. The primary construction uses the full mapped mother-tongue distribution and therefore avoids making the district measure depend on an arbitrary three-language cutoff. I retain the top-three measure only for comparison with the earlier analysis.

**Language classification review.** Not every Census mother tongue has an unambiguous published Shastry category. I therefore use a reviewed crosswalk rather than assigning degrees from names alone. The published Shastry concordance is applied first. For finer Indo-European mother tongues absent from that concordance, reviewed decisions use historical classification evidence from the *Linguistic Survey of India* [@griersonLinguisticSurveyIndia1903] and a named published Shastry anchor. Lexical comparisons from @koganGenealogicalClassificationNew2017 and ASJP v21 [@sorenwichmannCLDFDatasetDerived2025] are used only when they help distinguish nearby classifications; they are not averaged into the preferred distance. The current frozen review ledger contains both accepted assignments and explicitly unresolved rows. A mother tongue for which the evidence does not identify one defensible degree remains unmapped in the preferred construction rather than being assigned the degree that produces greater statistical relevance.

**Population shares and unresolved languages.** Unmapped speaker mass is retained explicitly. For the preferred weighted mean, only mapped positive-distance speakers enter the numerator and denominator, and the mapped-speaker share is reported separately. For population-share measures, unresolved and native-English speakers remain in the district population denominator. Thus the *share speaking languages distant from Hindi* is the percentage of all mother-tongue speakers whose mapped language is at least three Shastry degrees from Hindi. The five positive-distance shares similarly record the percentage of district speaker mass at each degree from one through five; when those shares are entered jointly, distance zero is the omitted reference category. I also retain low- and high-degree assignments recorded for reviewed ambiguous languages as bounding sensitivity checks, but those alternatives never replace the frozen preferred assignments.

**Historical 1991 measure.** The 1991 reconstruction applies the same conceptual rules to the *Language Atlas of India 1991*, which reports the geographic distribution of 114 classified languages [@officeRegistrarLanguageAtlasIndia2004]. The Atlas is a PDF publication rather than the district-language file used in my regressions. I extract the district-wise language sheets using the text and page coordinates embedded in the PDF. Cells with multiple numeric groups, missing text, or other ambiguities enter a review ledger rather than receiving a guessed value. District identity follows the 1991 Census state and district coding structure rather than name similarity [@censusIndiaCodingManual1991]. I check the extracted district populations against the 1991 Primary Census Abstract before constructing language shares. Each Atlas language is then linked to the same Shastry basis, and accepted district-language counts are checked against the Atlas population totals. Because unresolved cells can alter a weighted mean even when they represent little population, I calculate bounds that assign unresolved nonzero language mass to the lowest and highest feasible Shastry degrees. The preferred historical series requires at least 99 percent accepted speaker coverage and a resulting distance-bound width no greater than 0.5 degree before a district is retained. I then place eligible 1991 districts on Census-2001 geography using the lineage rules in @sec-app-lineage. This historical series is used for persistence and independent reconstruction checks, not as a replacement for the modern measure.

**Genealogical and lexical alternatives.** Two independent alternatives deliberately change what linguistic distance means. The *genealogical distance from Hindi* uses the reviewed Census-to-Glottolog 5.3 crosswalk and counts separation along Glottolog's language-family genealogy [@hammarstromGlottologGlottologGlottolog2026]. Glottolog provides stable language identifiers and an explicit genealogical classification; I do not rescale this edge count onto Shastry's 0--5 scale. The *lexical noncognacy with Hindi* measure instead uses the Dyen--Kruskal--Black Indo-European cognate data [@dyenIndoeuropeanClassification1992; @dyenComparativeIndoeuropeanDatabase1997]. At the language level it is one hundred minus the percentage of cognate items shared with Hindi. Following the convention described by Shastry, reviewed non-Indo-European languages outside the Dyen comparison receive five percent shared words, or a noncognacy value of 95; unresolved Indo-European mother tongues remain unresolved. Kogan and ASJP evidence are not numerically substituted for missing Dyen observations. These alternatives are robustness measures because they rely on different linguistic information, not competing versions from which I select the strongest empirical result.


## Price Series and Consumption Harmonization {#sec-app-consumption-construction}

Comparing district consumption across two decades requires more than placing several survey rounds in one file. The harmonization system registers nine consumption-survey designs from NSS 56 in 2000--01 through HCES 2023--24, including both NSS 66 questionnaire designs. The main longitudinal analysis uses NSS 61 as the baseline, NSS 66 Type 2 and NSS 68 Type 2 as intermediate endpoints, and the two HCES rounds as long-run endpoints; the remaining registered designs provide pretrend, treatment-period, and questionnaire-design checks. These surveys differ in questionnaire design, recall periods, district codes, and price environments [@nssHouseholdConsumerExpenditure2004; @nationalstatisticalofficeIndiaHouseholdConsumer2011; @nssHouseholdConsumerExpenditure2009Type1; @nssHouseholdConsumerExpenditure2009Type2; @nssHouseholdConsumerExpenditure2011Type2; @nsoHCES2022; @nsoHCES2023]. I reconstruct household monthly per-capita consumption expenditure (MPCE) within each survey first, validate that nominal measure against published survey totals, and only then harmonize prices and geography. The outcome used in the paper is real consumption per person, not a nominal rupee comparison across survey years.

**NSS consumption definitions.** For the NSS Schedule 1.0 rounds, I retain the consumption concept reported for that design rather than forcing every round into a nominally identical recall rule. The source metadata identify the household MPCE field, its scale, and the survey-design variables required for weighting; the benchmark metadata separately record whether the published comparison is based on the mixed reference period (MRP) or modified mixed reference period (MMRP). Under MRP, less frequently purchased categories use a 365-day recall while most other spending uses 30 days; MMRP additionally uses a 7-day recall for selected food and related items [@nsoHCESReport2023]. Some releases report MPCE directly; others require reconstruction from the split household files. I check the resulting household series against the official rural and urban MPCE for every registered round before any price adjustment. Across the nine surveys used here, the 18 reconstructed rural/urban means differ from their published counterparts by less than ₹0.50 in every case. A mismatch stops construction of that round rather than being carried into the district estimates.

**HCES three-visit reconstruction.** The modern HCES releases require an additional reconstruction because they replace the single long consumption interview used by earlier NSS rounds with three questionnaires administered in separate visits: the Food Questionnaire (FDQ), Consumables and Services Questionnaire (CSQ), and Durable Goods Questionnaire (DGQ) [@nsoHCES2022; @nsoHCESReport2023]. The questionnaires use reference periods of 7, 30, or 365 days depending on the item. I identify the registered expenditure items in each questionnaire, convert each reported amount to a monthly equivalent using its stated reference period, and then sum the three questionnaire components. Imputed house and garage rent is explicitly excluded from this MPCE reconstruction, consistent with the official published MPCE concept used for the benchmark comparison. The 2023--24 release identifies visit order, so I use the multiplier from the third visit, as prescribed in the official estimation procedure [@nsoHCESReport2023]. The 2022--23 release does not preserve visit order in the same way; its three questionnaire records are therefore required to carry the same multiplier before I use their common value. Households missing one of the three questionnaire components are not treated as complete HCES consumption observations.

**Temporal price adjustment.** Nominal MPCE must then be made comparable across dates. I use a *temporal price relative* to measure price change through time within a state and rural or urban sector. Before 2013, rural temporal prices use the Labour Bureau's Consumer Price Index for Rural Labourers (CPI-RL), whose 1986--87-base series covers 20 states and all-India [@labourBureauCPIRL]. Urban prices use the Labour Bureau's 2001-base Consumer Price Index for Industrial Workers (CPI-IW), aggregated from its centre series to the state level [@labourBureauCPIIW]. From 2013 onward, I use the state rural and urban CPI series with base 2012=100 from the national CPI system [@reserveBankIndiaDBIE]. I link the pre-2013 series to the 2012-base state CPI using their common state-sector months in 2013--14; state-sector cells without sufficient overlap fail rather than receiving an assumed link. A separate overlap between the 2010- and 2012-base rural/urban CPI series places the July 2011--June 2012 reference-year index on that same scale.

**Spatial price adjustment.** I preserve cross-state and rural/urban price-level differences using the state-specific 2011--12 Tendulkar poverty lines [@planningCommissionPovertyEstimates2013]. Each state-sector poverty line is divided by the all-India rural reference of ₹816 to obtain the spatial relative; using a common denominator retains the national rural-urban price difference rather than normalizing it away. Direct state observations are preferred throughout. When a historical index is unavailable for a small state or union territory, the price crosswalk permits only documented donor or all-India substitutions; the source and reason remain attached to the resulting household deflator.

**Household deflation and survey timing.** The household price deflator is therefore

$$
P_{srt}=P^{\text{spatial}}_{sr}\times P^{\text{temporal}}_{srt},
$$

where $s$ denotes state, $r$ rural or urban sector, and $t$ survey time. Real MPCE is nominal MPCE divided by this deflator. For NSS rounds, I average the monthly state-sector indices over the three months represented by the household's survey sub-round. For HCES, the price assignment follows the household's three-month panel. This timing avoids using an annual price index when the fieldwork itself identifies a narrower interview period.

**District aggregation and common support.** I attach each household's survey district to the Census-2001 geography using the lineage system described in @sec-app-lineage. Unresolved or ineligible geography is recorded rather than silently assigned. Person-weight coverage ranges from about 92.6% in 2000--01 to complete coverage in 2007--08, is about 97--98% in 2009--12, and is about 91.6% in each modern HCES round. HCES is designed to report consumption at the national and state or union-territory levels; the district means in this paper are estimates I construct from the household microdata, not official district HCES statistics [@nsoHCESReport2022]. I therefore retain district sample information alongside every estimate. District means are person-weighted: a household's survey weight is multiplied by household size so that the district mean represents consumption per person rather than the mean of equally weighted households. The district welfare system records the number of sampled households and primary sampling units, calculates a Kish effective sample size, and applies predeclared support thresholds before an estimate enters the preferred analysis. Comparisons that require the same districts use explicit common-support samples rather than allowing the set of districts to change unnoticed across outcomes or rounds.

**Cross-round validation.** The two modern HCES rounds provide a useful internal check because they use the same three-visit design in adjacent years. @fig-app-consumption-hces compares district estimates on common preferred support. The correlation is 0.853 for real mean MPCE, 0.852 for mean log real MPCE, and 0.855 for weighted median real MPCE. These correlations do not make the modern HCES directly interchangeable with the older NSS designs; they show that the reconstruction produces similar district rankings when the survey design is held nearly fixed across adjacent rounds. Together with the national MPCE benchmark checks, this is why the paper treats consumption as a harmonized multi-survey measurement system rather than as one nominal outcome carried unchanged across time.

![District-level consistency of reconstructed consumption across the 2022--23 and 2023--24 HCES rounds. Each panel uses districts satisfying the preferred support rule in both rounds; the dashed line is the 45-degree line. Sources: National Sample Survey Office, HCES 2022--23 and HCES 2023--24 [@nsoHCES2022; @nsoHCES2023].](../outputs/figures/appendix/appendix_consumption_hces_consistency.pdf){#fig-app-consumption-hces fig-pos="H" width="90%"}

## Additional Harmonized Outcome Panels {#sec-app-additional-data}

The local-development analysis uses several additional district panels that require the same attention to changing geography and denominators. For the 2001--11 Census comparisons, I construct migration, household human-capital, housing and asset, and worker-industry measures from published district counts [@officeoftheregistrargeneralandcensuscommissionerofindiaCensusIndia20012001; @censusIndia2011]. I carry 2011 counts to Census-2001 districts only where the lineage permits the required aggregation, and I pool numerators and denominators before calculating rates. The resulting measures describe comparable places rather than whichever district label happened to exist in each Census year.

For the Economic Census, I use SHRUG's district products for the 2005 and 2013--14 measures that they publish comparably [@asherDevelopmentResearch2021]. The predetermined 2005 information-technology measure requires finer industry detail, so I read the official Fifth Economic Census establishment records directly [@centralstatisticsofficeIndiaFifthEconomic2008]. Those records identify district, major or subsidiary activity, four-digit NIC-2004 activity, and total workers. I retain major-activity nonagricultural establishments and define computer-related activity as NIC-2004 Division 72. I use that measure only as a 2005 baseline characteristic because the Sixth Economic Census files available here do not permit an exact four-digit NIC-2004 comparison [@office2019a]. Labor-force participation and employment are constructed separately from the 2009--10 NSS employment survey and the 2017--18 PLFS, which measure residents rather than employment located at establishments [@nationalsamplesurveyofficeIndiaEmploymentUnemployment2010; @nationalstatisticalofficeINDIAPeriodicLabour2019].

# Migration {#sec-app-migration}

## Literature Review

Migration changes the interpretation of district outcomes. The schooling measures in this paper describe children living in a district in 2007--08, while later consumption describes the population living there many years afterward. If people relocate, the economic trajectory of a place need not equal the later economic trajectory of the children originally observed there. I therefore interpret the long-run district outcomes as place-based outcomes for changing populations, not as individual returns to the children exposed to a particular schooling environment.

Earlier work suggested that migration might be quantitatively modest for these comparisons. @chakraborty2016a report low interdistrict mobility in West Bengal, Haryana, and Punjab and approximate an individual's district of residence or employment with the district of schooling. Using earlier NSS rounds, @topalovaTradeLiberalizationPoverty2005 similarly documents low decadal interdistrict mobility, especially in rural areas and for employment-related moves. These measures describe recent-duration migration rather than the accumulated stock of everyone who has ever migrated. Lifetime-migrant measures accumulate moves over a longer exposure window as respondents age and can be dominated by moves made long before the economic period under study [@mahapatroInternalMigrationEmerging2020].

I do not assume that mobility remained negligible. Evidence using Census cohorts and railway passenger flows points to substantially more interdistrict movement after 2001 than the late-twentieth-century evidence implied [@divison2015a]. More recent work also shows why movement can respond to skill-intensive opportunity. @ghoseTradeInternalMigration2026 studies India's information-technology expansion in a setting where education and internal migration respond endogenously, so regional welfare gains depend in part on who acquires skills and where workers relocate. At the same time, migration remains structured by institutions. Caste-based rural insurance networks can discourage migration when leaving weakens access to informal insurance [@munshiNetworksMisallocationInsurance2016], while state borders themselves sharply reduce migration between otherwise neighboring districts [@koneInternalBordersMigration2018]. These findings make migration neither negligible nor frictionless.

My migration measures come from the 2011 Census D-series tables [@censusIndia2011]. D-02 describes migrants by place of last residence and provides the interstate share among migrants. D-03 records reason for migration. For work-related sorting I use D-07, which identifies migrants who have lived at the destination for at most nine years, report work or employment as the reason for migration, and previously lived elsewhere in India. I refer to this group as *recent work migrants*. The D-07 total is required to agree exactly with the corresponding D-03 work-migrant subtotal before the measures are used. Counts from the 2011 district geography are then pooled to the Census-2001 geography only where the district lineage is complete and deterministic, as described in @sec-app-lineage.

The distinction between all migrants and recent work migrants is important for the paper's economic question. Overall migrant composition combines moves for marriage, education, household relocation, and work. Restricting attention to recent work migrants instead asks whether destinations differ in the kinds of workers they recently attracted for employment. The result therefore concerns the composition of in-migration observed at the destination; it does not measure the skills of people who left the district.

## Empirical Results

@tbl-app-migration summarizes the preferred within-state reduced forms. Each coefficient compares districts within states and conditions on the predetermined Census controls used elsewhere in the paper. The first panel retains broad migration margins even when their estimates are null. The second focuses on recent work migrants and includes both their geographic origin and skill composition, together with the predeclared Hindi-belt restriction.

```{r}
#| label: tbl-app-migration
#| echo: false
#| results: asis
render_public_table("../outputs/tables/appendix/appendix_migration_summary.tex", "appendix_migration_summary")
```

The broad mobility measures provide little evidence that linguistic distance is associated with simply receiving more geographically distant migrants. The separate 2001--11 changes in population-denominated migrant rates remain descriptive and are not part of this predeclared regression family, so I do not use them to make an inferential claim about total migration incidence here. The interstate share among all migrants is close to zero in the preferred specification. The work- and education-reason shares provide no family-adjusted evidence of an association. Among recent work migrants, the share arriving from outside the state is also near zero. These null results are useful because they distinguish a change in migrant composition from a general increase in longer-distance mobility.

The skilled composition result is different. A one-degree increase in speaker-weighted linguistic distance from Hindi is associated with a 0.0129 increase in the graduate-or-technical-degree share among recent work migrants; the state-clustered standard error is 0.0044 and the Holm-adjusted $p$-value is 0.027. I interpret this as evidence of skill sorting across destinations, not as evidence that linguistic distance causes migration or that migrants mediate an EMI effect. The dependent variable is the skill composition of recent work in-migrants conditional on being observed at the destination.

The predeclared Hindi-belt restriction is weaker. Within the Hindi-belt states, the corresponding estimate is 0.0079 with a standard error of 0.0057 ($p=0.171$). The national within-state association therefore is not clearly reproduced inside that more linguistically similar institutional setting. This boundary is one reason I do not reduce the migration evidence to a universal behavioral response to linguistic distance.

Taken together, the migration results sharpen the interpretation of the later district outcomes. Linguistic conditions are associated with *who arrives for work* more clearly than with broad migration composition or interstate movement. Migration can therefore change the skill composition of places even when aggregate mobility margins show little association. Later district consumption should consequently be read as local development for a changing population, not as the individual economic return to the children observed in the earlier schooling data.

# Education Selection {#sec-app-selection}

The district comparisons in the main text show that schooling access differs sharply by social group even within the same local market. I retain the original child-level enrollment model here as a complementary description of *who is enrolled at all*. The model does not identify a causal effect of any household characteristic or schooling program, and I do not use it as a control function or selection correction in the district analysis.

I use the NSS 64th Round *Participation and Expenditure in Education* survey, restricting the child roster to ages 5--19 [@nationalsamplesurveyoffice2008]. The outcome equals one when the child appears in the survey's enrolled-child education record and zero otherwise. The fitted probit includes age, sex, household size, religion, social group, rural or urban residence, distance to the nearest primary class, and the education of the child's father or the registered male household proxy. It also includes district averages of selected schooling-cost and benefit variables among enrolled children. Those district averages are useful adjustment variables for reproducing the original model, but they are partly determined by school provision and the composition of enrolled children. I therefore keep them in the fitted specification while omitting them from the reader-facing coefficient table.

The final specification uses the NSS survey design: household weights, first-stage sampling units corresponding to sampled villages or urban blocks, and the survey's strata and substrata. Because a probit coefficient is expressed on a latent index, I report quantities on the probability scale. For continuous variables, the table reports the average slope of the predicted enrollment probability. For categorical variables, it reports the average discrete change in predicted enrollment probability when each observed child is assigned the comparison category rather than the stated reference category, with the remaining covariates left at their observed values. These are average marginal quantities over the estimation sample rather than effects evaluated at a synthetic "average child" [@hansenEconometrics2022; @arel-bundockHowInterpretStatistical2024].

```{r}
#| label: tbl-appendix-selection-ame
#| echo: false
#| results: asis
render_public_table("../outputs/tables/appendix/appendix_selection_ame.tex", "appendix_selection_ame")
```

The social-identity results reinforce the main-text evidence that access to education is not evenly distributed. Relative to Hindu children with the same modeled covariates, Muslim children have an average predicted enrollment probability about 12.6 percentage points lower; the corresponding Christian comparison is about 4.3 points lower. The Muslim difference is consistent with evidence of persistent educational disadvantage among Indian Muslims [@borooah2021a]. Relative to children in the residual "Other" social-group category, the predicted enrollment probability is about 10.2 points lower for Scheduled Tribe children, 7.5 points lower for Scheduled Caste children, and 2.9 points lower for Other Backward Class children. I interpret these as conditional associations, not as effects of religion or caste. That caution is especially important in a setting where social institutions and caste networks can alter schooling choices and the occupational opportunities attached to them [@munshi2006a].

Parental education displays the largest systematic gradient. Relative to an illiterate father or male household proxy, the average comparison rises from roughly 9.7 percentage points for a literate adult with no formal schooling to about 14.2 points for primary education, 21.2 points for secondary education, and 24.5 points for postsecondary education. Father's education can proxy for several household characteristics at once, including permanent income, information about schooling and labor markets, and preferences or expectations transmitted across generations. The model cannot separate those channels. What it establishes is that enrollment is strongly stratified by intergenerational human capital even after conditioning on the other observed covariates.

Several other estimates are useful mainly as descriptive boundaries. Enrollment probability declines by about 2.8 percentage points per year of age across the 5--19 range. Girls have a roughly 4.4-point higher predicted probability than boys conditional on the other modeled characteristics; I do not attribute that difference to any particular program, although Sarva Shiksha Abhiyan and girls-focused schooling initiatives were already operating before the survey [@pressinformationbureauInitiativesHumanResource2008]. Each additional household member is associated with a small decline in predicted enrollment probability. That sign is consistent with the resource-dilution logic of quantity--quality models, but Indian evidence on whether family size causally reduces child investment is mixed [@beckerEconomicAnalysisFertility1969; @kumarTestingChildrenQuantityQuality2011; @onurBirthOrderSibling2021].

School access is also unevenly related to enrollment. Children whose nearest primary class is reported one to two kilometers away have a predicted enrollment probability about 7.7 percentage points lower than children with a primary class within one kilometer. The more distant categories are estimated much less precisely and do not form a monotone distance gradient. I therefore treat the one-to-two-kilometer comparison as evidence that local access conditions can matter, not as an estimate of a stable distance-response function.

The original model also includes district averages of free schooling, fee waivers, scholarships, textbooks, stationery, mid-day meals, and enrollment cost. I do not display or interpret those coefficients here. These variables are measured among enrolled children and are jointly determined by program targeting, school type, local provision, and household selection. In particular, the negative coefficient on the district share receiving free education cannot establish that free schooling reduces enrollment. Retaining those variables as controls preserves the original specification without turning selected program receipt into a policy effect.

The child roster contains 127,181 observations, but the probit is estimated on complete cases. Missingness is concentrated in three required covariates: father's education, district-average enrollment cost, and distance to the nearest primary class. The registered missingness checks show that approximately 9.7 percent of children are missing at least one required model covariate. Missingness is also predictable from observed characteristics, especially for district-average enrollment cost and school distance, so the estimation sample should not be treated as a random subsample of all children.

```{r}
#| label: tbl-appendix-selection-missingness
#| echo: false
#| results: asis
render_public_table("../outputs/tables/appendix/appendix_selection_missingness.tex", "appendix_selection_missingness")
```

This limitation is another reason I use the model only as corroborating descriptive evidence. Its central message agrees with the within-district schooling-access analysis in the main text: educational participation is strongly associated with social identity and family background. The district decomposition is more directly suited to the paper's question about access to English-medium schooling; the probit shows that unequal selection begins one step earlier, at entry into schooling itself.

\newpage

# Linguistic Distance as an Instrumental Variable {#sec-app-identification}

The main text uses linguistic distance principally as a measure of inherited linguistic conditions. I separately ask whether it can serve as an instrumental variable for English-medium schooling: that is, whether variation in linguistic distance can isolate variation in English-medium exposure that is sufficiently strong and otherwise excluded from the later consumption equation. The answer is no. The purpose of this appendix is to show where the design fails and why changing the linguistic measure does not repair it. This distinction is important because weak instruments can make conventional two-stage least squares (2SLS) estimates and their standard errors unreliable even when a familiar first-stage threshold appears to be met [@keanePracticalGuideWeak2024].

## First Stage Collapses

The broad national relationship between linguistic distance and English-medium schooling is not itself in doubt. On the common sample used for the registered first-stage comparison, the unadjusted excluded-instrument $F$ statistic is 11.73. Adding region fixed effects and the main predetermined Census-2001 controls reduces it to 3.24; using state fixed effects and the same controls reduces it to 0.75. @fig-first-stage-absorption shows the corresponding partial relationships, residualizing both linguistic distance and all-child EMI under the adjustment shown in each panel. The same attenuation is summarized alongside the alternative linguistic measures in @tbl-app-iv-relevance. The state-fixed-effects partial $R^2$ is 0.0039: after comparing districts within the same state and conditioning on predetermined characteristics, linguistic distance explains very little of the remaining variation in all-child English-medium exposure.

![Partial relationship between linguistic distance and all-child EMI exposure with progressively finer geographic adjustment. Points are equal-frequency bin means on a common sample; lines are OLS first-stage fits. The region and state panels residualize both axes using the corresponding fixed effects and the main predetermined Census-2001 controls. Panel labels report state-clustered HC1 $F$ statistics. Sources: Census of India 2001 and NSS 64th Round, 2007--08 [@officeoftheregistrargeneralandcensuscommissionerofindiaCensusIndia20012001; @nationalsamplesurveyoffice2008; @asherDevelopmentResearch2021].](../outputs/figures/main/poster_first_stage_specs.pdf){#fig-first-stage-absorption fig-pos="H" width="100%"}

The distinction between the conventional first-stage $F$ statistic and the Montiel Olea--Pflueger effective $F$ statistic matters here. The conventional statistic is useful for comparing the same first-stage specification as geographic controls are added. The effective $F$ is designed for assessing weak-instrument relevance when errors may be heteroskedastic or clustered [@montieloleaRobustTestWeak2013]. I therefore use the conventional $F$ statistics to describe geographic absorption and the effective $F$ statistic when evaluating the preferred IV specification. I do not treat a value of 10 as a universal pass/fail rule; contemporary weak-IV guidance shows that important distortions in 2SLS inference can persist well above that threshold [@keanePracticalGuideWeak2024].

The first-stage collapse is substantively informative. Linguistic distance and English-medium schooling are strongly related across broad Indian geography, but most of that relationship is absorbed when districts are compared within states. The residual panel in @fig-linguistic-geography shows where linguistic-distance variation remains under that preferred adjustment, while @fig-first-stage-absorption shows that this remaining variation has little relationship with residual EMI. This fits the institutional evidence in the main text: English-medium schooling, especially in government schools, is strongly organized at the state level. It also means that linguistic distance cannot provide the within-state excluded variation needed to interpret the later schooling--consumption association as a causal effect of English-medium instruction.

## Spatial Dependence

District schooling measures, linguistic conditions, and later outcomes are geographically clustered, so I also test whether the preferred specifications leave positive spatial autocorrelation in their residuals. I construct row-standardized contiguity weights from the Census-2001 district polygons and calculate Moran's $I$, a global measure of whether neighboring districts have similar values [@anselin2001]. Under rook contiguity, English-medium exposure has $I=`r report_value("moran_emi_i")`$, linguistic distance has $I=`r report_value("moran_linguistic_distance_i")`$, real consumption growth has $I=`r report_value("moran_consumption_growth_i")`$, and the change in the Gini coefficient has $I=`r report_value("moran_gini_change_i")`$; each test rejects the null of no positive spatial autocorrelation at $p<0.001$. The geographic concentration visible in the underlying variables is therefore substantial.

For the preferred consumption specification, the second-stage residuals have Moran's $I=`r report_value("moran_iv_residual_i")`$ ($p=`r report_value("moran_iv_residual_p")`$), while the corresponding first-stage residuals have $I=`r report_value("moran_first_stage_residual_i")`$ ($p=`r report_value("moran_first_stage_residual_p")`$). These tests do not reject the null of no positive residual spatial autocorrelation. Replacing rook with queen contiguity leads to the same conclusion: the second-stage residual test has $p=`r report_value("moran_iv_residual_p_queen")`$ and the first-stage residual test has $p=`r report_value("moran_first_stage_residual_p_queen")`$. Three offshore districts---Lakshadweep, South Andaman, and Nicobars---have no contiguous district neighbor. I retain them with zero spatial lag rather than altering the district boundaries or increasing the contiguity tolerance. Taken together, the tests document strong spatial clustering in the underlying variables and no evidence of positive spatial autocorrelation in the preferred first- or second-stage residuals.

## Robustness to Measures and Data

The weak within-state relationship is not specific to one way of measuring linguistic distance. @tbl-app-iv-relevance compares the preferred speaker-weighted measure with the legacy top-three measure, genealogical distance, lexical noncognacy, the population share speaking languages distant from Hindi, a five-component language-distance composition, and the independently reconstructed 1991 distance measure. For each construction, I report the excluded-instrument $F$ statistic without geographic fixed effects, with region fixed effects, and with state fixed effects. The last column reports the state-fixed-effects partial $R^2$.

```{r}
#| label: tbl-app-iv-relevance
#| echo: false
#| results: asis
render_public_table("../outputs/tables/appendix/appendix_iv_relevance_summary.tex", "appendix_iv_relevance_summary")
```

The scalar alternatives all tell the same story. The top-three measure has a raw $F$ of 16.30 but a state-fixed-effects $F$ of 1.96. The genealogical and lexical measures fall from 10.43 and 5.29 to 0.28 and 0.15, respectively, while the distant-language share falls from 4.08 to 0.59. The richer five-component language composition is strongly jointly predictive without geographic fixed effects, but its state-fixed-effects joint $F$ is only 3.73. A richer set of language shares can therefore predict broad geographic differences without supplying strong within-state excluded variation.

Moving the linguistic measure backward in time does not solve the problem either. On the validated common historical support, the 1991 speaker-weighted distance has an $F$ statistic of about 9.98 without geographic fixed effects, 0.31 with region fixed effects and Census controls, and 0.03 with state fixed effects and those controls. The historical construction is valuable because it separates measurement of the linguistic environment from later schooling, but it does not restore relevance. The broader historical checks reach the same boundary: linguistic geography is persistent, yet predeterminedness and persistence do not imply that the exclusion restriction holds or that the first stage is strong. Pretreatment balance and negative-control checks can inform an IV design, but they cannot by themselves establish exclusion, and their interpretation also depends on specification choices [@danieliNegativeControl2026].

Measurement error in English-medium schooling is also an implausible explanation for the collapse. The independently reconstructed DISE measure agrees closely with the NSS measure in levels and continues to agree after state means are removed; @fig-nss-dise-agreement reports correlations of about 0.896 and 0.607, respectively. The language-measure sensitivities likewise permit alternative crosswalks and language classifications, yet none produces a persuasive state-level first stage. These results make the identification boundary robust to the two most immediate measurement objections: changing the schooling source and changing the linguistic-distance construction.

The richer language-share instruments do not offer a credible escape through multiple instruments. In the registered state-fixed-effects specifications, constituent first stages remain weak and the falsification-adaptive sets contain zero. A falsification-adaptive set summarizes parameter values that remain compatible with the model under minimally sufficient relaxations of overidentifying restrictions [@mastenSalvagingFalsified2021]. Here it reinforces the simpler relevance evidence rather than recovering a stable causal estimate.

## Weak-IV-Robust Inference

Weak first stages change how the second stage should be read. Conventional 2SLS estimates remain useful as a description of what the ratio estimator produces, but their usual $t$ tests are not the basis for inference. I instead emphasize the Montiel Olea--Pflueger effective $F$ statistic and Anderson--Rubin (AR) tests. The AR test evaluates candidate values of the structural coefficient without requiring a strong first stage and is therefore appropriate when identification is weak [@keanePracticalGuideWeak2024].

@tbl-app-iv-weak reports the two long-change specifications used in the main identification discussion. I do not attach conventional significance stars to the 2SLS coefficients. The table instead places the weak-IV statistics directly beside them.

```{r}
#| label: tbl-app-iv-weak
#| echo: false
#| results: asis
render_public_table("../outputs/tables/appendix/appendix_iv_weak_inference.tex", "appendix_iv_weak_inference")
```

For 2004--05 to 2022--23, the 2SLS coefficient is 0.117 with a state-clustered standard error of 0.152; the effective $F$ statistic is 0.69. For the 2023--24 endpoint, the corresponding estimate is 0.137 with a standard error of 0.234 and an effective $F$ of 0.41. Neither conventional estimate is informative enough to support a causal interpretation. More importantly, the AR confidence sets are disconnected. On the evaluated grid, the 2022 set accepts values at or below approximately $-0.071$ and at or above $0.025$, while the 2023 set accepts values at or below approximately $-0.051$ and at or above $0.015$. Both sets extend to the grid boundaries. Rejecting zero on such a grid does not identify the sign of the effect: both negative and positive values remain compatible with the weak-IV-robust test.

The result is not a peculiarity of these two specifications. Across every registered consumption-IV robustness family, no specification satisfies the project's predeclared effective-$F$ strength criterion. The families vary the linguistic measure, English-medium treatment, welfare outcome, controls, parameterization, and historical adjustment. I retain the full model-level results in the replication outputs rather than printing the hundreds of specifications here. Their collective implication is narrower and more useful: there is no nearby specification choice that turns the design into a strongly identified IV strategy.

Finally, I relax exact exclusion rather than treating it as guaranteed. Following the logic of plausibly exogenous instruments, I allow linguistic distance to have a bounded direct effect on the outcome and ask how large that direct effect must be before $\beta = 0$ enters the AR confidence set [@conleyPlausiblyExogenous2012]. For the 2022 long change, the minimum bound is about 0.0267 log points, or 41.7% of the observed reduced form. For 2023, the corresponding threshold is about 0.0112 log points, or 19.9% of the reduced form. These are sensitivity thresholds, not estimates of a direct effect. They show that the 2023 result is especially fragile to modest departures from exact exclusion, while even the more persistent 2022 reduced form does not overcome the weak first stage.

I therefore do not use linguistic distance to estimate a causal effect of English-medium schooling. The appendix supports a more limited conclusion: linguistic distance is a persistent and economically meaningful feature of India's language environment, but it is too weakly related to within-state English-medium schooling, and too plausibly related to broader local conditions, to isolate that schooling channel. The descriptive schooling--welfare relationships in the main paper remain descriptive.

# References {-}
